### 2020-04-05 papers/sugar-synth/reviews-rebuttal.txt
asd

We thank the reviewers for their insightful comments. We will carefully
address the concerns in the final version related to the writing quality,
limited perspective of a domain expert, discussion on the guarantees of the
algorithm, and a mis-citation.

Biological systems are typically DIFFICULT to model because many components
and interactions are unknown. We have identified glycosylation as a problem
that is exceptionally suited to mathematical analysis, as its biochemistry
is well-characterized. As reviewerC has noted "The paper demonstrates
application of synthesis techniques to a new and potentially impactful
problem domain". We consider the identification of problem, as being
amenable to formal analysis, one of the key contributions of our work.

Here are the responses for reviewer's questions/concerns.

ReviewerB:"if the authors have consulted with a biologist"

One of the authors (Mukund Thattai) is a computational biologist, a domain
expert. MT has done pioneering work on Golgi biogenesis
(Mani&Thattai,eLife2016). He has developed the mathematical framework for
glycosylation that the present paper is based on
(Jaiman&Thattai,bioRxiv2018). That work has been presented at the Golgi2018
meeting and is already cited. MT’s work on glycosylation has been developed
in close interaction with Ajit Varki (acknowledged in bioRxiv preprint),
editor of the influential textbook `Essentials of Glycobiology’.

ReviewerA:"Is the mathematical model in Section 4 novel, and what is
its biological justification?"

The model is from Jaiman&Thattai[biorXiv,2018], which was cited. We will
expand our section 2 for clearer justification.

ReviewerA:"how the proposed synthesis algorithm helps"

There is as yet no good tool to predict protein glycan profiles. As a
crucial step to solving that major problem, we use glycan profiles to infer
synthesis rules. Our work is the very first, to our knowledge, to attempt
this general problem. Previous work (Jaiman&Thattai,bioRxiv2018) solves a
restricted version: the synthesis of a single glycan, while most protein
glycan profiles relevant to biology consist of multiple structures as
considered here. Further, our introduction of important enzymatic
constraints, in the form of the “specificity rules”, makes this work more
relevant to practical situations.

ReviewerA: "Can you comment on the correctness of results ...?"

In this work we make predictions about glycosylation enzyme biochemistry
and localization. These predictions are, in principle, testable. Two of the
authors(MT&AB) are presently undertaking experiments to test such
predictions. However, such experimental verification is beyond the scope of
the current paper. Our learned rules in some examples matches with the
rules that has been hypothesized in Jaiman&Thattai[biorXiv,2018]. We will
mention this in the paper.

ReviewerB:"why .... the set of production rules cannot generate
any glycan outside of the input set?"

Protein glycan profiles regulate cell-cell interactions in eukaryotes.
Synthesis of incorrect glycans leads to disorders, including cancers. This
is why we have adopted a very important constraint: we impose the condition
that glycans outside the input set are NOT synthesized.

ReviewerB:"highlight some of the challenges in coming up with their
algorithm/encoding"

The algorithm is not a trivial modification of known CEGIS tools. Our
encoding has constraints that enforce ordering of applications of rules,
support various stay model, and the other variations. The 'quantifier
instantiation' optimization helped us in making the tool work.

ReviewerA:"approach relies on clean data, whereas real-world.. is... noisy."

We discussed in section6.5 noisy data and how to adapt our method for it.
And our tool supports it by allowing generation of additional similar
glycans. We have applied the tool to real-world observations. A future
extension of this work is to use such data to develop a good noise model.

POPL 2020 Paper #214 Reviews and Comments
===========================================================================
Paper #214 Automated inference of production rules for glycans

Review #214A
===========================================================================

Overall merit
-------------
C. Weak Reject

Reviewer expertise
------------------
Y. Knowledgeable

Paper summary
-------------
The paper proposes an algorithm for synthesizing production rules for “glycans”, which are biological objects constructed from primitive components according to a fixed set of rules. The paper proposes a synthesis algorithm in the CEGIS framework, and empirically evaluates the scalability of their approach.

Strengths
---------
+ Interesting application domain
+ Paper is well written

Weaknesses
----------
- Technical approach is not particularly novel
- Approach does not handle noisy data
- Experiments are very weak

Comments for author
-------------------
The technical novelty in this paper does not appear to be very strong. Their algorithm works within the standard CEGIS framework, where they iteratively synthesize a candidate set of production rules (i.e., a “program”), and then synthesize a counterexample to the correctness of those rules, continuing until either either no counterexample can be found (success) or no valid program can be found (failure). The main contribution, then, is taking a mathematical model for how glycans are produced and convert it into constraints for both the program synthesis and counterexample detection steps.

For such a paper, I would expect the application to be justified in more detail. In particular, I would expect the paper to justify how the proposed synthesis algorithm helps improve some aspect of the biological application. However, their experiment is only in terms of synthesis time. There is essentially no discussion of whether the results are scientifically meaningful, or even correct -- i.e., can we be sure that the mathematical model of glycan construction corresponds to reality?

Another question I have is whether the mathematical model in Section 4 is novel, or if it is based on prior work. If it is novel, the authors should emphasize it as a contribution, but need to justify its biological plausibility. If it is not novel, then the authors should give a citation to the relevant work.

Finally, as the authors point out, their approach relies on clean data, whereas real-world data is often very noisy. I think their approach would be more interesting if they are able to handle noisy data, which generally speaking is a relatively understudied problem in the program synthesis literature.

Minor

- The CEGIS citation should probably point to the original work, i.e., Armando Solar-Lezama’s Ph.D. thesis “Program Synthesis by Sketching”

- Typo on line 305: “tread-off” should be “trade-off”

- On line 425, what is the definition of “subtree”?

Questions for authors’ response
---------------------------------
Can you comment on the correctness of results you obtain in your experiments?

Is the mathematical model in Section 4 novel, and what is its biological justification (i.e., is there any justification why it is a good model)?

Review #214B
===========================================================================

Overall merit
-------------
C. Weak Reject

Reviewer expertise
------------------
Y. Knowledgeable

Paper summary
-------------
This work presents an algorithm for synthesis of production rules for glycans (i.e. polymers) from sugar monomers.  In each iteration, the algorithm seeks to synthesize production rules that can generate the input glycans. If unsuccessful, the algorithm terminates indicating failure. Otherwise, the method checks for the existence of a "counterexample glycan" which is not included in the input glycans but can be generated by the synthesized rules. The algorithm terminates with the synthesized rules if there is no such counterexample and proceeds with the next iteration otherwise (with additional constraints that disallow generation of the counterexample glycan). Both the synthesis and counterexample searches are encoded as SMT problems. The tool is evaluated on some existing glycan datasets and is shown to have a reasonable run time.

Strengths
---------
The paper explores a new domain for synthesis, in particular, a domain that often presents a significant barrier to understanding the underlying computational problems.

Weaknesses
----------
The paper does not present the perspective of domain experts about the utility of this approach. The approach by itself is technically straight-forward, once the computational space is identified. Finally, the writing in the paper needs significant improvement.

Comments for author
-------------------
- The paper does not make enough of a case that the modeling of the problem is adequate and that the results of the tool are helpful for a biologist. Is the modeling of production rules accurate and complete? How do we know that the synthesized rules are meaningful? There is no domain knowledge encoded into the problem in the form of an objective function or additional constraints. There can clearly be many different production rules that satisfy the synthesis constraints, not all of which will be acceptable to a domain expert.

- Please clearly state the guarantees that Algorithm 1 provides, e.g., soundness, completeness, termination?

- From a biological standpoint, why is it important that the set of production rules cannot generate any glycan outside of the input set?

- The paper cites [Jha et al 2010] for CEGIS. This is incorrect and should be replaced with the following paper or Armando Solar-Lezama's PhD thesis:

   A. Solar-Lezama, L. Tancau, R. Bod ́ık, S. A. Seshia, and V. Saraswat, “Combinatorial sketching
   forfinite programs,” ASPLOS, 2006

- The paper's writing needs careful revising/rewriting --- the current version is distracting because of the typos and errors. The presentation of the algorithm dives right into the constraints, without first giving a high-level overview of the algorithm. Some of the description about how the constraints are actually used (e.g. in Lines 622--634) should precede the presentation of the constraints. Without this, for instance, reading the constraint on Line 603 was confusing, as this is listed amongst "conditions for valid assignments" (Line 580). Here is a small, non-exhaustive list of suggestions to improve the writing:

   . Line 9: "build" --> "built"

   . Line 67: Fix "observed glycan molecules a cell"

   . Line 76: Fix "the depth of surrounding that are". Ideally replace the word surrounding with a more suitable word through the paper.

   . Line 485 (and elsewhere): $SVars$ cannot be a "set of uncountable variables"! Perhaps you mean set with an unbounded number of variables..

   . Line 479: Clarify where the maximuma arity of a monomer is obtained from right here.

   . Line 505: "decadents" --> "descendents"

   . Line 508: "decedents" --> "descendents"

Questions for authors’ response
---------------------------------
1. I am curious to know if the authors have consulted with a biologist and can share some feedback/insights they may have obtained about the accuracy of the model and the quality of the synthesized production rules.

2. Can the authors highlight some of the challenges in coming up with their algorithm/encoding in case I missed something?

Review #214C
===========================================================================

Overall merit
-------------
A. Strong Accept

Reviewer expertise
------------------
X. Expert

Paper summary
-------------
The paper describes a new application of synthesis to infer production rules for glycans (a type of sugar tree). The basic idea is that glycans are grown by a set of enzymes that can be modeled as production rules that grow a tree (the glycan). A production rule matches a branch in a tree and grows that branch by adding extra sugar building blocks to it.

The goal is to find a set of rules that can produce all the molecules in a given set $\mu$, but only those molecules. That second requirement, of only being able to produce the given molecules, introduces universal quantification, because it has to be the case that forall molecules not in the set, the given rules cannot produce them. This constraint is enforced through CEGIS; the system produces a hypothesis set of rules, and then a checker looks for counterexample molecules that can be produced by the rules but are not in $\mu$.

One thing that makes this trickier than the standard CEGIS problem is that the counterexample molecules itself introduces a quantified constraint, since one must now ensure that the new set of rules generated cannot produce the counterexample molecule for any possible application of the rules. This turns out not to be a big problem, as the system simply passes the quantified constraint to Z3.

There are several additional complications to the problem, such as the fact that reactions can sometimes take place in compartments, which restrict which rules can apply at which times, or the fact that we may have incomplete visibility to the true set $\mu$, which from what I understand from 6.3, could even be infinite. However, the technique is general enough to handle all of these variations.

Strengths
---------
Very exciting application area.

Formalization of the problem is interesting and non-trivial.

Comments for author
-------------------
Overall, I was very excited to read the paper. The paper demonstrates the application of synthesis techniques to a new and potentially impactful problem domain. I think that in itself is a strong contribution.

In terms of the technical contribution, the encoding of the problem into constraints is non-trivial, and I am not familiar with other synthesis work that has looked at the generation of this kind of rewrite system. Even though at a high-level, the approach is a very standard application of CEGIS, the particular execution model of rewrite systems means this is not just a trivial synthesis problem that you could feed to an existing SyGuS solver or to a synthesis tool like Sketch.

One minor but important note, the paper cites Jha 2010 as the origin of CEGIS. The CEGIS technique was first described in the paper "Sketching Concurrent Data Structures" from PLDI 2008, as a generalization of an algorithm from "Combinatorial sketching for finite programs" from ASPLOS 2006.

### 2020-04-05 papers/sugar-synth/algo.tex
asd

In this section, we present a method to solve the synthesis problem of finding
a set of production rules that produce a given set $\mu$ of molecules.
Here, we are considering only single compartment case.
The generalizations are discussed in section~\ref{sec:variations}.
Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
At a high level, our method is iterative.
The method first synthesizes a set of rules $R$ that can produce the molecules in input set $\mu$, i.e., $\mu \subseteq P(S,R)$.
However, there is also an additional requirement that a molecule
that is not in $\mu$ is not producible by the synthesized rules $R$.
Therefore, the method looks for the producible molecules that are not in $\mu$, called
{\em counterexample molecule}.
If there is no such molecule, our method terminates and reports
the synthesized rules.
If there is a molecule in $P(S,R) - \mu$, we add constraints that says
that look for rules
that do not generate
the molecule and go for another iteration.
Now let us look at the details of the method.

\input{alg-synth.tex}

\subsection{\textsc{SugarSynth} in detail}

The method assumes that the input set $\mu$ is finite, which a
natural assumption.
Even if a set of rules can produce
an unboundedly large number of molecules.
No real-life biology will exhibit such a set in a cell.
Since we are synthesizing production rules, our method
needs to bound the search space.
The method also takes two numbers as input:
$d$ is the maximum height of the learned rules
and 
$n$ is the maximum number of the learned rules.
The parameters eliminate potential
unbounded search space.
If the method fails to find production rules,
the user can always call the method with larger parameters.
However, we do not expect to find a large number of or sized rules
because chemical reactions do not usually have many conditional
on the precursors and do not apply large changes in a single step.
First, the method initializes $S$ with the set of monomers occurring
in $\mu$
and sets $w$ to be the maximum arity of any monomer in $S$.

We use templates to model the search space of rules.
A template is also a tree that has a certain depth and
the internal nodes of the tree have the same number of children.
Two variables label each node of the template.
One variable is for choosing the sugar at the node and the other is for describing
the `situation' of the node.
The first variable at the node encodes choices of
sugar monomers or their absence.
The domain of the variables is $S \union \{\bot\}$.
Let $SVars$ be the set of uncountable variables with the domain.
We will use the pool of $SVars$ to add variables to the templates.

A node in a production rule can be in four situations.
In Figure~\ref{fig:rule-parts}, we illustrate the situations of the nodes in a rule.
The first situation is when a node position may be in the expanding part.
The nodes in the dark gray area are the expanding nodes.
The second situation is when a node position may not be present in the rule.
The dashed area includes absent nodes
from the full tree.
Among the matching nodes, we make two cases, namely if the expanding
nodes are the descendants of the matching node or not. 
The third situation is when a node position is in the matching part
and has expanding decadents.
The solid path from $v_0$ to the root $v_e$ of the expanding part includes
the nodes that are ancestors of expanding part among the matching nodes.
Finally, the fourth situation is when a node position is in the matching part and
has no expanding decedents.
The nodes in the rest of the light gray area are in the fourth situation.
Another variable is mapped to a node to encode the four situations.
Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
indicate the set of four situations.
Let $KVars$ be the set of uncountable variables with domain $K$. 
We will use the pool of $KVars$ to add variables to the templates.
Formally, the templates are defined as follows.

\begin{figure}[t]
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
    \node[loc] (v0) {$v_0$};
    \node[sqloc, below right of=v0] (v1) {$v_e$};
    \draw [->] plot [smooth] coordinates { (v0.south)
          ($(v0.south)+(3mm,-2mm)$)
          ($(v0.south)+(2mm,-6mm)$)
          ($(v0.south)+(9mm,-6mm)$)
          % ($(v0.east)+(8mm,-8mm)$)
          (v1.north) };
    % \path[->] (v0) edge (v1);
    % \path[->,dashed] (v1) edge (v2);
    % \path[->] (v1) edge (v3);
    \fill[fill=gray,opacity=0.2] 
    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
    \fill[fill=gray,opacity=0.4] 
    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
  \end{tikzpicture}

  \caption{Illustrating parts of production rules that are to be encoded in the rule templates.}
  \label{fig:rule-parts}
\end{figure}

\begin{df}
For given integers $d$ and $w$,
a {\em rule template} $t = (V, \nu, \kappa, C, v_{0r})$ is a labeled full tree with depth $d$ and 
each internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to distinct
sugar choice variables, 
$\kappa : V \maps KVars$ maps nodes to distinct situation variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0r} \in V$ is the root of the tree.
\end{df}
In the method at line~\ref{line:createRtemp}, we use the
call to \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
height $d$ and children width $w$. We get $d$ and $n$ as input, and
we let $w$ be the maximum arity of any sugar that occurs in $\mu$.
This allows us to map any node to any sugar.
We also assume that templates
returned by $\textsc{MakeTemplatesRule}$ do not share variables.
We obtain a production rule from a template by assigning values
to the variables.
If for a node $v$ in a template if we assign $\kappa(v) = Absent$,
we call the node {\em absent}. Otherwise, we call the node {\em present}.
Not all the assignments may result in a valid rule.
Next at line~\ref{line:ruleCorr}, we will construct constraints
that encode the set of valid rules.

A valid assignment to the variables in
a template  $t = (V, \nu, \kappa, C, r)$ must
satisfy the following six conditions.
The first two encode the basic agreement among variables.
The third encodes the tree structure of the rule.
The last three encode the relative positions of the four
parts of the rule.
\begin{enumerate}
\item If a node is present, then it is labeled with a sugar. Otherwise, we ignore the label.
$$
\Land_{s \in S} \Land_{v \in V} (\kappa(v) \neq Absent \limplies \nu(v) \neq \bot )
$$
\item  The number of children of a node should match with the arity of the labeled sugar.
  The following formula states that the children that are at greater arity than that of the label
  must be absent.
$$
\Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
$$
\item If a node is present, then the parent of the node is also present.
  $$
  \Land_{{\text{internal node } v \in V}}
  \Land_{i\in[1,w]} (\kappa(C(v,i) ) \neq Absent \limplies \kappa( v ) \neq Absent )  
  $$
\item If a node is $Expand$, then its children are also $Expand$ if present. The following constraints state that once one reached the expanding part, there
are no descendants that are matching.
  $$
\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Expand \limplies \kappa(C(v,i)) \in \{Expand,Absent\}   )
$$
\item If a node is $Match$, then its children are also $Match$ if present.
  Similarly, the following constraints state that once one reached the matching part without expanding descendants, there are no descendants that are expanding.
$$
\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Match \limplies \kappa(C(v,i)) \in \{Match,Absent\}   )
$$
\item If a node is $MatchAns$, then exactly one child is $MatchAns$ or $Expand$.
  The following constraints encode that there is exactly one subtree that is expanding part
  by stating that there is no branching in $MatchAns$ nodes and they form a path.  
$$
\Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = MatchAns \limplies \sum_{i\in[1,w]}( \kappa(C(v,i)) \in \{MatchAns,Expand\}) = 1
$$
\end{enumerate}

In the method, the call to $\textsc{RuleTemplateCorrectness}$ at
line~\ref{line:ruleCorr} creates the above constraints
and store them in $tCons$.
Next, we will also be searching for the molecules that may be produced by the learned rules.
Therefore, we need to define the search space for the molecules.
We use similar templates for defining the search space.
Since we always search in bounded space, we limit the template size using a parameter,
namely the height of the template.
\begin{df}
For given integers $h$ and $w$,
a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
every internal node has $w$ children, where 
$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
and
$v_{0m} \in V$ is the root of the tree.
\end{df}
At line~\ref{line:createMtemp}, we use the call to \textsc{MakeTemplateMol}($S$, $h$, $w$) to
create a template of height $h$ and children width $w$.
%
Our method searches for unwanted producible molecules up to the height of $h$,
which we choose to be the maximum of the heights of molecules in $\mu$.
%
The choice of $h$ is somewhat arbitrary.
%
We assume if the rules could have produced a molecule with height less than $h$,
then it must be in $\mu$.
On the other hand, we ignore the potential formation of the large size of molecules.
%
Since we observe only finite molecules in any wet experiment, we may assume
there may be large producible molecules, but we did not observe them.
However, this point is debatable and $h$ may be viewed as a parameter of the whole
method.

%
Similar to the rule templates, not all assignments to molecule template variables
are valid.
We add the following four conditions for
valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
The first  two conditions encode the tree structure of the molecule and
matching the number of children with label arity for each node.
The second encodes that the height of the molecule is large enough.
The last encodes that new molecule must not be the same as any molecule in $\mu$.
\begin{enumerate}
\item If a node is present, then the parent of the nodes is also present.
  $$
  \Land_{i\in[1,w]}\Land_{\mathrlap{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
  $$
\item The number of children of a node should match with the arity of the labeled sugar.
  $$
  \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
  $$
\item At least one of the leaves is present. This constraint is for
  avoiding the trivial solution.
  $$
  \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
  $$
  where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
  This allows us to look for all molecules that are close
  to height $h$.
  We choose $d' = h-d$, where $d$ is the height of the rule templates.
  This $d'$ ensures that we do not miss producible molecules that
  can not have exact height $h$ but is slightly larger than $h$.
  Then, there must be another producible
  molecule that is a subtree of the molecule and has height in
  between $h$ and $h-d$
  because the maximum depth of a rule is $d$.
\item We find a molecule that is not in $\mu$.
  To encode the constraints, let predicate $Neq$ be recursively defined as follows.
  \begin{align*}
  Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
  \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
    Neq(v,\bot) :=\;&  \nu(v) \neq \bot
  \end{align*}
  The disequality constraint is
  \begin{align*}
    \Lor_{\mathclap{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).
  \end{align*}
  In our implementation, we have an option to strengthen
  the constraints.
  We have an option to also ignore all the subtrees of
  the molecules of $\mu$.
\end{enumerate}
%
In our method, the call to $\textsc{MolTemplateCorrectness}$ at
line~\ref{line:molCorr} implements above constraints and store them in
$mCons$.

We need to encode that the rules do generate the molecules
in $\mu$ and do not generate any other.
Using procedure $\textsc{EncodeProduce}$, we generate
the corresponding constraints.
We will discuss the procedure in-depth shortly.
Let us continue with~\textsc{SugarSynth}.
At~\ref{line:molenc} line, we call $\textsc{EncodeProduce}$ for
each molecule in $\mu$ and generate constraints $pCons$ stating that
the solutions of the templates will produce the molecules in $\mu$.

After producing the constraints $tCons$, $mCons$, and $pCons$,
the method enters in the loop.
It first checks the satisfiability of conjunction
$tCons \land pCons \land nCons$
using some SMT solver, where $nCons$ is $\ltrue$ in the first iteration
and will encode constraints related to counterexample molecules.
If the conjunction is unsatisfiable, there are no rules of
the input number and height,
and the method returns failure of synthesis.
If it is satisfiable, it constructs the rules at line~\ref{line:getR} from
the satisfying assignment $a$
and stores them in $R$.

At line~\ref{line:consNewR}, we construct constraints $rCons$ again using
$\textsc{EnocdeProduce}$ that says that template molecule $\hat{m}$
is generated by rules $R$.
We check the satisfiability of $rCons \land mCons$.
If it is not satisfiable, we have found the rules that generate exactly
the molecules in $\mu$ within our search range of counterexample molecules,
and the loop terminates.
Otherwise, we use the satisfying assignment to create a
counterexample molecule $m'$.
At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
all possible applications of templates rules in $T$ must not produce $m'$.
We use shorthand $F \landplus G$ for $ F := F \land G$.
As we will see that $\textsc{EnocdeProduce}$ introduces fresh variable
maps $\tau$ and $cuts$ in the encoding.
Since we negate the returned formula by $\textsc{EnocdeProduce}$ 
and then we check the satisfiability, the encoding will not be correct
due to the fresh variables.
Therefore, we will need to introduce universal quantifiers to the
fresh variables.
We introduce universal quantifiers over $\tau$ and $cuts$
variables before adding to $nCons$.
Intuitively, the universal quantifiers say that all possible
application patterns of the learned rules do not produce $m'$.
Afterword, the loop goes to the next iteration.
Since the domain of all the variables is finite, eventually the loop must
terminate.

\input{alg-encode-produce}

\subsection{\textsc{EncodeProduce} in detail}

Now let us look at the encoding generated by $\textsc{EncodeProduce}$
in detail.
The process of production of molecules adds pieces of trees of nodes
one after another.
In order to show that a molecule is producible by a set of production rules,
we need to find the nodes where the production rules are applied,
the rules that are applied on the nodes, and
the order of the application of the rules.
The encoding is like encoding bounded executions of programs.

To model the production due to the application of the rules,
we need to attach three maps to the nodes of the molecules.
\begin{itemize}
\item Let $cuts$ map each node to a Boolean variable
indicating the node is the point where a rule is applied and the
molecule was expanded.
We say points of the applications of the production
rules as {\em cuts} of the tree.
All the nodes that are in between the cuts are added in a single step.

\item Let $rmatch$ map each node to a rule indicating
  the rule that is applied to expand at the node.
  We need to match a rule to a node if it is a cut point.
  % If a node is cut point, we need to the rules that added the nodes  
\item Let $\tau$ map each node to an integer variable
  indicating the time point when the node was added to the molecule.
  The order of the addition is an important constraint since already
  existing nodes in a molecule decide what can be added in the future.
  If we do not capture the order in our encoding, we may wrongly declare
  that a molecule is producible while it is not due to the matching
  constraints.
\end{itemize}

Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
that returns the encoding.
It takes a molecule $m$ and a set of rules $T$.
Both the inputs can be template or concrete.
The function is tolerant of the variations, but our presentation
assumes that the molecule is concrete and the rules are templates.
This will cover the case when the $\textsc{EncodeProduce}$ is called at
line~\ref{line:molenc} in algorithm~\ref{alg:sugar-synth}.
However, at line \ref{line:consNewR} the molecule is a template
and the rules are concrete.
We will discuss the handling of this situation at the end of this subsection.
$\textsc{EncodeProduce}$ is implemented with the help of three other functions,
$\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
All of them are presented in algorithm~\ref{alg:produce}.
%

$\textsc{EncodeProduce}$ returns constraints stating for each node $v$
and rule template $t$ that
if $v$ is at a cut and $t$ is applied at $v$, then $t$ must match
at the node $v$.
We require $rmatch(v)$ to be equal to some rule.
Therefore, if $v$ is a cut node, some template must be applied. 
Since we are matching with template rule, we do not know the
position of the head of expanding nodes.
We enumerate to all possible depths from $1$ to $d-1$ for finding
the head.
For each $\ell \in [1,d)$, we call $\textsc{EncodeP}( v, t, \ell)$
to construct the constraints encoding that $t$ is applied at $v$
and the depth of the root of the expanding nodes is at depth $\ell$.

In $\textsc{EncodeP}$,
we need to match rule template $t$ such that the root of the expanding
nodes matches with $v$.
Since we are given a guess $\ell$ for the depth of the
node.
We can traverse up from $v$ for $\ell$ steps to find the node to match
the root of $t$.
Naturally, the $\textsc{EncodeP}$ ignores returns $\lfalse$ if there
is no $\ell$th ancestor of $v$, i.e., there is no possibility of a match.
The variable $mark$ is assigned the timestamp for node $v$.
During matching, we will add constraints stating that the matching part of the
template should match to the nodes added before $mark$
and expanding part of the template should match with nodes that
are added at or after $mark$.
The local variable $c$ collects a conjunction of constraints as they are
generated in the function and $c$ is initially $\ltrue$.
The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
and traverses the nodes in $t$.
The while loop at~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
the $\ell$th ancestor from $v$,
matches all the ancestors up to the parent of $v$.
As loop traverses down, it also traverses $t$ along the matching
using variable $v_r$.
In each iteration, $v_r$ is updated to the $j$th child due to lines
\ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
if the ancestors of $v$ also traverse along with the $j$th child
(see if condition at line \ref{line:encodep-jth-child-cond}).

Let us consider the body of the loop in detail.
Let $v'$ be the $i$th ancestor at some iteration of the loop.
At lines~\ref{line:encodep-ans-match} and~\ref{line:encodep-ans-order}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
declared $MatchAns$, sugar matches in $v_r$ and $v'$, and
node $v'$ is added before $mark$.
The for loop at \ref{line:encodep-sub-for-loop}, iterates over
children of $v'$.
If there is a child of $v'$ that is not along the path to
$v$, it is matched at line~\ref{line:mtree} with the corresponding
child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
rule template and molecule and generate constraints to encode that their
nodes match.
$\textsc{MatchTree}$ is also called with a third parameter
$mark$ and a Boolean value as the fourth parameter.
We will discuss the procedure shortly.
If the $j$th child of $v'$ is along the path to $v$,
we get the node name $v'_r$ for updating
$v_r$ for the next iteration of the while loop.
After the while loop in $\textsc{EncodeP}$, $v_r$ is the
candidate node for the root of expanding nodes.
At line \ref{line:encodep-vr-expand}, we declare that $v_r$
is an $Expand$ node.
Afterword, we match the node $v$
with the corresponding node $v_r$ in the template rule by calling
$\textsc{MatchTree}$ again at line~\ref{line:encodep-mtree-expand}.
At line~\ref{line:mcut}, we also call to match the cut pattern at
the subtree of $v$ with the template rule.

$\textsc{MatchTree}$ is a recursive procedure and matches
sugar assignments between the molecule and the rule template.
If the rule template node $v_r$ does not exist, then
there is nothing to match and it returns $\ltrue$ at
line~\ref{line:matchtree-vr-absent}.
If the molecule node $v$ does not exist, then the rule node must also
be flagged absent. Otherwise, the match will fail.
The corresponding constraints $\kappa(v_r) = Absent$ is added
at line~\ref{line:matchtree-v-absent}.
If both the nodes are present, then we add constraints
that if $v_r$ is not flagged to absent in the template, then
the labels of $v$ and $v_r$ must match at line~\ref{line:matchtree-cons}.
At the same line, $\textsc{MatchTree}$ also adds constraints that
nodes are added in the molecule in the correct order.
The third input of the function is $mark$, which is the time when
node $v$ was added. 
The function also takes $isExpand$ bit as the fourth input if
$isExpand$ is true then the subtree must be added after
or with $v$, and is
false then nodes must be added before $v$.
The calls to $\textsc{MatchTree}$ from \textsc{EncodeP} use the
$isExpand$ appropriately to indicate that nodes correspond to the
matching part of the rule are added before and expanding part nodes
are added later.
Afterwords at line~\ref{line:matchtree-recurse},
the procedure calls itself for the children of $v$ and $v_r$ to
generate the same constraints for the subtree.

Once a rule is applied to a molecule, all the
nodes inside the expanding part are added together.
Therefore, there should be no cuts with in the set of added nodes.
Furthermore, if there are children nodes below the added nodes,
they must be added due to the application of some other rule.
Therefore, the children are at the cut points.
We call the above requirement {\em cut pattern}. 
$\textsc{MatchCuts}$ is a recursive function over the trees, and matches
cut patterns between the molecule node $v$ and the rule template node $v_r$.
The cuts must occur whenever nodes of the rule template transition
from present to absent.
For helping to detect the transition,
the third parameter is a constraint that encodes that the parent of $v_r$
is absent or not.
If $v$ does not exist, then there is no constraint to add therefore
the procedure returns $\ltrue$ at line~\ref{line:mcut-no-v}.
If $v_r$ does not exist, then it returns constraints encoding
that if the parent of $v_r$ was not absent, then $v$ must be at a cut at line~\ref{line:mcut-no-vr}.
If both $v$ and $v_r$ exist, then we add constraints
that if the parent of $v_r$ is not absent, $v_r$ is
flagged to be absent in the learned rule
if and only if $v$ is the cut points at line~\ref{line:mcut-cons}.
$\textsc{MatchCuts}$ recursively traverses the children of $v$ and $v_r$
and generates the constraints for the cut pattern.
The constraints added by the procedure are guarded by the presence of parents
because we need to enable constraints only when there is a possibility of transition
from presence to absence.
However, the call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
Even if the parent of $v_r$ at line \ref{line:mcut} in \textsc{EncodeP} is present,
$v_r$ is a cut point adding cut pattern constraints for the node will create inconsistency.
Therefore, we are passing $\lfalse$.

In the case when we call \textsc{EncodeProduce} with template molecule and
concrete rules.
We need to swap the roles of $\nu$ and $M$ at their occurrences
at line \ref{line:encodep-ans-match} in \textsc{EncodeP} and
line \ref{line:matchtree-cons} in \textsc{MatchTree}.
Furthermore, $\kappa(v_r)$ is a concrete value depending on the situation of $v_r$ in
the rule.
In the functions, the variable is to be replaced by the
evaluated value of $\kappa(v_r)$.

% On the other hand,
% if $v_r$ is also present, then $v$ must not be any cut point.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/bio.tex
asd

\begin{figure}[ht!]
\centering
\includegraphics[width=0.8\linewidth]{gfig1.png}
\caption{Biological details of glycan production. There are many types of sugar monomer building blocks (each represented by standard colored shapes); for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}. A glycan is a tree-like sugar tree (nodes linked by edges) attached to a substrate protein at the root (labeled `R'). Distinct edge orientations correspond to covalent bonds of distinct carbons on the sugar monomer. Curved boxes represent reaction compartments within cells, which are the site of glycan production. Each step of glycan growth (black arrows) represents the addition of a single new monomer to a specific attachment point on the tree. Each such step is catalyzed by an enzyme, labeled $E_i$. At any stage of growth, the tree can exit the reaction compartment as an output. Alternatively, it can be passed to a subsequent reaction compartment for further growth driven by different enzymes. Note that the enzymatic rule is sensitive to the two monomers being linked by a bond, as well as any branches. For example, enzyme $E_2$ will add a Galactose to a GalNAc only if the GlcNAc branch is present; otherwise, the reaction will not proceed (`X'). The structures, reactions, and enzymes shown here are illustrative, and they do not correspond to any measured data set; see the following section for a real example. In biological experiments, the combined outputs of every compartment are measured; the underlying reactions must be inferred. The inference problem is the subject of this paper.}
\end{figure}

DNA and RNA are made by copying template DNA, in a process called transcription carried out by an enzyme called RNA polymerase. Proteins are made by copying a template messenger RNA, in a process called translation carried out by a molecular machine called a ribosome \cite{alberts2013essential}. In contrast, glycans are grown without a template, in a process called glycosylation. Glycosylation is carried out, not by a single enzyme, but by a large collection of so-called GTase enzymes that assemble one sugar monomer at a time into a final glycan tree. This process is reminiscent of a factory assembly line to make a car \cite{Jaiman2018}. However, the assembly process operates without a blueprint: the final glycan structure is determined by the behavior of the enzymes themselves.

The process of glycosylation is stochastic, since enzymes can operate in different time orders \cite{Spahn2016}. It is as if factory workers could operate in many different orders while building the car, first adding doors and later windows. Moreover, the enzymes are promiscuous: they can add new monomers to many different places on the growing tree. This is as if the factory workers could add headlights at many different points on the car. Since there is no template, the existing tree determines where new monomers are added. Given the stochastic and promiscuous nature of the GTase enzymes, it is not surprising that the final product is highly variable \cite{Spahn2014}. The same set of enzymes can build many different glycan trees.

This variability is evident in the glycans observed to be produced by living cells. In a typical experiment, a protein is purified from a cell and the glycans attached to it are separated and their structure is characterized. Such an experiment produces a spectrum of glycan trees termed the protein's glycan profile \cite{Spahn2014}. A single glycan profile typically contains ten to twenty trees in measurable abundance, each tree being a tree of depth two to ten bonds.

In \cite{Jaiman440792}, the authors had reported a method to infer the production rules when a single glycan is produced. However, the biologically-interesting case is when the data set contains many glycan trees. This raises the following question: given a set of glycan trees produced by a cell,
can we infer the set of enzymes that produce the glycans? This is the problem we tackle here.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/experiments.tex
asd

In this section, we  present our implementation of the proposed method for the synthesis
and discuss the interesting optimizations in the implementation.
%
We also apply our tool on several data sets and illustrate the usability of our method.

\subsection{Implementation} 
We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool} for
analyzing the performance of the technique in practice.
%
The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
to discharge the generated satisfiability queries.
%
We have developed a specialized file format for giving input to~\ourtool.
%
The format allows one to declare monomers, list molecules over the monomers
in the SMTLIB like format,
and pass the input parameters of the method.

\subsection{Optimizations}
We have incorporated several optimizations to gain efficiency or reduce
the number of iterations in the method. Let us discuss some of the remarkable
optimizations.

\paragraph{Ordered templates}
Our method constructs a list of templates.
Since the templates are symmetric, there is a potential of unnecessary
iterations.
Let us suppose the method rejects a set of learned rules.
In the subsequent queries include the constraints that reject the set of rules,
the solvers may choose the same set of rules again by permuting the mapping
from the templates to the learned rule.
To avoid these iterations, we define an arbitrary total order over the rules.
We add ordering constraints stating that any assignments to the list of templates should produce
ordered rules with respect to the total order.

\paragraph{Quantifier instantiations}
The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
to ensure that the query is not unsatisfiable.
The instantiations sufficiently slow down the solving process.
We assist the solver by also adding an instantiation of the quantifiers that is equal
to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
We observe that for some inputs only adding the instantiations and not the universally quantified
formula is more effective.

\paragraph{Constraints for counterexample molecules}
In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
at line~\ref{line:consNewR} in each iteration.
The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
the construction of many terms repeatedly. In our implementation, we construct the constraints
once outside the while loop.
We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
of the template variables that were returned by the solver at line~\ref{line:posModel}.
Due to the assignments, the templates become concrete rules in the context of solving.

\subsection{Benchmark}

We have applied our tool on four sets of molecules form
respiratory mucins of a cystic fibrosis patient (D1),
human chorionic gonadotropin from a cancer cell line (D2), horse chorionic gonadotropin (D3), and
hydra (D4)~\cite{Jaiman2018,hydra}.
The availability of clean data, where we are clear about the source and the collection
methodology used, limits our choices for the data.

\subsection{Results}
\input{tbl-results}

We have applied \ourtool~on the data set. For each data set, we choose several
parameter combinations to illustrate the relative performance of the tool.
If we did not give large enough parameters, then the tool fails to synthesize the rules.

For D1, we synthesize the rules in 1.2 seconds. If we reduced any of the first
two parameters, we failed.
Giving an extra compartment in the first row did not impact the performance.
We synthesize the rules for D2, which is also our motivating example, in 0.4 seconds.
The performance of the tool is also robust against variations of the parameters.
We synthesize the rules for D3 in 0.9 seconds if we use optimal values for the parameter.
However, if we use too many resources or too little, the tool runs for a long time,
which is typical. The search in combinatorial space is highly sensitive to the parameters.

The synthesis for D4 takes a long time. We also observe the trade-off between the number of compartments
and the depth of the rules at the first and second row of D4.
For D3, we also learned large rules, i.e., they are adding many nodes at a single time.
Currently, the tool does not have any objective function to optimize the synthesized rules.
We are hoping to see more data sets such that we can develop a clearer picture
for a reasonable objective function.

% Our synthesized production rules for the first three data sets
% match with the reported rules in the literature.
% For the fourth, there is no hypothesis available to us.

The results suggest that the tool is potentially applicable to larger data sets with
the variants discussed earlier. However, we have not fully modeled all the information
and biological intuition --- for example, the objective function for the rules.

% As we have discussed earlier, we have implemented variants that encode

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/intro.tex
asd

The ability to control the assembly of small building blocks into large structures is of fundamental importance in biology and engineering. Familiar examples of this process from biology include the synthesis of linear DNA from nucleotide building blocks and the synthesis of linear proteins from amino-acid building blocks. In both these examples the synthesis is templated: the new DNA or protein molecule is essentially copied from an existing molecule. However, most biological assembly proceeds without a template. For example, when an adult animal is grown from a fertilized egg, the genome within the egg contains a dynamical recipe encoding the animal rather than a template. The genome restricts and controls the set of events that can take place subsequent to fertilization.

While the process of animal development is too complex to study comprehensively, the same themes arise in the synthesis of complex tree-like sugar polymers known as glycans \cite{Varki2017} that are covalently attached to proteins. Unlike linear proteins and DNA, glycans are tree-like structures: their nodes are sugar monomers, and their edges are covalent carbon-carbon bonds. The tree-like structure of a glycan is a direct consequence of the fact that a sugar monomer can directly bond to at least three neighboring sugar monomers (in contrast to nucleotides or amino acids, which can only bind to two neighbors and are constrained to make a chain).

A given cell produces a specific set of glycan molecules that are present in the cell. Since different cells produce different sets of the molecules,
the assembly process must be programmable: the assembly process includes a set of production rules. The reactions that underlie glycan production are carried out by enzymes known as GTases \cite{Varki2017}. There are hundreds of enzymes present in a given cell: each enzyme is a protein encoded by a distinct gene, which carries out a distinct biochemical reaction. The enzymes thus execute the production rules. A glycan tree is assembled piece by piece in successive steps. At each step, a production rule adds a small piece of a tree %(a monomer)
at the leaves or internal nodes of the current tree.
Not all the rules are applicable at all the leaves. The monomer at a leaf and current surroundings of the leaf controls
the applicability of a rule on the leaf.
Identifying the exact mechanism of the production by extensive
biochemical experiments are costly and often needs an initial
hypothesis for the rules to test. Biologists must identify the production rules and their control
conditions by manually analyzing the set of observed glycans in a cell
and using prior knowledge of biochemistry.
This is an error-prone process since the production rules must generate
exactly the set of molecules in the cells and nothing else; and moreover, the data sets about which glycans are present in which cells are often incomplete. Manually comprehending all possible tree generation rules is difficult and ad-hoc. It would be useful to automate the process of learning which rules are operating in a given cell, based on incomplete data.

In this paper, we are presenting a {\em novel} automated synthesis
method for the production rules.
Our method takes the observed glycan molecules a cell and synthesizes
the possible production rules that may explain the observation.
Our work is the {\em first} to consider the computational problem.
Our method of the synthesis is similar to counterexample guided
inductive synthesis(CEGIS)~\cite{cegis}.
Several methods for solving problems of searching in a complex combinatorial
space use templates to define and limit their search space,
such as learning invariants of programs~\cite{InvGenTACAS09},
and synthesizing missing components in programs~\cite{sygus,Solar-Lezama2005}.
We also use templates to model the space of the production rules.
The templates usually have parameters that limit the search space.
In our case, the size and number of templates are the parameters that
limit sizes of pieces of trees that are added in one step,
number of the production rules, and
the depth of surrounding that are checked to control the rules.

\begin{figure}[t]
  \begin{tikzpicture}[thick,shorten >=1pt,node distance=2cm,on grid]
    \node[sqloc]   (q0)                {Synthesis query};
    \node   (qi) [xshift=-1cm, left=of q0]                {Input glycans};
    \node[sqloc]           (q1) [xshift=2cm, right=of q0] {Counterexample query};
    \node   (qo) [ below=of q1]                {Synthesized rules};
    \node   (qf) [ below=of q0]                {Synthesis failed};
    \path[->] (q0) edge [bend left=45] node [above] {Synthesized rules} node [below,xshift=-1.2cm,yshift=-3mm] {sat} (q1);
    \path[->] (q1) edge [bend left=45] node [below] {Counterexample} node [above,xshift=1.2cm,yshift=3.5mm] {sat} (q0);
    \path[->] (qi) edge  (q0);
    \path[->] (q1) edge node [right,yshift=6mm] {unsat} (qo);
    \path[->] (q0) edge node [left,yshift=6mm] {unsat} (qf);  
  \end{tikzpicture}
  \caption{The architecture of our synthesis method}
  \label{fig:illus-method}
\end{figure}

Our method is iterative, as illustrated in Figure~\ref{fig:illus-method}.
Our tool first constructs quantifier-free constraints encoding that a set of unknown rules
defined by templates can assemble the input set of molecules.
The generated constraints involve Boolean variable, finite range variables, and integer ordering
constraints.
We solve the constraints using off-the-shelf SMT solver.
We call the query to the solver {\em synthesis query}.
If the constraints are unsatisfiable, there are no production rules with the search space
defined by the templates.
Otherwise, a solution of the constraints gives a set of rules.

However, there is also an additional requirement that a molecule
that is not in the input set must not be producible by the synthesized rules.
Therefore, the method looks for the producible molecules that are not in the input set.
Again the search of the molecule is assisted by a template, which bounds the height
of searched molecules.
We generate another set of constraints using the templates for the unknown molecule encoding that
are there other molecules that can be generated using the synthesized rules.
We again solve the constraints using an SMT solver.
We call the query to the solver {\em counterexample query}.
If there is no such molecule, our method terminates and reports
the synthesized rules.
Otherwise, we have found a producible molecule that is not in the input set.

We append our synthesis constraints with additional constraints stating that
no matter how we apply the synthesized rules, they will not produce the extra molecule.
Since there is a requirement that {\em all} possible applications of rules must satisfy
a condition, we have quantifiers in the constraints.
Now onward we use a solver that handles quantifiers over finite range variable in
the synthesis query.
We go to the next iteration of the method.
The method always terminates because the search space of rules is finite.

Our encoding to constraints depends on the model of execution of the rules.
The current biological information is not sufficient to make a precise
and definite model, and do the synthesis.
We have also explored the variations of the models. For example, all rules
may apply simultaneously. They apply in batches because they
stay in different compartments.
The molecule under-assembly may flow through the compartments.
The distribution of stay of the molecules in a compartment also affects
the execution model.
Furthermore, we may have variations in the type and quality of data available to us.
We have added the support for the variations in our method.

We have implemented the method in our tool~\ourtool. The tool takes input data in its own format.
We have applied the tool on data sets from published sources (these are available in the database UniCarbKB \cite{Campbell2013}).
We have obtained the rules for the data sets.
The outputs are within the expectations of biological
intuition.

The key contributions of this paper are the following.
\begin{itemize}
\item We identified {\em a new area} of application for formal methods in biology.
\item We have developed a {\em novel} synthesis method for discovering the production rules of glycan molecules from the output of the rules.
\item We have implemented the method in a tool and applied it to data sets.
\end{itemize}

We organize the paper as follows.
In section~\ref{sec:bio}, we introduce the biological background of the synthesis problem and present the basis of the formal model.
If the reader wants to avoid getting into the biological details, she may skip the section.
In section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
In section~\ref{sec:model}, we present the formal model of the glycans, their production
rules, and the templates.
In section~\ref{sec:algo}, we present our method of solving the synthesis problem.
In section~\ref{sec:variations}, we present the variations of the problem and the modifications
in our methods to support the variations.
In section~\ref{sec:experiments}, we present our implementation of the method and experiments on
some glycan data.
In section~\ref{sec:related} and section~\ref{sec:conclusion},
we present the related work, future directions, and conclude the paper.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/main.tex
asd

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

% \documentclass[conference,10pt]{IEEEtran}

\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\input{macro.tex}

\title{Automated inference of production rules for glycans}
\author{Ashutosh Gupta and Mukund Thattai and Ansuman Biswas}
%\institute{IITB and NCBS}

\date{May 2019}

\begin{document}

\begin{abstract}
  Glycans are tree-like polymers made up of sugar monomer building blocks. They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types. Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks. The rules by which these enzymes operate are not fully understood. In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans found in a cell. We have build a tool based on this method and have used it to infer rules based on published glycan data sets.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:intro}
\input{intro}

\section{Production of Glycans}
\label{sec:bio}
\input{bio.tex}

% \clearpage

\section{Motivating example}
\label{sec:motivation}
\input{motivation.tex}

\section{Modelling of the synthesis problem}
\label{sec:model}
\input{model.tex}

% \clearpage

\section{Method for the synthesis problem}
\label{sec:algo}
\input{algo.tex}

\section{Variations of the synthesis problem}
\label{sec:variations}
\input{variations.tex}

\section{Experiments}
\label{sec:experiments}
\input{experiments.tex}

\section{Related work}
\label{sec:related}
\input{related.tex}

\section{Conclusion}
% ##limitations in the model because of limited information,
% a)the boundary between two compartments is fuzzy, 
% b) the growths of two branches are correlated/ not correlated..and some more points like these, for which wealth of data is not available, hence the model was unable to incorporate the specifics
\label{sec:conclusion}
\input{conclusion.tex}

% % \bibliographystyle{unsrt}
\bibliography{biblio}

\end{document}

### 2020-04-05 papers/sugar-synth/model.tex
asd

In this section, we present the formal model for the synthesis problem.
We model glycan molecules as labeled trees and production rules are also modeled using small pieces of trees. The glycan molecules are assembled by applying the production rules repeatedly.
Our synthesis problem reduces into finding the pieces of trees that represent the production rules.

Let $S$ be the set of sugar monomers that builds glycans,
the oligomer molecules. Each $s \in S$ is associated with arity $m$ 
(written $arity(s) = m$), {\em i.e.}, the maximum number of children for $s$. The children of the monomers are indexed, which means we can refer to the $k$th child of $s$ for some $k \leq arity(s)$.
In the monomers, they correspond to bonds at specific positions from where children are connected.
Now we define the glycan molecules as labeled trees. Now onward we refer to the glycans simply as molecules.

\begin{df}
A {\em molecule} $m = (V,M,C,v_0)$ is a labeled tree, where 
$V$ is a set of nodes in the tree,
$M : V \maps S$ maps nodes to their label, 
$C : V \times \naturals \pmaps V$ maps the indexed children of nodes, and
$v_0 \in V$ is the root of the tree.
Furthermore, a molecule must satisfy the conditions of the arity of monomers.
Formally, let $M(v) = s$, if $C(v,n) = v'$ then $n \leq arity(s)$.
\end{df}

Let $m = (V,M,C,v_0)$ and $m' = (V',M',C',v_0')$ be molecules.
With some abuse of notation, we write $v \in m$ to denote $v \in V$.
For each $v \in m$, if $(v,n)$ is not in the domain of $C$, we write $C(v,n) = \bot$.
We assume that $C(v,0) = \bot$ and if $n > n' > 0$ and $C(v,n) \neq \bot$, then $C(v,n') \neq \bot$.
Let $NumberOfChildren(v)$ be equal to the number of $n$s such that $C(v,n) \neq \bot$. 
A node $v \in V$ is a {\em leaf} of $m$ if $C(v,n) = \bot$ for each $n$.
Let $depth(v)$ be the length of the path from $v_0$ to $v$.
A {\em branch} of $m$ is a path from $v_0$ to some leaf of $m$.
Let $height(m)$ be the length of the longest branch in $m$.
We define ancestor relation recursively as follows. Let $ancestor(m,v,0) = v$.
For for $d > 0$, let $ancestor(m,v,d) = ancestor(m,v', d-1)$ if $C(v',i) = v$ for some $i$.
Let recursively-defined predicate $Match(m,v,m',v')$ state that 
$v \in m$, $v' \in m'$, $s = M(v) = M(v')$, and $Match( m, C(v,n), m', C(v',n) )$ for each $n \leq arity(s)$ such that $C(v,n) \neq \bot$.
In other words, the subtree in $m$ rooted at $v$ is embedded in $m'$ at node $v'$.
Let $subtree(m)$ be the set of molecules such that
$m' = (\_,\_,\_,v_0') \in subtree(m) \lequiv Match(m,v_0,m',v_0')$.
Let us also define a utility to copy a subtree of a molecule into another molecule.
Let $Copy(m,v)$ return a molecule $m'' = (V'',M'',C'',v_0'')$ such that 
$V''$ is a set of fresh nodes, $Match(m,v,m'',v_0'')$, and $Match(m'',v_0'',m,v)$.
Both the $Match$ conditions say that the trees rooted at
$v$ and $v_0''$ are identical.

Now let us define the model of the production process of the molecules.
A production rule expands a molecule $m$ by attaching a new piece of tree at a node that has a vacant spot among its children if the surroundings of the
node satisfy some condition.
The rule is modeled as a tree that has two parts.
One part should already be there in $m$ and the other part will be appended to $m$.

\begin{df}
  A {\em production rule} $r = (V, M, C, v_0, v_e)$ is a labeled tree, where
  $V$ is a set of nodes in the tree,
  $M : V \maps S$ maps nodes to their label, 
  $C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
  $v_0 \in V$ is the root of the tree, and
  $v_e \in V$ is the root of expanding part of the rule.
\end{df}

If we {\em apply}, a rule $r$ on a molecule $m$, then it is extended at some node
$v \in m$. A copy of the descendants of $v_e$ will be be attached to
$v$ in $m$, and the rest of the nodes in the rule have to match $v$ and above.
We will refer to the descendants of $v_e$ as {\em expanding nodes}
and all the other nodes as {\em matching nodes}.

\begin{example}
  In Figure~\ref{fig:exrule}(a), we present a rule. That has
  two kinds of nodes.
  The rule adds the square node.
  Therefore, $v_e$ is the square node.
  The circular nodes $A$ and its right child $B$ are part of a pattern.
  The pattern must be present in the molecule to apply the rule.
  In Figure~\ref{fig:exrule}(b), we present an application of the rule.
  The solid tree with three nodes is the initial molecule.
  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
  Upon applying the rule, the rule adds a left child with label $A$ to the middle node.
  The rule is not applicable at the root $A$, since it has a right child $A$
  and the combination does not match the pattern.
\end{example}

\begin{figure}[t]
  \begin{minipage}{0.48\linewidth}
    \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$B$};
    \node[sqloc, below left of=v0] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->] (v0) edge (v2);
  \end{tikzpicture}\par
  (a)
  \end{minipage}
  \begin{minipage}{0.48\linewidth}
        \center
  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
    \node[loc] (v0) {$A$};
    \node[loc, below right of=v0] (v1) {$A$};
    \node[loc, below right of=v1] (v3) {$B$};
    \node[loc, below left of=v1] (v2) {$A$};
    \path[->] (v0) edge (v1);
    \path[->,dashed] (v1) edge (v2);
    \path[->] (v1) edge (v3);
  \end{tikzpicture}
  \par
  (b)
  \end{minipage}

  \caption{(a) An example of a rule. (b) An application of the rule.}
  \label{fig:exrule}
\end{figure}

We naturally extend the definitions related to molecules, including $Match$ and $Copy$,
to the production rules.
Let us formally define the molecule productions using the rules.
Let $m = (V,M,C,v_0)$ be a molecule and $r = (V_r, M_r, C_r, v_{0r}, v_e)$ be a production rule.
Let $d$ be such that $v_{0r} = ancestor(r,v_e,d)$, i.e., $v_e$ is at the depth $d$ in $r$.
Let $i$ be such that $C_r(v',i) = v_{0r}$ for some $v' \in r$.
We apply $r$ on $m$ at node $v \in m$ such that $C(v,i) = \bot$.
We obtain an expanded molecule as follows.
Let $(V',M',C',v_0') = Copy(r,v_e)$.
The expanded molecule is
$$
m' = (V \uplus V', M \uplus M', C \uplus C' \uplus \{(v,i) \mapsto v_0'\}, v_0)
$$ if
$Match( r, v_{0r}, m', ancestor(m', v_0', d) )$.
The match condition states that after attaching the new nodes $V'$
the rule tree must be embedded in $m'$ at the $d$th ancestor of $v'_0$. 
We write $m' = Apply(m, v, r)$ to indicate the application of $r$
on molecule $m$ at node $v$ that results in $m'$.
If $r$ is not applicable at $v$, we write $Apply(m, v, r) = \bot$.
We write $m' = Apply(m, r)$ if there is a $v \in m$ such that
$m' = Apply(m,v,r)$.

Let $R$ be a set of rules.
A molecule $m$ is {\em producible} by $R$ from a set of molecules $Q$
if there is sequence of molecules $m_0,...,m_k$
such that $m_0 \in Q$, $m_k=m$, and
for each $0<i\leq k$, $m_{i} = Apply(m_{i-1},r)$ for some $r \in R$.
Let $P(Q,R)$ denote the set of molecules that are producible from
rules $R$ from a set of molecules $Q$.
We have discussed in section~\ref{sec:bio} that all the production rules
are not applied at the same time.
The rules may live in compartments and
the rule sets of the compartments are applied one after another.
To model compartments for the rules,
let us suppose we have a sequence $R_1,...,R_k$ of set of rules.
Let $P(Q, R_1,..,R_k) = P(..P(P(Q,R_1),R_2),..,R_k)$ denoting the
trees obtained after applying the rule sets one after another.

\subsection{Synthesis problem}
As discussed in section~\ref{sec:bio},
we observe a set of glycan molecules present in a cell.
%
However, we do not know the production rules to produce the molecules.
%
We will be developing a method to find the production rules.
%
Let us suppose we are given a set of molecules $\mu$ produced by some
unknown set of production rules.
%
The {\em synthesis problem} is to find a set $R$ of production rules
such that $\mu = P(S,R)$,
where $S$ is the set of monomers.

%
We can define a more general form of the synthesis problem.
Let us suppose we are also given $k$ compartments with unknown rules.
The goal of synthesis is to find  $R_1,...,R_k$ such that
$\mu = P(S, R_1,..,R_k)$.
We may relax the requirement of exactly producing $\mu$.
We may say that molecules in $\mu$ have to be produced, but we are ok
if subtrees of the molecules of $\mu$ are also produced.
Formally, we may weaken the requirement to $\mu \subseteq P(Q, R_1,..,R_k) \subseteq subtree(\mu)$.

The above are a few simplified versions of the biological problem.
We may generalize the problem further where
%rules are partitioned and the partitions are applied in phases one after another,
rules are not applied exhaustively due to time constraints,
the given $\mu$ is finite and may not be exhaustive,
and we only know the weights of the parts of molecules in $\mu$.
We will first present a method for solving a simplified version of the problem.
However, our tool handles some of the above variations.
We will discuss the variations in section~\ref{sec:variations}.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/motivation.tex
asd

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{gfig2.png}
\caption{An example of a glycan data set. Figure credit: Anjali Jaiman, PhD thesis.}
\label{fig:dataset-gly}
\end{figure}

\input{fig-ex-gly}

In this section, we first present a motivating example to illustrate our method.
In Figure~\ref{fig:dataset-gly}, we show the glycan oligomers associated with human chorionic
gonadotropin~\cite{Harrd1992}.
The data set has four glycan oligomers (shown in boxes and numbered). If we assume that all these oligomers
are built by starting from a root GalNAc (yellow square) by adding one monomer at a
time (lines between glycans represent monomer addition reactions). At the top of the figure, we illustrate
if all enzymes (rules) operate in a
single compartment, a large number of glycans can potentially be made in addition to the measured ones.
In the lower part of the figure, we illustrate if the enzymes are split into three compartments
(separated by dotted lines), then certain reactions are prevented from occurring, so the set of structures
made is reduced.
In this case, we assume that only the terminal (bottom-most) structures will be produced as outputs.
Here we have assumed certain rules of operation that are most consistent with the observed glycan data set.
The goal of this paper is to infer the rules.

Now we may consider the abstract general problem. In Figure~\ref{fig:ex-gly}(a), we present a set of
glycan molecules present in a cell consists of three molecules, which are structurally similar to the
three glycan molecules in Figure~\ref{fig:dataset-gly}.
To keep things simple for illustration, we have dropped the third glycan molecule
(inside light blue square) from the four in
Figure~\ref{fig:dataset-gly}.
%
The molecules contain four types of monomers.
As we are considering the abstract case, we have named them $A$, $B$, $C$, $D$.
Each monomer is associated with an arity, i.e., the maximum number of potential children.
The arities of the monomers are $2$, $1$, $1$, and $0$, respectively.

Before considering the synthesis problem that produces exactly
the three molecules,
let us consider a set of rules in Figure~\ref{fig:ex-gly}(b)
that produce the molecules.
There are six rules.
All the rules are in the same compartment.
% The first compartment has four rules and the second compartment has three rules.
The rules have two kinds of nodes.
If the circular nodes are present around a leaf,
the rule is enabled and appends the molecule at the leaf with the square nodes.
In Figure~\ref{fig:ex-gly}(c), we show the steps of generating the middle
glycan molecule.
The first two steps add two nodes at a time.
The last step looks at the two ancestors before adding a single node.
We may avoid the need for compartments if the operations can add multiple nodes at
a time or look for a deep context for being enabled.

The second last rule in Figure~\ref{fig:ex-gly}(b) has a non-trivial condition
on the sibling of the anchor leaf node.
It requires, the parent of the new node should be $A$ and the right sibling
must be $B$.
If we do not have the sibling condition, we may be able to construct the molecule
in Figure~\ref{fig:ex-gly}(d) using the fourth and the modified fifth rule.
The molecule is not in a subtree of any of the three input glycan molecules.
Therefore, there are scenarios where rules must look into the context before applying
themselves.

Furthermore, there is another interesting case occurs in this example.
Please look at the last two rules.
The second last rule adds $C\rightarrow B$ branch.
The last rule checks if $C\rightarrow B$ nodes are present then adds $C$.
If both rules were adding one node at a time without checking deep context,
then the undesired molecule pattern in Figure~\ref{fig:ex-gly}(e) is
also producible.
We would have needed compartments to avoid the undesirable molecule.
Therefore, there is a tread-off between rule sizes and the number of compartments.

Our method for synthesis takes the three glycan molecules as input.
It also needs the budget of resources to search for the production rules.
If we allow an arbitrary number of rules,
the rules to look at their context up to
an arbitrary depth, and an arbitrarily large number of compartments,
then there is a trivial solution.
Therefore, our method needs three parameters.
The first parameter limits the number of rules.
The second parameter limits the maximum height of the rule trees.
The third parameter limits the maximum number of compartments.
If we give larger values to the parameters, then we have more chances to
find the production rules.
For this illustration, we searched for the production rules with
seven as the limit on the number of rules, three as the limit on the rule heights,
and a single compartment.

\ourtool, the tool that implements the method, reported the synthesized rules from
Figure~\ref{fig:ex-gly}(b) in 0.4 seconds.
In our tool, we first construct a synthesis query using the templates to encode
that a set of rules produces the input molecules.
We call a solver to solve the synthesis query.
After the first query, we obtain the rules presented in Figure~\ref{fig:ex-gly}(f).
Next, we look for another molecule that may be produced by the rule.
Using another query to the solver, we find a counterexample molecule that is
producible by the rules but is not in the input set.
Coincidently, we have already presented the counterexample molecule in Figure~\ref{fig:ex-gly}(e).
Afterword, we add constraints that the molecule must not
be producible by the synthesized rules in our synthesis query and go for the next iteration.
After 19 iterations, our tool synthesizes a set of rules that satisfies the requirements.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/related.tex
asd

% Success of formal methods
Formal methods have successful in applications to a vast range of
problems from the analysis of systems.
%
In particular,
the verification of hardware designs
using SAT solver based model-checking has been applied
to industrial-scale problems~\cite{biere1999symbolic2}.
%
The verification of software systems is a much harder problem, and
many methods like bounded model-checking~\cite{biere2003bounded},
abstract interpretation~\cite{lattice77}, and
counterexample guided abstraction refinement (CEGAR)~\cite{ClarkeCEGAR} have
been successful for the problem.
%

Synthesis of programs that implements a given behavior
has been a focus of research for a while~\cite{PnueliSynthesis}.
%
In recent years, the synthesis of programs using SAT/SMT solvers
has gained momentum.
%
The approach encodes the search of a program that exhibits a certain
behavior into a satisfiability problem.
%
The solvers attempt to find a solution of the satisfiability problem.
%
The solution is the synthesized program.
%
In~\cite{SrivastavaSynthesis,Solar-Lezama2005},
the method has been successfully applied
to find programs that satisfy quantified specifications like sorting
programs, and have missing ``holes'' and need proper implementations
for them.
%
In~\cite{exampleSynth},
a set of pairs of input and output examples is the specification of a program,
and the synthesis method searches for programs in a space defined by a
template~\cite{sygus}.

% Boolean networks

% Program synthesis
Since the formal methods have been exhibiting effectiveness in the vast
range of problems, the
systems biology community has been applying the methods for various
biological problems~\cite{fisher2007executable}.
The formal methods approach is distinctly different from the statistical
approach traditionally used in biology.
In formal methods, we aim to match precisely the expected behavior rather
than learn approximate artifacts from biological data.
The use of formal methods belongs to two broad categories:
the analysis of biological models and the synthesis of the models.

The key focus has been to model gene regulatory networks (GRNs), since
they are the core of the central dogma of biology.
Boolean networks are often used to model
GRNs to find stable points or attractors, i.e., nodes in the networks
where a system eventually reaches no matter where it starts~\cite{wang2012BooleanOverview}.
% In~\cite{boolean-networks-stabilty}, the networks are used to evaluate
% stability of GRNs under mutations.
The Boolean networks are not often sufficient to adequately model the behavior
of GRNs.
The continuous-time Markov chains (CTMCs) are used to bring in the aspects of
timing and probabilistic constraints to GRNs.
The transient behaviors of GRNs are also crucial for various aspects
of biology.
% For example, the decision making in some cells is based on
% response timing to stimulus~\cite{response-timing-paper}.
In~\cite{delayedCTMC}, a methods based on
model-checking is used to estimate the time evolution of the probability
distributions over states of GRNs.

There are many biological processes where we can observe the system behavior,
but the exact mechanisms of the processes are not known.
%
Recently, the methods for formal synthesis are finding their applications in biology
\cite{dunn2014defining, xuPluripotency, booleanModelKarp13, paoletti2014analyzing}.
%
In~\cite{koksal2013synthesis},
GRNs with unknown interactions are modeled using
a automata-like model with missing components
and the specification is the variations of the behavior of the system under mutations.
Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
Otherwise, the method learns a new sample where the program violates the specifications.
It adds the sample to the set and goes to the next iteration.
Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
solving the generated queries.
%
Our method follows a similar pattern, where
a set of sampled observations is a specification, i.e., output molecules,
and we need to find the governing programs, i.e., production rules.
%
In a subsequent work~\cite{fisher2015synthesising}, a Boolean network model is used,
the functions attached to the nodes are considered unknown, and
a different kind of data sets provides the behavior specification.
In this approach, their earlier method is adopted to work in the new situation.
%

As far as we know, there has been no similar computation based analysis of glycan production rules.
%
However, there has been a theoretical analysis of the properties of the production
rules in~\cite{Jaiman2018}. The work identifies the conditions for the production
of a finite set of molecules and the cases of unique production methods.
%
In our work, we are taking the computational approach of synthesis from formal methods
instead of looking for the theoretical conditions.

% 
% To resolve this paradox we borrow from the field of
% al-gorithmic self-assembly, which explores how small buildingblocks
% with stochastic local interactions assemble into various global
% configurations [23]–[25].  A central inverse problemin self-assembly
% is to design building blocks that assembleinto a target shape.  This
% process has been studied using thetheoretical framework of Wang tiles:
% square building blockswith colored sides that stick to one another
% along sides withmatching colors [26].  For efficient assembly of Wang
% tiles,their coloring must be cleverly chosen so the target shape isthe
% unique terminal output of the stochastic assembly process[27].   That
% is, their assembly must be algorithmic.  Glycans may be considered a
% natural realization of the Wang construct,with monomers acting like
% tiles whose stickiness is encodedby GTase enzymes.  If glycan
% biosynthesis could be madealgorithmic, cells could suppress unwanted
% byproducts andgenerate only desired glycan oligomers with high yield.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 papers/sugar-synth/variations.tex
asd

In the previous section, we have presented a simplified version of
the synthesis problem.
However, the biological problem has more details.
There are unsettled modeling choices.
We have developed and implemented several variations of the
above method.
The user may choose the variation depending on the choice of model.

\subsection{Compartments}

The production rules may live in different compartments.
The compartments are ordered.
A molecule moves from one compartment to the next.
The rules of the current compartment apply to the molecule.
We have presented the formal description of the compartments in
section~\ref{sec:model}.
To support the compartments, we take an additional integer
input $k$ in $\textsc{SugarSynth}$ to indicate the maximum number
of compartments.
We construct each template rule with a new variable with domain $[1,k]$
indicating the compartment of the rule.
Let $v_r$ be a node of some template rule. We will write $compart(v_r)$
for the compartment variable for the template rule.

We will alter $\textsc{EncodeProduce}$ and its sub-procedures to
ensure that they enforce the compartment order.
Essentially, we need to encode that a rule is applied when all the
pattern nodes were added in the current or earlier compartment.
For the encoding, we add a new map $compart$ that maps
nodes to variables with domain $[1,k]$.
We modify the function $\textsc{MatchTree}$ by replacing
the {\bf if} by the following code.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:matchtree-v-absent}
  \If{$isExpand$}
   \State $tCons := ( mark \leq \tau(v)  \land compart(v_r) = compart(v) )$
  \Else
   \State $tCons := ( \tau(v) < mark  \land compart(v_r) \geq compart(v) )$
   \EndIf
   \vspace{1ex}
\end{algorithmic}
In the above code, we have inserted constraints related to compartments
as discussed earlier.
Similarly, we also need to modify the function
$\textsc{EncodeP}$ by replacing line~\ref{line:encodep-ans-order}
by the following line, which also inserts extra constraints.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:encodep-ans-match}
  \State $c \landplus \tau(v') < mark \land compart(v_r) \geq compart(v')$
  \vspace{1ex}
\end{algorithmic}
After adding the above constraints, we can divide the learned rules
into compartments.

\subsection{Compartment stay model}

When molecules pass through compartments, they are expected to stay
for a while in the compartments.
%
The length of the stay determines the number of rule applications
occurring on the molecule.
%
In our method, we assume that the molecules may stay any length of time
in a compartment, which may result in any number of applications of the rules.
%
% This allows the maximum number of molecules are produced for a given set of rules.
%
We can assume a whole range of {\em stay models} for the synthesis.
%
We have considered one more stay model, where a molecule stays
in the compartment until no more rule applications are possible.
%
We call the stay model as {\em infinite stay}.

To encode the infinite stay model, we need to add constraints that
say when a molecule goes to the next compartment, no rule of
the current compartment is applicable.
%
We add the following constraints in \textsc{EncodeProduce}.
$$
\Land_{v\in m}\Land_{i \text{ such that } C(v,i) \neq \bot}\Land_{t \in T}
compart( v ) \leq compart(t) < compart(C(v,i)) 
\limplies \Land noMatch(C(v,i),t)
$$
where constraint $noMatch(v,t)$ is collected in \textsc{EncodeP}
by inserting the following code after the while loop and an update on $c$ before making the last two updates on $c$.
$noMatch(v,t)$ is $\ltrue$
before the call to \textsc{EncodeP}.
\begin{algorithmic}[1]
  \vspace{1ex}
  \setcounterref{ALG@line}{line:encodep-vr-expand}
  \State $nomatch(v,t) \landplus \lnot c$
  \vspace{1ex}
\end{algorithmic}
The constraints state that for each node $v$ and template rule $t$
if a rule adds node $v$ before or at the compartment of $t$
and its $i$th child after the compartment of $t$,
then $t$ must not be applicable at the $i$th child.
The above constraints effectively encode that the molecule
cannot expand at some node until all the relevant rules on
earlier compartments are disabled.
Note that if there is no child at the $i$th node, we are adding no constraints of disabling rules of future compartments.
One may also add those as a requirement, depending on
the interpretation of the stay model.

\subsection{Unbounded molecules}

We always observe a finite set of molecules. However,
a set of production rules is capable of producing
an unboundedly large size of molecules.
If due to some biological intuition, we believe that a subtree in a molecule
will form a repeating pattern by attaching a copy of itself to
one of its leaves.
We pass such information to our method.
We add constraints to our method to encode that
the last compartment consists of rules for the repeating pattern,
the production rules add a new copy of the subtree only after looking
within the context of the previous copy of the subtree,
and once the rules for repeating pattern are active
all the other rules of the last compartment must not be applicable.

\subsection{Full organism data vs single cell data}

The experiments that observe the set of glycan molecules are of two kinds. In one kind, we isolate a single cell type of an organism and identify all the glycans present in the cells. In our presentation, we have assumed this kind of source of data. However, the experiments are difficult to conduct.
In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
So the information that which glycans are coming from which
cells is unknown.
In this situation, the synthesis has another task to map the molecules
in $\mu$ to several cell types.
A glycan may be present in multiple cell types.

Our method is easily adapted to consider this kind of data.
We search for a small covering set of subsets of
$\powerset{\mu}$ such that each set in the cover allows a small
set of production rules.
The cover sets indicate the different types of cells in terms of
the presence of glycans.
This variation makes the problem particularly hard.
There can be a potentially large number of covering subsets.
So far, we have not encountered a large enough data set such that
we apply the variation effectively.

\subsection{Incomplete data}

The experiments are imperfect. They may not detect all possible glycans
in a cell.
We may need to leave the possibility of allowing a few more molecules to be
producible beyond $\mu$.
We replace the no extra molecule constraint by a constraint that allow molecules
that are `similar' to molecules in $\mu$.
However, we could not imagine any measure of similarity that naturally stems from
biological intuition.
For now, our tool supports the count of the number of monomer differences
as a measure of similarity.

%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2020-04-05 src/bash_script.sh

@@ -1,0 +1,5 @@
+#!/bin/sh
+
+cd ./build/depqbf/examples
+gcc -o depqbf-file depqbf-file.c -L.. -lqdpll
+./depqbf-file

### 2020-04-05 src/main.cpp

@@ -1,0 +1,365 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <z3++.h>
+#include <future>
+#include <vector>
+
+#include <unistd.h>
+#include <chrono>
+#include <ctime>
+#include <fstream>
+#include <iostream>
+//#include <time.h>
+
+#include <boost/program_options/options_description.hpp>
+#include <boost/program_options/parsers.hpp>
+#include <boost/program_options/variables_map.hpp>
+#include <boost/token_functions.hpp>
+#include <boost/tokenizer.hpp>
+
+using namespace boost;
+using namespace boost::program_options;
+
+#include <exception>
+
+#include <load-vts.h>
+#include <vts.h>
+#include "sugar/sugar.h"
+#include "z3-util.h"
+
+
+int funcType = -1;
+int synthVar = -1;
+bool useZ3 = false;
+bool displayGraph = false;
+bool printModel = false;
+bool inputFile = false;
+std::vector<std::string> inputFilename;
+std::string mode = "vts";
+
+int parse_cmdline(int ac, char* av[]) {
+
+  //-----------------------------
+  // Setup options
+  //-----------------------------
+
+  options_description general("General options");
+  general.add_options()("help", "produce a help message")
+        //  ("help-module", value<std::string>(),
+        //   "produce a help for a given module")
+        //  ("version", "output the version number")
+        ;
+
+    /*
+    options_description gui("GUI options");
+    gui.add_options()
+      ("display", value<std::string>(), "display to use")
+      ;
+   */
+
+  options_description variation("Variation options");
+  variation.add_options()
+    ("func-type,f", value<int>(),"fix a function type for QBF")
+    ("synth-var,s", value<int>(), "fix a synthesis variation for vts");
+
+  options_description options("Available options");
+  options.add_options()
+    ("use-z3,z", "use z3 for QBF solving")
+    ("print-model,p", "print vts model")
+    ("display-graph,g","display the vts as graph")
+    ("input-file,i", value<std::vector<std::string> >(), "input file.");
+
+  options_description sugar_options("sugar synthesis");
+  sugar_options.add_options()
+    ("sugar-mode,r", "run sugar synthesis");
+
[…]

### 2020-04-05 src/solver/card-constr.cpp

@@ -1,0 +1,484 @@
+#include<vts.h>
+#include<z3-util.h>
+//#include <vector>
+#include <iterator>
+
+VecExpr vts::flattern_3d ( Vec3Expr& dump ) {
+  VecExpr d1;
+  for ( unsigned int i = 0; i < N; i++ ) {
+    for( unsigned j = 0 ; j < N; j++ ) {
+      if (j == i)
+        continue;
+      for ( unsigned q = 0; q < E_arity; q++ ) {
+        d1.push_back( dump[i][j][q] );
+     }
+    }
+  }
+  return d1;
+}
+
+z3::expr vts::cardC (z3::expr_vector d1, unsigned drop_count) {
+   z3::expr expr = ctx.int_val(0);
+        auto tt = ctx.bool_val(true);
+        for (unsigned i=0; i < d1.size(); i++ ) {
+         expr = expr + z3::ite( d1[i], ctx.int_val(1), ctx.int_val(0) ) ;
+       }
+     // std::cout << "The total count is : " << expr;  
+      return (tt && (expr == ctx.int_val(drop_count)) );
+}
+
+
+// Use PbEq for exactly k.
+z3::expr vts::exactly_k_drops( unsigned drop_count, Vec3Expr& dump ) { //
+  // D2: Flattening the array. Avoid i == j.
+  VecExpr d1 = flattern_3d( dump );
+  
+  // Print the flattern arrar.
+  // for ( auto& i: d1 ) {
+  //	  std::cout << i  << "\n";
+  // }
+  // exit(0);
+
+  //unsigned int L = d1.size();
+  
+  // Only have support for exactly 2 and 3.
+  if ( drop_count == 1 ) {
+    z3::expr al = at_least_one ( d1 );
+    z3::expr am = at_least_two ( d1 );
+    return ( al && !am );
+  }
+
+  if (drop_count == 2) {
+    z3::expr al = at_least_two( d1 ); 
+    z3::expr am = at_least_three( d1 ); 
+    return ( al && !am );
+  }
+  
+  else if (drop_count == 3) {
+    z3::expr al = at_least_three( d1 ); 
+    z3::expr am = at_least_four( d1 ); 
+    return ( al && !am );
+  } 
+  
+  else if (drop_count == 4) {
+    z3::expr al = at_least_four( d1 ); 
+    z3::expr am = at_least_five( d1 ); 
+  } 
+  
[…]

### 2020-04-05 src/solver/depqbf-printer.cpp

@@ -1,0 +1,164 @@
+#include <fstream>      
+#include <iostream>
+#include <sstream>
+#include <string>
+#include "z3-util.h"
+
+
+
+// Adds all the var to the depqbf file. 
+void add_vars( std::string s, std::ofstream& ofs) {
+  std::stringstream iss(s);
+  do 
+  {
+    std::string subs;
+    iss >> subs;
+    ofs << "qdpll_add (depqbf," << subs << ");\n"; 
+    if ( subs == "0" ) {
+      ofs << "\n";
+      break;
+    }
+  } while (iss);
+}
+
+void depqbf_file_creator(VecExpr& edgeQuant, unsigned int equant_len) {
+
+  // Begin the printing basic structure
+  std::ofstream ofs;
+  //ofs.open( "/tmp/depqbf.c", std::ofstream::out );
+  ofs.open( "./build/depqbf/examples/depqbf-file.c", std::ofstream::out );
+
+  ofs << "#include <string.h>\n";
+  ofs << "#include <stdlib.h>\n";
+  ofs << "#include <stdio.h>\n";
+  ofs << "#include <assert.h>\n";
+  ofs << "#include \"../qdpll.h\"\n\n";
+  //ofs << "#include <map>"
+
+  ofs << "int main (int argc, char** argv) \n";
+  ofs << "{ \n";
+
+  ofs << "unsigned int interpretation[" << equant_len << "];\n";
+  ofs << "unsigned int index = 0; \n"; 
+  ofs << "int num;\n";
+  ofs << "FILE *fptr;\n";
+  ofs << "fptr = fopen ( \"/tmp/out.txt\",\"w\");\n";
+  ofs << "if(fptr == NULL) { \n";
+  ofs << "  printf(\"Error!\");\n";   
+  ofs << "  exit(1);\n } \n";             
+  //ofs << "std::map<VecExpr,bool> interpretation;"; 
+  ofs << "/* Create solver instance. */\n";
+  ofs << "QDPLL *depqbf = qdpll_create ();\n";
+
+  ofs << "/* Use the linear ordering of the quantifier prefix. */\n";
+  ofs << "qdpll_configure (depqbf, \"--dep-man=simple\");\n";
+  ofs << "/* Enable incremental solving. */\n";
+  ofs << "qdpll_configure (depqbf, \"--incremental-use\");\n\n";
+
+  ofs << "/* Enable solution. */\n";
[…]

### 2020-04-05 src/solver/depqbf-run.cpp

@@ -1,0 +1,46 @@
+#include <iostream>
+#include <thread>
+#include <mutex>
+#include <condition_variable>
+#include <chrono>
+#include <ctime>
+
+
+
+int depqbf_run_with_timeout()
+{
+    auto start = std::chrono::system_clock::now();
+
+    std::chrono::milliseconds span (1000);
+    std::mutex m;
+    std::condition_variable cv;
+    int retValue;
+    
+    std::thread t([&m, &cv, &retValue]() 
+    {
+        retValue = system("./src/bash_script.sh");
+        cv.notify_one();
+    });
+    
+    t.detach();
+    
+    {
+        std::unique_lock<std::mutex> l(m);
+        if(cv.wait_for(l, span) == std::cv_status::timeout) { 
+            auto end = std::chrono::system_clock::now();
+            std::chrono::duration<double> elapsed_seconds = end-start;
+            std::cout << "Elapsed time is : " << elapsed_seconds.count() << "\n";
+            throw std::runtime_error("Timeout");
+        }
+    }
+   
+    if(retValue == -1){
+      std::cout << "SYTEM ERROR !!!\n";
+     }
+
+ //   std::cout << "Return Value is : " << retValue << "\n";
+ //   auto end = std::chrono::system_clock::now();
+    //std::chrono::duration<double> elapsed_seconds = end-start;
+//    std::cout << end - start << "\n";
+    return retValue;    
+}

### 2020-04-05 src/solver/gates-encoding.cpp

@@ -1,0 +1,175 @@
+#include<vts.h>
+#include<z3-util.h>
+
+/** Boolean Gates Implementation ***
+ * L := a G L | a          a : Bool_var [args]   G : gate
+ * G == AND | OR 
+ */
+
+// The expression can take any gates. gth gate.
+z3::expr vts::gates( VecExpr& u, z3::expr x, z3::expr y ) {
+  assert( u.size() == 1 );
+  auto g1 = x && y;
+  auto g2 = x || y;
+
+  return ( ( u[0] && g1 )  ||  ( !u[0] && g2 ) );
+
+  /** Implement in case of more than two gates **
+  auto g3 = ctx.bool_val(true);
+  auto g4 = x;
+  auto g5 = y;
+   z3::expr_vector g_list(ctx);
+   z3::expr_vector c_list (ctx);
+
+   c_list.push_back( u[0] );
+   c_list.push_back( u[1] );
+
+   // Exactly One
+   auto cSum = at_least_one( c_list ) && !at_least_two( c_list );
+
+   auto lsCons =  ( u[k][g][0] && g1 )  ||  ( u[k][g][1] && g2 );
+
+   auto cons = cSum && lsCons;
+   return cons;
+   */
+}
+
+z3::expr vts::var_fml ( VecExpr& chooseVar, 
+                        z3::expr_vector& sideConstraintFormula, 
+                        z3::expr_vector& sideConstraintLiteral,
+                        unsigned i, unsigned m, bool e,
+                        unsigned j = 0, unsigned q = 0 ) {
+  z3::expr_vector ls(ctx);
+  z3::expr_vector cl_list(ctx);
+  z3::expr_vector il_list(ctx);
+
+  for ( unsigned m1 = 0; m1 < M; m1++ ) {
+    if ( m1 == m )  continue;
+    if ( e == true ) {
+       ls.push_back( ( chooseVar[m1] && presence_edge [i][j][q][m1] ) 
+                  || ( chooseVar[m1+M] && !presence_edge[i][j][q][m1] ) );
+    } else{
+       ls.push_back( ( chooseVar[m1] && nodes[i][m1] )  
+                  || ( chooseVar[m1+M] && !nodes[i][m1] ) ); 
+    }
+
+    cl_list.push_back( chooseVar[m1] );
+    cl_list.push_back( chooseVar[m1+M] );
+    il_list.push_back( !chooseVar[m1] || !chooseVar[m1+M] ) ;
+  }
+
[…]

### 2020-04-05 src/solver/normal-form.cpp

@@ -1,0 +1,22 @@
+#include <vector>
+#include "z3-util.h"
+
+std::vector<z3::expr> cnf_converter( z3::expr& fml ) {
+  z3::context& c = fml.ctx();
+  z3::solver s(c);
+  z3::goal g(c);
+  g.add( fml );
+
+   z3::tactic t1(c, "simplify");
+   z3::tactic t2(c, "tseitin-cnf");
+   z3::tactic t = t1 & t2;
+   z3::apply_result r = t(g);
+   z3::goal r_g = r[0];
+   std::vector<z3::expr>  expr_list; 
+   for (unsigned int i = 0; i < r_g.size(); i++) {
+      expr_list.push_back( r_g[i]);
+   }
+  
+   return expr_list;
+}
+

### 2020-04-05 src/solver/qbf.cpp

@@ -1,0 +1,191 @@
+#include<vts.h>
+#include<z3-util.h>
+
+//#include <vector>
+//#include <iterator>
+
+z3::expr vts::create_qbf_formula ( int funcType ) {
+
+  /** Build Constrint Of the Form := [[1]] && [[2]] && [[3]]
+   * [[1]] :: Connectivity Constraint : kConnected Graph
+   * [[2]] :: V5 Constraint : Self edge not allowed
+   * [[3]] :: Function Requirements :
+   *          - No Boolean func exists.
+   *          - Function structure: 3 CNF. 
+   */
+
+  /***** Building [[1]] ****/
+  VecExpr ee_list = flattern_3d ( edges );
+  z3::expr_vector listE = edge_list();
+
+  // Arg true -> u need edge quantified outside.
+  z3::expr kconnectedConstraint = k_connected_graph_constraint ( C, false );
+
+  /***** Building [[2]] ****/
+  // [[2]] : V5  
+  z3::expr V5 = no_self_edges();                              
+  
+  /***** Building [[3]] ****/
+  z3::expr_vector listN = node_list();
+  z3::expr_vector listActiveN = active_node_list(); 
+  z3::expr_vector listPresentE = presence_edge_list();
+  z3::expr_vector listActiveE = active_edge_list();
+  z3::expr_vector listPairingM = pairing_m_list();
+  z3::expr_vector listReach = reach_list();
+
+  /* Avoid writing forall( x, forall( y, .... ))
+  z3::expr_vector qvarQbf( ctx ); 
+
+  qvarQbf.reserve( listN.size() + listActiveN.size() + listPresentE.size() + listActiveE.size() + listPairingM.size() + listReach.size() ); // preallocate memory
+  
+  qvarQbf.insert( qvarQbf.end(), listN.begin(), listN.end() );
+  qvarQbf.insert( qvarQbf.end(), listActiveN.begin(), listActiveN.end() );
+  qvarQbf.insert( qvarQbf.end(), listPresentE.begin(), listPresentE.end() );
+  qvarQbf.insert( qvarQbf.end(), listActiveE.begin(), listActiveE.end() );
+  qvarQbf.insert( qvarQbf.end(), listPairingM.begin(), listPairingM.end() );
+  qvarQbf.insert( qvarQbf.end(), listReach.begin(), listReach.end() );
+  
+  */
+  //for (const auto&& vect : listN ) {
+  //  std::cout << vect << "\n";
+  //}
+  
[…]

### 2020-04-05 src/solver/qdimacs-printer.cpp

@@ -1,0 +1,196 @@
+#include <vector>
+#include <numeric>
+#include <utility>
+#include <iterator>
+#include <fstream>      
+#include <map>
+
+#include <iostream>
+#include <sstream>
+#include <string>
+#include "z3-util.h"
+
+
+//use Z3_ast
+void collect( z3::expr e ,
+              std::vector<z3::expr>& r,
+              std::map <Z3_ast, int>& var_id_map ) {
+  if ( e.is_bool() ) {
+    z3::func_decl f = e.decl();
+    if ( f.decl_kind() == Z3_OP_OR) {
+      unsigned num = e.num_args();
+      for (unsigned i = 0; i < num; i++) {
+        collect(  e.arg(i), r,  var_id_map);
+      }
+    }else if ( f.decl_kind() == Z3_OP_NOT) {
+      collect( e.arg(0), r, var_id_map  );
+    }else{
+      // Check if element present in the list.
+      if( var_id_map.count(e) == 0 ) {
+        var_id_map[e] = var_id_map.size() + 1;
+        //std::cout << var_id_map[e];
+        r.push_back (e);  // check types
+      }
+    }
+  }
+}
+
+// todo: check type of second argument?
+std::vector <z3::expr> visit( std::vector <z3::expr>& cnf_fml, std::map <Z3_ast, int>& var_id_map ) {
+    //std::set<z3::expr> r;
+    std::vector <z3::expr> r;
+
+    for( auto& e : cnf_fml) {
+        collect( e, r , var_id_map);
+    }
+    
+    return r;
+}
+
+
+inline void print_variable( z3::expr& v, std::ofstream& o,
+                            std::map <Z3_ast, int>& vmap ) {
+  if( v.is_app() && v.num_args() == 0 && v.is_bool() ) {
+    o << vmap.at( v ) << " ";
+  }else{
+    assert(false);
+  }
+}
+
+inline void print_variables( std::vector<z3::expr>& vs, std::ofstream& o,
+                             std::map <Z3_ast, int>& vmap ) {
+  for (unsigned int i = 0; i < vs.size(); i++) {
+    z3::expr arg = vs.at(i);
+    print_variable( arg , o, vmap );
+  }
+  o << "0\n";
+}
+
+inline void print_literal( z3::expr& v, std::ofstream& o,
+                    std::map <Z3_ast, int>& vmap ) {
+  z3::func_decl f = v.decl();
+  if ( f.decl_kind() == Z3_OP_NOT) {
[…]

### 2020-04-05 src/solver/z3-hoister.cpp

@@ -1,0 +1,143 @@
+#define _NO_OMP_
+#define _MP_INTERNAL
+// #define _MP_GMP
+
+#include "ast/expr_abstract.h"
+#include "ast/ast_util.h"
+#include "ast/rewriter/quant_hoist.h"
+#include "ast/ast_pp.h"
+#include "ast/rewriter/th_rewriter.h"
+#include "ast/rewriter/expr_replacer.h"
+#include "model/model_v2_pp.h"
+#include "model/model_evaluator.h"
+#include "smt/smt_kernel.h"
+#include "smt/params/smt_params.h"
+#include "smt/smt_solver.h"
+#include "solver/solver.h"
+#include "solver/mus.h"
+#include "qe/qsat.h"
+#include "qe/qe_mbp.h"
+#include "qe/qe.h"
+#include "ast/rewriter/label_rewriter.h"
+#include "api/api_context.h"
+#include "ast/rewriter/expr_safe_replace.h"
+
+#include "z3-util.h"
+
+//----------------------------------------------------------------------------
+// Basic Z3 utils
+
+ast_manager& get_ast_manager( z3::context& c ) {
+  Z3_context z3_ctx = c;
+  ast_manager& m = mk_c(z3_ctx)->m();
+  return m;
+}
+
+ast_manager& get_ast_manager( z3::expr& f ) {
+  auto& c = f.ctx();
+  return get_ast_manager( c );
+}
+
+expr_ref get_z3_internal_expr_ref( z3::expr& f ) {
+  auto& m = get_ast_manager( f );
+  expr* f_expr = to_expr(f);
+  expr_ref f_ref(m);
+  expr_safe_replace subst(m);
+  subst(f_expr, f_ref);
+  return f_ref;
+}
+
+z3::expr get_z3_expr_from_internal_expr( z3::context& c, expr_ref& f ) {
+  Z3_ast f_middle = of_ast(f);
+  return z3::expr( c, f_middle );
+}
+
+z3::expr get_z3_expr_from_internal_expr( z3::context& c, ast* const f ) {
+  Z3_ast f_middle = of_ast(f);
+  return z3::expr( c, f_middle );
+}
+
+//z3::expr exp = to_ast(f_ref);
+//----------------------------------------------------------------------------
+// prenex normal form
+
+// void filter_vars( qe::pred_abs& m_pred_abs, app_ref_vector const& vars ) {
+//   for (unsigned i = 0; i < vars.size(); ++i) {
[…]

### 2020-04-05 src/solver/z3-util.cpp

@@ -1,0 +1,132 @@
+#include "z3-util.h"
+
+std::string to_string(z3::sort e) {
+  std::stringstream ss; ss << e; return ss.str();
+}
+
+std::string to_string(z3::expr e) {
+  std::stringstream ss; ss << e; return ss.str();
+}
+
+std::string to_string( VecExpr es) {
+  std::stringstream ss;
+  for( auto& e : es ) ss << e << ",";
+  return ss.str();
+}
+
+void print_exprs( std::ostream& o, VecExpr& es) {
+  for( auto& e : es ) o << e << "\n";
+}
+
+void print_expr_vec( std::ostream& o, z3::expr_vector& es) {
+  for( unsigned i = 0; i < es.size(); i++ ) o << es[i] << "\n";
+}
+
+void print_expr_set( std::ostream& o, expr_set& es) {
+  for( auto& e : es ) o << e << "\n";
+}
+
+void print_expr( std::ostream& o, z3::expr e) {
+    o << e << "\n";
+}
+
+void dump( z3::expr e )         { print_expr( std::cerr, e );      }
+void dump( VecExpr& es )         { print_exprs( std::cerr, es );    }
+void dump( expr_set& es )        { print_expr_set( std::cerr, es ); }
+void dump( z3::expr_vector& es ) { print_expr_vec( std::cerr, es ); }
+
+z3::expr parseFormula( z3::context& c, std::string str,
+                       const std::vector< std::string >& names ) {
+  unsigned s = names.size();
+  Z3_symbol* symbols = new Z3_symbol[s];
+  Z3_func_decl* decls = new Z3_func_decl[s];
+  std::vector<z3::expr> vars; // for keeping smart pointer alive
+  for (unsigned i=0; i < s; i++) {
+    const char* name = names[i].c_str();
+    z3::expr v = c.bool_const( name );
+    vars.push_back( v );
+    decls[i] = v.decl();
+    symbols[i] = Z3_mk_string_symbol(c, name);
+  }
+  str = sanitise_string(str);
+  std::string cmd = str.find_first_of(' ')!=std::string::npos
+    ? "(assert (" + str + "))" : "(assert " + str + ")";
+  z3::expr ast(c);
+  try {
+    Z3_ast e = Z3_parse_smtlib2_string(c, cmd.c_str(), 0, NULL, NULL, s, symbols, decls);
+    delete[] symbols;
+    delete[] decls;
+    ast = to_expr(c, e);
+  }
+  catch( z3::exception e ) {
[…]

### 2020-04-05 src/solver/z3-util.h

@@ -1,0 +1,240 @@
+#ifndef VTS_Z3_UTIL_H
+#define VTS_Z3_UTIL_H
+
+#include <vector>
+#include <unordered_set>
+#include <z3++.h>
+// #include <boost/regex.hpp>
+#include <boost/algorithm/string.hpp>
+
+//#define toDigit(c) (c-'0')
+
+
+template <class T>
+inline
+std::vector<T> to_vector( const std::vector<std::vector<T>>& double_vec ) {
+  std::vector<T> output;
+  for(typename std::vector<std::vector<T>>::size_type i=0; i < double_vec.size(); ++i) {
+    output.insert(output.end(), double_vec[i].begin(), double_vec[i].end());
+  }
+  return output;
+}
+
+struct expr_hash {
+  size_t operator () (const z3::expr& a) const {
+    Z3_ast ap = a;
+    size_t hash = std::hash<Z3_ast>()(ap);
+    return hash;
+  }
+};
+
+struct expr_equal :
+  std::binary_function <z3::expr,z3::expr,bool> {
+  bool operator() (const z3::expr& x, const z3::expr& y) const {
+    return z3::eq( x, y );
+    }
+};
+
+typedef std::unordered_set<z3::expr,  expr_hash, expr_equal> expr_set;
+typedef std::vector< std::vector<z3::expr> > VecsExpr;
+
+typedef std::vector< z3::expr > VecExpr;
+typedef std::vector< VecExpr > Vec2Expr;
+typedef std::vector< Vec2Expr > Vec3Expr;
+typedef std::vector< Vec3Expr > Vec4Expr;
+
+std::string to_string( z3::sort );
+std::string to_string( z3::expr e);
+std::string to_string( VecExpr es);
+void print_expr_set( std::ostream& o, expr_set& es);
+void print_exprs( std::ostream& o, VecExpr& es);
+void print_expr_vec( std::ostream& o, z3::expr_vector& es);
+void print_expr( std::ostream& o, z3::expr e);
+void dump( z3::expr e );
+void dump( expr_set& e );
+void dump( VecExpr& e );
+void dump( z3::expr_vector e );
+
+
+// Create typedef for tuples
+typedef std::vector< std::tuple<unsigned,unsigned> > Tup2Expr;
+typedef std::vector< std::tuple<unsigned,unsigned,unsigned> > Tup3Expr;
+typedef std::vector< std::tuple<unsigned,unsigned,unsigned,unsigned> > Tup4Expr;
[…]

### 2020-04-05 src/sugar/rule.cpp

@@ -1,0 +1,329 @@
+#include <sugar/sugar-encode.h>
+#include <fstream>
+
+//-------------------------------------------------------------------------
+//  rule: object for holding to be synthesized rules
+//-------------------------------------------------------------------------
+
+rule::rule( z3::context& ctx_, sugar_t_vec& sugars_, unsigned num_sugars,
+            rule_vec& children_, bool is_conc )
+  : ctx(ctx_), sugars(sugars_), children( children_ ), depth(1),
+    is_extended(ctx_), is_condition(ctx_), compartment(ctx_), is_fast(ctx_),
+    local_constraints(ctx_) {
+  //set depth
+  for( unsigned i = 0; i < children.size(); i++ ) {
+    if( children[i] ) {
+      unsigned new_depth = children[i]->get_depth() + 1;
+      if( new_depth > depth ) depth = new_depth;
+    }
+  }
+  if( is_conc ) {
+    is_extended = mk_false(ctx);
+    is_condition = mk_false(ctx);
+    compartment = ctx.int_val( 0 );
+    is_fast = ctx.bool_val(false);
+    for(  unsigned i = 0; i < num_sugars; i++ ) {
+      node_sugar.push_back( mk_false( ctx ) );
+    }
+    return;
+  }else{
+    is_extended = get_fresh_bool( ctx, "r_e" );
+    is_condition = get_fresh_bool( ctx, "r_c" );
+    node_sugar = get_fresh_bool_vec( ctx, num_sugars, "r_s" );
+  }
+  //make constraints
+  VecExpr cs;
+  //only one of sugar occurs
+  cs.push_back( atmost( ctx, node_sugar, 1 ) );
+  cs.push_back( z3::implies( get_occur_cons(), atleast( ctx, node_sugar, 1 )) );
+
+  // number of children should match with sugar type
+  for( unsigned i = 0; i < node_sugar.size(); i++ ) {
+    unsigned bound = sugars.at(i)->get_children_num();
+    VecExpr not_occurs;
+    for( unsigned j = bound; j < children.size(); j++ ) {
+      if( children[j] )
+        not_occurs.push_back( !children[j]->get_occur_cons() );
+    }
+    cs.push_back( z3::implies( node_sugar.at(i), mk_and( ctx, not_occurs )));
+  }
+
+  // constraints for monomer additions.
+  z3::expr is_expand_tree = is_extended && !is_condition;
[…]

### 2020-04-05 src/sugar/rule.h

@@ -1,0 +1,83 @@
+#ifndef SUGAR_RULE_H
+#define SUGAR_RULE_H
+
+#include <solver/z3-util.h>
+#include <sugar/sugar.h>
+
+class rule;
+typedef std::shared_ptr<rule> rule_ptr;
+typedef std::vector<rule_ptr> rule_vec;
+
+class rule {
+public:
+  rule( z3::context&, sugar_t_vec&, unsigned num_sugars,
+        rule_vec& children_, bool is_conc );
+
+  static void
+  make( z3::context&, sugar_t_vec&, unsigned num_rules, unsigned num_sugars,
+        unsigned depth, unsigned num_children, bool is_conc, rule_vec& rs );
+  void initialize_non_recursive( unsigned max_compartments_ );
+
+  unsigned get_children_num() const { return children.size(); }
+  rule_ptr get_child( unsigned i ) const { return children[i]; }
+  unsigned get_depth() const { return depth; }
+
+  unsigned get_num_sugar_bits() const { return node_sugar.size(); }
+  const VecExpr& get_sugar_bits() const { return node_sugar; }
+  z3::expr get_sugar_bit( unsigned i ) const { return node_sugar[i]; }
+  z3::expr get_is_extended() const { return is_extended; }
+  z3::expr get_is_condition() const { return is_condition; }
+  z3::expr get_occur_cons() const { return is_condition || is_extended; }
+  z3::expr get_is_expand_cons() const { return is_extended && !is_condition; }
+  z3::expr get_compartment() const { return compartment; }
+  z3::expr get_is_fast() const { return is_fast; }
+  void collect_local_cons(VecExpr& , VecExpr& ) const;
+
+  // for concerete rules
+  bool check_is_empty();
+  void set_sugar( unsigned idx);
+  void set_is_extended( bool );
+  void set_is_condition( bool );
+  void set_compartment( unsigned num );
+  void set_is_fast( bool );
+  void add_no_empty_condition();
+
+  z3::context& get_context() const { return ctx; }
+
+  void dump_dotty( std::ostream& out ) const;
+  void dump_dotty( std::string filename ) const;
+  void pp( std::ostream& out ) const;
+
+  static z3::expr sorted( const rule_ptr& r1, const rule_ptr& r2 );
+private:
+  z3::context& ctx;
+  sugar_t_vec& sugars;
+  rule_vec children;
+  unsigned depth;
+
+  // ------------------
+  // Encoding variables
+  VecExpr node_sugar;
[…]

### 2020-04-05 src/sugar/sugar-encode.cpp

@@ -1,0 +1,1116 @@
+#include <sugar/sugar-encode.h>
+
+//----------------
+// missing constraints
+//   * rule children match with sugar children
+//   * is sorting constraints correct
+
+
+unsigned get_true_idx( const VecExpr& bits, z3::model& m ) {
+  unsigned i = 0;
+  for(; i < bits.size(); i++ ) {
+    if( is_true( bits.at(i), m ) ) {
+      break;
+    }
+  }
+  return i;
+}
+
+//-------------------------------------------------------------------------
+//  mol_cons: object for holding constraint bits for molecule nodes
+//-------------------------------------------------------------------------
+
+mol_cons::mol_cons( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size )
+  : ctx(ctx_), m(m_.get()), cut_bit(ctx), tstamp(ctx),
+    compartment(ctx), local_constraints(ctx)
+{
+  cut_bit = get_fresh_bool( ctx, "cut" );
+  tstamp = get_fresh_int( ctx, "ts" );
+  compartment = get_fresh_int( ctx, "c" );
+  rule_match = get_fresh_bool_vec( ctx, rule_size, "rm" );
+
+  local_constraints = atmost( ctx, rule_match, 1 ) &&
+    z3::implies( cut_bit, atleast( ctx, rule_match, 1 ) );
+}
+
+mol_cons_ptr
+mol_cons::make( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size ) {
+  return std::make_shared<mol_cons>( ctx_, m_, rule_size );
+}
+
+void
+mol_cons::collect_local_cons( VecExpr& l_cons) const {
+  assert(m);
+  for( unsigned i = 0; i < m->get_children_num(); i++ ) {
+    if( m->get_child(i) ) {
+      auto cons_i = m->get_child(i)->get_cons();
+      cons_i->collect_local_cons( l_cons );
+    }
+  }
+  l_cons.push_back( local_constraints );
+}
+
+// mk concrete constraints for sugar mol node
+mol_cons::mol_cons( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size,
+                    unsigned matched_rule, bool is_cut, int time, int comp )
+  : ctx(ctx_), m(m_.get()), cut_bit(ctx), tstamp(ctx),
+    compartment(ctx), local_constraints(ctx) {
+  cut_bit = is_cut ? mk_true( ctx_ ) : mk_false( ctx_ );
[…]

### 2020-04-05 src/sugar/sugar-encode.h

@@ -1,0 +1,165 @@
+#ifndef SUGAR_ENCODE_H
+#define SUGAR_ENCODE_H
+
+#include <solver/z3-util.h>
+#include <sugar/sugar.h>
+#include <sugar/rule.h>
+
+
+class mol_cons {
+public:
+  mol_cons( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size );
+  mol_cons( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size,
+            unsigned matched_rule, bool is_cut, int time, int comp );
+  static mol_cons_ptr
+  make( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size );
+  static mol_cons_ptr
+  make( z3::context& ctx_, sugar_mol_ptr m_, unsigned rule_size,
+        unsigned matched_rule, bool is_cut, int time, int comp );
+
+  inline const VecExpr& get_rule_bits() const { return rule_match; }
+  inline z3::expr get_rule_bit( unsigned i ) const { return rule_match.at(i); }
+  inline z3::expr get_cut_bit() const { return cut_bit; }
+  inline z3::expr get_tstamp() const { return tstamp; }
+  inline z3::expr get_compartment() const { return compartment; }
+  inline z3::expr get_local_cons() const { return local_constraints; }
+  void collect_variables( expr_set& );
+  void collect_local_cons(VecExpr& ) const;
+
+  unsigned read_rule( z3::model& ) const;
+  bool read_cut_bit( z3::model& ) const;
+  int read_tstamp( z3::model& ) const;
+  int read_compartment( z3::model& ) const;
+  std::string read_tstamp_as_string( z3::model& ) const;
+  std::string read_compartment_as_string( z3::model& ) const;
+
+private:
+  z3::context& ctx;
+  sugar_mol* m;
+  VecExpr rule_match;
+  z3::expr cut_bit; // if the molecule at the cutpoint
+  z3::expr tstamp;  // time stamp for the time for addition
+  z3::expr compartment; // in the compartment the node was added
+
+  // collect conditions on the local_cons
+  z3::expr local_constraints;
+};
+
+class unknown_label {
+public:
+  unknown_label( z3::context& ctx_, sugar_mol_ptr m_, unsigned sugar_size );
+  static unknown_label_ptr
[…]

### 2020-04-05 src/sugar/sugar.cpp

@@ -1,0 +1,431 @@
+#include "sugar/sugar.h"
+#include "sugar/sugar-encode.h"
+#include "solver/z3-util.h"
+#include <fstream>
+
+// #include <iostream>
+
+sugar_t_ptr sugar_t::make( std::string symb_, unsigned num_children_,
+                           unsigned idx_ ) {
+  return std::make_shared<sugar_t>( symb_, num_children_, idx_ );
+}
+
+//-----------------------------------------------------------------------------
+// sugar_mol utils
+
+sugar_mol::sugar_mol( sugar_t_ptr& sugar_, sugar_mol_vec& children_ )
+  : sugar(sugar_), children( children_ ), parent(nullptr), sibling_number(0) {
+  if( sugar ) assert( sugar->get_children_num() == children.size() );
+  set_depth();
+}
+
+sugar_mol::sugar_mol( sugar_t_ptr& sugar_, sugar_mol_ptr child )
+  : sugar(sugar_), children(), parent(nullptr), sibling_number(0)
+{
+  if( sugar ) assert( sugar->get_children_num() == 1 );
+  children.push_back( child );
+  set_depth();
+}
+
+sugar_mol::sugar_mol( sugar_t_ptr& sugar_, sugar_mol_ptr child1, sugar_mol_ptr child2 )
+  : sugar(sugar_), parent(nullptr), sibling_number(0)
+  {
+    if(sugar) assert( sugar->get_children_num() == 2 );
+    children.push_back( child1 );
+    children.push_back( child2 );
+    set_depth();
+  }
+
+sugar_mol::sugar_mol( sugar_t_ptr& sugar_ )
+    : sugar(sugar_)
+  {
+    if(sugar) {
+      for( unsigned i = 0; i < sugar->get_children_num(); i++ ) {
+        sugar_mol_ptr null_sugar;
+        children.push_back( null_sugar );
+      }
+    }
+    set_depth();
+  }
+
+// void sugar_mol::update_child( unsigned child_num, sugar_mol_ptr& new_child ) {
+//   assert( sugar->get_num_children() > child_num );
+//   children[child_num] == new_child;
+// }
+
+// void sugar_mol::add_child( unsigned child_num, sugar_mol_ptr& new_child ) {
+//   if( new_child ) {
+//     if( children[child_num] ) {
+//       bio_synth_error( "sugar", "adding already existing child!" );
+//     }else{
+//       update_child( child_num, new_child );
+//     }
+//   }
+// }
+
[…]

### 2020-04-05 src/sugar/sugar.h

@@ -1,0 +1,101 @@
+#ifndef SUGAR_H
+#define SUGAR_H
+
+#include "utils/utils.h"
+#include "solver/z3-util.h"
+#include <map>
+
+class sugar_t;
+typedef std::shared_ptr<const sugar_t> sugar_t_ptr;
+typedef std::map<std::string,sugar_t_ptr> sugar_t_map;
+typedef std::vector<sugar_t_ptr> sugar_t_vec;
+
+class sugar_t{
+public:
+  sugar_t( std::string symb_, unsigned num_children_, unsigned idx_ )
+    : symbol(symb_), num_children( num_children_ ), idx(idx_) {}
+  inline std::string get_name() const { return symbol; }
+  inline unsigned get_children_num() const { return num_children; }
+    inline unsigned get_idx() const { return idx; }
+  static sugar_t_ptr make( std::string symb_, unsigned num_cs_, unsigned idx );
+private:
+  std::string symbol;
+  unsigned num_children;
+  unsigned idx;
+};
+
+class sugar_mol;
+typedef std::shared_ptr<sugar_mol> sugar_mol_ptr;
+typedef std::vector<sugar_mol_ptr> sugar_mol_vec;
+//
+class mol_cons;
+typedef std::shared_ptr<mol_cons> mol_cons_ptr;
+//
+class unknown_label;
+typedef std::shared_ptr<unknown_label> unknown_label_ptr;
+
+class sugar_mol{
+public:
+  sugar_mol( sugar_t_ptr& sugar_, sugar_mol_vec& children_ );
+  sugar_mol( sugar_t_ptr& sugar_, sugar_mol_ptr child );
+  sugar_mol( sugar_t_ptr& sugar_, sugar_mol_ptr child1, sugar_mol_ptr child2 );
+  sugar_mol( sugar_t_ptr& sugar_ );
+  // void add_children( sugar_mol_vec& children );
+  // void add_child( unsigned child_num, sugar_mol_ptr& new_child );
+  // void update_child( unsigned child_num, sugar_mol_ptr& new_child );
+  void pp( std::ostream& out ) const;
+  void dump_dotty( std::ostream& out, z3::model& m ) const;
+  void dump_dotty( std::string filename, z3::model& m ) const;
+  void dump_dotty( std::string filename ) const;
+
+  inline unsigned     get_depth() const { return depth; }
+  inline sugar_mol_ptr get_child( unsigned i ) const { return children.at(i); }
[…]

### 2020-04-05 src/utils/utils.h

@@ -1,0 +1,82 @@
+#ifndef VTS_UTILS_H
+#define VTS_UTILS_H
+
+#include <string>
+#include <stdexcept>
+#include <cassert>
+#include <iostream>
+#include <ostream>
+#include <sstream>
+#include <memory>
+
+#include <vector>
+// #include <boost/regex.hpp>
+#include <boost/algorithm/string.hpp>
+
+#define toDigit(c) (c-'0')
+
+inline std::vector<std::string> get_coordinates ( std::string text, bool splitOnUnder ) {
+  std::vector<std::string> results;
+  if ( splitOnUnder == false ) {
+    boost::split(results, text, [](char c){ return c == '['; });
+  } else {
+    boost::split(results, text, [](char c){ return c == '_'; });
+  }
+  return results;
+}
+
+inline std::pair<int, int> getxy (std::string var) {
+  std::pair<int, int> p2( toDigit(var[2]) , toDigit(var[4]) );
+  return p2;
+}
+
+class vts_exception : public std::runtime_error
+{
+public:
+  vts_exception(const char* what) : runtime_error(what) {}
+  vts_exception(const std::string what) : runtime_error(what.c_str()) {}
+};
+
+// everntually remove EXTERNAL_VERSION
+// should be NDEBUG == EXTERNAL_VERSION??
+// #ifndef NDEBUG
+#ifdef EXTERNAL_VERSION
+#define triggered_at ""
+#else
+#define triggered_at " (triggered at " <<  __FILE__ << ", " << __LINE__ << ")"
+#endif
+
+#define issue_warning( ss )  { std::cerr << ss.str() << "\n";}
+
+#ifndef NDEBUG
+#define issue_error( ss )  { std::cerr << ss.str() << "\n"; assert( false );}
+#else
+#define issue_error( ss )  { throw vts_exception( ss.str() ); }
+#endif
+
+
+#define vts_error( M, S ) { std::stringstream ss;                   \
+    ss << "# vts " << M << " Error - " << S                         \
+       << triggered_at << std::endl;                                  \
+    issue_error( ss ); }
+
+
+#define vts_warning( M, S ) { std::stringstream ss;                   \
+    ss << "# vts " << M << " Warning - " << S                         \
+       << std::endl;                                                    \
+    issue_warning( ss );  }
+
+
+#define bio_synth_error( M, S ) { std::stringstream ss;     \
+    ss << "# vts " << M << " Error - " << S                         \
+       << triggered_at << std::endl;                                  \
[…]

### 2020-04-05 src/vts/graph-connectivity.cpp

@@ -1,0 +1,80 @@
+#include<vts.h>
+#include<z3-util.h>
+
+// z3::expr vts::k_connected_graph_constraint ( unsigned K, bool eQuantVar ) {
+
+//   /***** Building [[1]] ****/
+//   VecExpr ee_set = flattern_3d ( edges );
+//   //edgeQuant = ee_set;
+
+//   z3::expr_vector setR1 = flattern2d ( d_reach1, N, N, true );
+//   z3::expr_vector setR2 = flattern2d ( d_reach2, N, N, true );
+  
+//   // Create:  Exists (setR1, reach_d1 && d1_all-conn )
+//   z3::expr is_reach1 = exists( setR1, reachability_under_drop_def( d_reach1 , drop1, 0 ) && remains_connected( d_reach1 )  ); 
+//   z3::expr is_reach2 = exists( setR2, reachability_under_drop_def( d_reach2 , drop2, 1 ) && gets_disconnected( d_reach2 )  );
+  
+//   z3::expr_vector setD1 = flattern3d ( drop1, N, N, E_arity, true );
+//   z3::expr_vector setD2 = flattern3d ( drop2, N, N, E_arity, true );
+  
+//   z3::expr k_min_1_connected = forall ( setD1, implies 
+// 		  (  (exactly_k_drops ( K-1, drop1 ) && only_present_edges_can_be_dropped ( drop1 )), is_reach1 ) );
+ 
+//   z3::expr k_not_connected = exists ( setD2,  
+// 		    (exactly_k_drops ( K, drop2 ) && only_present_edges_can_be_dropped ( drop2 )) && is_reach2 ) ;
+ 
+//   z3::expr at_least_k_edges = at_least_three( ee_set );
+  
+//   // [[1]] : K Connected Graph 
+//   // EXISTS [ setE, kedges && k-1Conn && knotConn ]
+//   z3::expr_vector setE = flattern3d ( edges, N, N, E_arity, true );
+
+//   if ( eQuantVar ) { 
+//     z3::expr cons = exists ( setE, at_least_k_edges && k_min_1_connected && k_not_connected );
+//     return cons;
+//   }else {
+//     z3::expr cons = at_least_k_edges && k_min_1_connected && k_not_connected;
+//     return cons;
+//   }
+
+// }
+
+z3::expr vts::k_connected_graph_constraint ( unsigned K, bool eQuantVar ) {
+
+  /***** Building [[1]] ****/
[…]

### 2020-04-05 src/vts/kcnf-encoding.cpp

@@ -1,0 +1,119 @@
+#include<vts.h>
+#include<z3-util.h>
+
+/*********   N [here = 3] CNF  Encoding *************************************
+ *
+ *   lit_list ::= a1 C1 || a2 C2 || ... || an Cn 
+ *   cl_list  ::= a1 + a2 + a3 + ... + an == N (here 3)
+ *   il_list  ::= !Coeff (C1) || !Coeff (C2) 
+ *   
+ *   outer_list ::=  Sum_i { [lit_list] && [cl_list] && [il_list] } 
+ * 
+ * * *************************************************************************/
+
+// Bool e :: edge or not (node) : [arg_list_2d, N, M, 0/1, N, E_arity]
+z3::expr vts::literal_cnf ( Vec2Expr s, z3::expr_vector& sideCons, unsigned i, unsigned k, bool e, unsigned j = 0, unsigned q = 0 ) {
+
+  z3::expr_vector outer_list(ctx);
+  z3::expr_vector lit_listC(ctx);
+  z3::expr_vector lit_listI(ctx);
+  
+  // d: no of clauses
+  for ( unsigned d = 0; d < D; d++ ) {
+    z3::expr_vector il_list(ctx);
+    z3::expr_vector cl_list(ctx);
+    z3::expr_vector inner_list(ctx);
+    
+    //std::cout << "value of k is :" << k << "\n";
+    for ( unsigned k1 = 0; k1 < M; k1++ ) {
+      if ( k1 == k )  continue;
+      
+      if ( e == true ) { 
+        inner_list.push_back( ( s[d][k1] && presence_edge[i][j][q][k1] ) 
+                          || ( s[d][k1+M] && !presence_edge[i][j][q][k1] ) );
+      } else { 
+        inner_list.push_back( ( s[d][k1] && nodes[i][k1] ) 
+                           || ( s[d][k1+M] && !nodes[i][k1] ) );
+      //std::cout << "value of k1 is: " << k1 << "\n" ; 
+      //std::cout << "Interested var is: " << s[d][k1] << " and " << s[d][k1+M] << "\n";
+      }
+      
+      cl_list.push_back( s[d][k1] );
+      cl_list.push_back( s[d][k1+M] );
+      il_list.push_back( !s[d][k1] || !s[d][k1+M] ) ;
+    }
+
+    // Exactly three : 3 CNF 
+    // To make compuational challenge easy lets fix at_most_one`
[…]

### 2020-04-05 src/vts/load-vts.cpp

@@ -1,0 +1,224 @@
+#include "load-vts.h"
+#include <iostream>
+#include <fstream>
+
+// Format
+// M
+// num
+// N
+// num
+// n
+// 2 | 2+ 3-
+// e
+// 2 3 | 2* 3* 4 5 6 7 
+// e
+// 2 3 | 2* 3* 4 5 6 7 
+
+char load_vts::peek_skip_space() {
+  while( in.peek() == ' ') {
+    in.ignore();
+  }
+  return in.peek();
+}
+
+bool load_vts::peek_num_or_eol() {
+  char ch = in.peek();
+  if( ch == '\n') return true;
+  if( ch >= '0' && ch <= '9' ) return true;
+  return false;
+}
+
+unsigned load_vts::read_num_line() {
+  std::string line;
+  std::getline ( in, line);
+  line_num++;
+  return (unsigned)atoi( line.c_str() );
+}
+
+char load_vts::read_char_line() {
+  std::string line;
+  std::getline ( in, line);
+  line_num++;
+  if( line.size() == 0 ) {
+    return 0;
+  }else{
+    return line[0];
+  }
+}
+
+std::string load_vts::read_string_line() {
+  std::string line;
+  std::getline ( in, line);
+  line_num++;
+  return line;
+}
+
+char load_vts::get_command() {
+  return read_char_line();
+}
+
+void load_vts::get_mol_num() {
+  if( M != 0 ) load_error( "re intializing number of molecules", line_num );
+  M = read_num_line();
+  mol_vars.clear();
+  for( unsigned i = 0 ; i < M ; i++ ) {
+    mol_vars.push_back( "m" + std::to_string(i));
+  }
+}
+
+void load_vts::get_node_num() {
+  if( N != 0 ) load_error( "re intializing number of nodes", line_num );
+  N = read_num_line();
+}
+
+void load_vts::get_max_edge_num() {
+  if( Q != 0 ) load_error( "re intializing max edges", line_num );
+  Q = read_num_line();
+}
+
+void load_vts::get_model_version() {
+  if( V != UN_INIT ) load_error( "re intializing model version", line_num );
+  std::string line = read_string_line();
+  if      ( line  == "MODEL_1" ) { V = MODEL_1;
+  }else if( line  == "MODEL_2" ) { V = MODEL_2;
+  }else if( line  == "MODEL_3" ) { V = MODEL_3;
[…]

### 2020-04-05 src/vts/load-vts.h

@@ -1,0 +1,56 @@
+#ifndef VTS_LOAD_H
+#define VTS_LOAD_H
+
+#include <fstream>
+
+#include "utils/utils.h"
+#include "vts/vts.h"
+
+#define load_error( M, Line ) vts_error( "", "line " << Line << ":"<< M )
+
+class load_vts {
+
+private:
+  z3::context& c;
+  std::string file_name;
+  std::ifstream in;
+  vts_ptr v = nullptr;
+  unsigned line_num = 0;
+  bool initialized_vts = false;
+
+  //model parameters
+  unsigned int  M = 0;
+  unsigned int  N = 0;
+  unsigned int  Q = 0;
+  model_version V = UN_INIT;
+  unsigned int  D = 0;
+  std::vector< std::string > mol_vars;
+
+  char peek_skip_space();
+  bool peek_num_or_eol();
+  unsigned read_num_line();
+  char read_char_line();
+  std::string read_string_line();
+  char get_command();
+  void get_label( std::vector<unsigned>& mols,
+                  std::vector<unsigned>& activity);
+  void get_mol_num();
+  void get_node_num();
+  void get_max_edge_num();
+  void get_model_version();
+
+  void get_node();
+  void get_edge();
+  void get_pairing();
+  void qr_get_pairing();
+  void get_node_function();
+  void get_edge_function();
+public:
+  load_vts( z3::context& c_, std::string file_name_ )
+    : c(c_)
+    , file_name(file_name_) {}
+  void load();
+  inline vts_ptr get_vts() { return v; };
+};
+
+#endif

### 2020-04-05 src/vts/pp_vts.cpp

@@ -1,0 +1,327 @@
+#include <string>
+#include <iostream>
+#include <fstream>
+#include <vts.h>
+#include <z3-util.h>
+#include <vector>
+#include <utility>      // std::pair, std::make_pair
+#include <map>
+#include <tuple>
+#include <algorithm>
+
+#include <boost/algorithm/string.hpp>
+
+#define toDigit(c) (c-'0')
+
+
+// Printing the Graph
+void vts::edge_dump_dot( std::string filename, z3::model mdl ) {
+  std::string style = "solid";
+  std::string color = "blue";
+  std::string node_vec;
+
+  std::cout << "dumping dot file: " << filename << "\n";
+
+  std::ofstream ofs;
+  ofs.open ( filename, std::ofstream::out );
+
+  ofs << "digraph prog {\n";
+  for( unsigned i = 0 ; i < N; i++ ) {
+    node_vec = std::to_string(i) + ":";
+    for ( unsigned m = 0; m < M; m++ ) {
+      if( is_true( nodes[i][m], mdl ) ) {
+        node_vec = node_vec + "1";
+      }else if ( is_false ( nodes[i][m], mdl ) ) {
+        node_vec = node_vec + "0";
+      }else{
+        node_vec = node_vec + "*";
+      }
+      if( is_true( active_node[i][m], mdl) ) {
+        node_vec = node_vec + "-";
+      }
+    }
+    ofs << std::to_string(i) << "[label=\"" << node_vec << "\"]\n";
+
+    for( unsigned j = 0 ; j < N; j++ ) {
+      if (i == j)
+        continue;
+      for( unsigned q = 0; q < E_arity; q++ ) {
+        if ( q == 0 ) {
+          if ( is_true(edges[i][j][q], mdl ) ) {
+            style = "solid";
+            color = "black";
+
+            std::string label = "m";
+            ofs << std::to_string(i) << "-> " << std::to_string(j)
+              <<  "[label="  << label << ",color=" << color
+              << ",style=" << style << "]\n";
+          }
+        }
+
+        if ( q == 1 ) {
+          if ( is_true( edges[i][j][q], mdl ) ) {
+            style = "solid";
+            color = "yellow";
+            std::string label =  "m" ;
+            ofs << std::to_string(i) << "-> " << std::to_string(j)
+              <<  "[label="  << label << ",color=" << color
+              << ",style=" << style << "]\n";
+          }
+        }	  
+      } // end q
+    } // end j
+  } // end i
+
+  ofs << "}\n";
+
[…]

### 2020-04-05 src/vts/synth-func.cpp

@@ -1,0 +1,434 @@
+#include <string>
+#include <iostream>
+#include <fstream>
+#include <vts.h>
+#include <z3-util.h>
+#include <vector>
+#include <utility>      // std::pair, std::make_pair
+#include <map>
+#include <tuple>
+#include <algorithm>
+#include <utils/utils.h>
+
+//#include <boost/algorithm/string.hpp>
+
+//#define toDigit(c) (c-'0')
+
+void dVar_update ( unsigned m, unsigned M, std::string& depth_var, 
+                   unsigned noOfLiterals, unsigned litCount, bool lit, bool gateCase ) {
+  std::string var;
+  if ( lit == true ) 
+    var = "arg";
+  else 
+    var = "!arg";
+    
+  //if ( m == M-1 || gateCase == true ) {
+  if ( litCount == noOfLiterals || gateCase == true ) {
+    depth_var = depth_var + var + std::to_string (m);
+  }else { 
+    depth_var = depth_var + var + std::to_string (m) + " || ";
+  }
+}
+
+void chosen_var ( unsigned candidateVar, std::vector <int> & func_arg, 
+                  std::vector<std::string>& chosenVars ) {
+  for ( unsigned v = 0; v < candidateVar; v++ ) {
+    int var = func_arg[v]; 
+    if ( var == 1 ) {
+        if ( v == candidateVar - 2 ){
+          chosenVars.push_back( "true" );
+        } else if ( v == candidateVar - 1 ) {
+          chosenVars.push_back( "false" );
+        } else {
+          auto pickedVar = "arg" + std::to_string (v);
+          chosenVars.push_back( pickedVar );
+        }
+    } else if( var == -1 ) {
+        auto pickedVar = "!arg" + std::to_string (v);
+        chosenVars.push_back( pickedVar );
+    }else {
+        continue;
+    }
+  }
+  //std::cout << "The size of Candiate var is: " << chosenVars.size() << "\n";
+}
+
+void chosen_gates ( std::vector <int>& wVarStr, 
+                    std::vector<std::string>& chosenGates ) {
+  if ( wVarStr[0] == 1 ) {
+    chosenGates.push_back( "AND" );
+  } else { 
+    chosenGates.push_back( "OR" );
+  }
+}  
+
+void print_learned_function ( std::vector<std::string>& chosenVars, 
+                              std::vector<std::string> chosenGates, 
[…]

### 2020-04-05 src/vts/vts-synthesis.cpp

@@ -1,0 +1,756 @@
+#include <utils/utils.h>
+#include <vts/vts.h>
+#include <solver/z3-util.h>
+#include <algorithm>
+
+/** To do list :
+ * 1. Take Care of at_least_k and at_most_k with asserts() or false adds
+ *
+ */
+
+/** Synthesis Variations
+ * 1. Add edge to achieve graph stability and k connected.
+ * 2. Add flow of molecules to fix fusion.
+ *   - use available molecules.
+ *   - allow other type of molecule.
+ * 3. KCnf
+ *   - Low depth circuit
+ * 4. Gates: And Or
+ *   - Function dependence with var occuring once
+ * 5. VTS repair.
+ * 6. Biological Operations
+ *  - Activate
+ *  - Deactivate
+ */
+
+z3::expr vts::annotate_mukund_graph(z3::expr_vector& fixN,
+                                    z3::expr_vector& fixActiveN,
+                                    z3::expr_vector& fixE,
+                                    z3::expr_vector& fixPresenceE,
+                                    z3::expr_vector& fixActiveE,
+                                    z3::expr_vector& fixPairingP) {
+  /* M = 14, Molecules subgraph of Mukunds VTS`
+   * [ Qa2, Qa5, Qa7, Qa8 ] ::> [ M0, M1, M2, M3 ]
+   * [ Qb8 ] ::> [ M4 ]
+   * [ Qc7, Qc8 ] ::> [ M5, M6 ]
+   * [ Qbc2, Qbc2/3, Qbc7 ] ::> [ M7, M8, M9 ]
+   * [ R2, R3, R7, R8 ] ::> [ M10, M11, M12, R13 ]
+   *
+   * Nodes: N = 3
+   *        EE :: 0
+   *        LE :: 1
+   *        PM :: 2
+   */
+
+  /*
+  // Straighten out few things.
+  z3::expr_vector listE = edge_list();
+  z3::expr_vector listN = node_list();
+  z3::expr_vector listActiveN = active_node_list();
+  z3::expr_vector listPresenceE = presence_edge_list();
+  z3::expr_vector listActiveE = active_edge_list();
+  z3::expr_vector listPairingM = pairing_m_list();
+  z3::expr_vector listReach = reach_list();
+  */
+
+  // Fix nodes of the graph
+
+  auto n0 = nodes[0][0];
[…]

### 2020-04-05 src/vts/vts.cpp

@@ -1,0 +1,1173 @@
+#include <vts.h>
+#include <z3-util.h>
+#include <algorithm>
+#include <tuple>  // std::tuple
+#include <vector>
+
+template <int index>
+struct TupleCompare {
+  template <typename Tuple>
+  bool operator()(const Tuple& left, const Tuple& right) const {
+    return std::get<index>(left) < std::get<index>(right);
+  }
+};
+// define n dimentional vectors
+//#include "boost/multi_array.hpp"
+//#include <cassert>
+
+/*
+// Functor to compare by the Mth element
+template<int M, template<typename> class F = std::less>
+struct TupleCompare
+{
+    template<typename T>
+    bool operator()(T const &t1, T const &t2)
+    {
+        return F<typename tuple_element<M, T>::type>()(std::get<M>(t1),
+std::get<M>(t2));
+    }
+};
+*/
+
+void vts::add_known_edge(unsigned n1, unsigned n2, unsigned q,
+                         std::vector<unsigned>& mols,
+                         std::vector<unsigned>& act) {
+  // std::vector < std::tuple <unsigned, unsigned> > knownEdgesTuple;
+  // std::vector < std::tuple <unsigned, unsigned, unsigned, unsigned> >
+  // knownPresenceEdgesTuple; std::vector < std::tuple <unsigned, unsigned,
+  // unsigned, unsigned> > knownActiveEdgesTuple;
+  assert(n1 < N);
+  assert(n2 < N);
+  assert(q < E_arity);
+
+  knownEdges.push_back(edges[n1][n2][q]);
+  for (unsigned i = 0; i < mols.size(); i++) {
+    assert(mols[i] < M);
+    knownPresenceEdges.push_back(presence_edge[n1][n2][q][mols[i]]);
+    if (act[i] == 0) {
+      knownActiveEdges.push_back(!active_edge[n1][n2][q][mols[i]]);
+    } else if (act[i] == 1) {
+      knownActiveEdges.push_back(active_edge[n1][n2][q][mols[i]]);
+    }
+  }
+
+  /*
+  // Tuple version
+  for ( unsigned i = 0; i < mols.size(); i++ ) {
+     knownPresenceEdgesTuple.push_back( std::make_tuple (n1, n2, q, mols[i]) );
+     knownEdgesTuple.push_back( std::make_tuple( n1, n2, q ) );
+     if ( act[i] == 2 ) {
+       continue;
[…]

### 2020-04-05 src/vts/vts.h

@@ -1,0 +1,284 @@
+#ifndef VTS_H
+#define VTS_H
+
+#include <string>
+#include <memory>
+#include <z3++.h>
+#include <z3-util.h>
+#include <map>
+
+enum model_version {
+  UN_INIT,
+  MODEL_1,
+  MODEL_2,
+  MODEL_3,
+  MODEL_4,
+  MODEL_5,
+  MODEL_6
+};
+
+
+class vts{
+public:
+  vts( z3::context& _ctx,
+       unsigned _M, unsigned _N, unsigned _Q, model_version _V, unsigned _C, unsigned _D )
+    : ctx(_ctx), M(_M), N(_N), E_arity(_Q),
+      V(_V), C(_C), D(_D),
+      knownNodes( _ctx ),
+      knownActiveNodes( _ctx),
+      knownEdges(_ctx),
+      knownPresenceEdges( _ctx ),
+      knownActiveEdges( _ctx ),
+      knownPairingMatrix( _ctx )
+      
+      /** vector version *
+      knownNodesTuple(),
+      knownActiveNodesTuple(),
+      knownEdgesTuple(),
+      knownPresenceEdgesTuple(),
+      knownActiveEdgesTuple(),
+      knownPairingMatrixTuple()
+      */
+      
+  {
+    init_vts();
+  }
+
+private:
+  z3::context& ctx;
+
+  unsigned   M;        // Molecules
+  unsigned   N;       // Nodes
+  unsigned   E_arity;    // Edge arity
+  model_version V;   // Variation
+  unsigned   C;     // connectedness
+  unsigned D;      // CNF conjunction depth 
+  unsigned L;
+
+  void init_vts();
+
+  //variables
+  Vec2Expr nodes;
+  Vec2Expr active_node;
+
+  Vec3Expr node_parameter_var;
+  Vec3Expr edge_parameter_var;
+  Vec3Expr gate_selector_var_node;
+  Vec3Expr gate_selector_var_edge;
+
+  Vec3Expr edges;
+  Vec4Expr presence_edge;
+  Vec4Expr active_edge;
+
+  Vec2Expr pairing_m;
+  Vec4Expr pairing_m_4d;
+  std::vector<z3::func_decl> node_funcs;
+  std::vector<z3::func_decl> edge_funcs;
+
+  Vec4Expr reach;
+  Vec3Expr drop1;
[…]

### 2020-04-05 src/sugar/sugar-encode.cpp
ads

@@ -296,53 +296,53 @@ distinct_mol_pairs( const sugar_mol_ptr& m1, const sugar_mol_ptr& m2 ) {
 }
 
 
-// In a tree v1----------->v2
-// check if there is a repeating subtree rooted at v2
+// // In a tree v1----------->v2
+// // check if there is a repeating subtree rooted at v2
 
-z3::expr sugar_encoding::
-eq_rules( const sugar_mol_ptr& m1, const sugar_mol_ptr& m2) {
-  auto m1_rule_bits = m1->get_cons()->get_rule_bits();
-  auto m2_rule_bits = m2->get_cons()->get_rule_bits();
-  VecExpr eqs;
-  for( unsigned i=0 ; i < mp_rule_bits.size(); i++ ) {
-    eqs.push_back( m1_rule_bits[i] == m2_rule_bits[i] );
-  }
-  return mk_and( ctx, equal);
-}
+// z3::expr sugar_encoding::
+// eq_rules( const sugar_mol_ptr& m1, const sugar_mol_ptr& m2) {
+//   auto m1_rule_bits = m1->get_cons()->get_rule_bits();
+//   auto m2_rule_bits = m2->get_cons()->get_rule_bits();
+//   VecExpr eqs;
+//   for( unsigned i=0 ; i < mp_rule_bits.size(); i++ ) {
+//     eqs.push_back( m1_rule_bits[i] == m2_rule_bits[i] );
+//   }
+//   return mk_and( ctx, equal);
+// }
 
 
-unsigned repeat_search_depth = 3
+// unsigned repeat_search_depth = 3
 
-z3::expr sugar_encoding::
-is_repeat_subtree_in_branch_with_tail( const sugar_mol_ptr& m,
-                                       const sugar_mol_ptr& tail_to_match ) {
-  // No impact return
-  return mk_true( ctx );
-  auto m_rule_bits = m->get_cons()->get_rule_bits();
-  VecExpr no_equal;
-  VecExpr match_sub_tree;
-  const std::vector<sugar_mol*> parent_trail,
-  parent_trail.push_back( m );
-  sugar_mol* mp_head = m->get_parent();
-  unsigned depth = 1;
+// z3::expr sugar_encoding::
+// is_repeat_subtree_in_branch_with_tail( const sugar_mol_ptr& m,
+//                                        const sugar_mol_ptr& tail_to_match ) {
+//   // No impact return
+//   return mk_true( ctx );
[…]

### 2020-10-14 papers/sugar-synth/algo.tex
sadsad

@@
 
-  \caption{Illustrating parts of production rules that are to be encoded in the rule templates.}
+  \caption{Parts of production rules in the rule templates.}
   \label{fig:rule-parts}
@@
 \begin{enumerate}
-\item If a node is present, then it is labeled with a sugar. Otherwise, we ignore the label.
+\item If a node is present, then it is labeled with a sugar.
+  %Otherwise, we ignore the label.
 $$
@@
   $$
-\item The number of children of a node should match with the arity of the labeled sugar.
+\item The children count of a node matches with the arity of the labeled sugar.
   $$
@@
   $$
-\item At least one of the leaves is present. This constraint is for
+\item At least one of the leaves is present. This is for
   avoiding the trivial solution.
@@
   \end{align*}
-  The disequality constraint is
-  \begin{align*}
-    \Lor_{\mathclap{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).
-  \end{align*}
-  In our implementation, we have an option to strengthen
-  the constraints.
-  We have an option to also ignore all the subtrees of
+  The constraint is
+  % \begin{align*}
+  $\Lor_{%\mathclap
+    {m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$
+  % \end{align*}
+  In our implementation, 
+  we have an option to also ignore the subtrees of
   the molecules of $\mu$.

### 2020-10-14 papers/sugar-synth/bio.tex
sadsad

@@
-\begin{figure}[ht!]
-\centering
-\includegraphics[width=0.8\linewidth]{gfig1.png}
-\caption{Biological details of glycan production.
-There are many types of sugar monomer building blocks (each represented by standard colored shapes); for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.
-A glycan is a tree-like sugar tree (nodes linked by edges) attached to a substrate protein at the root (labeled `R').
-Distinct edge orientations correspond to covalent bonds of distinct carbons on the sugar monomer.
-Curved boxes represent reaction compartments within cells, which are the site of glycan production.
-Each step of glycan growth (black arrows) represents the addition of a single new monomer to a specific attachment point on the tree.
-Each such step is catalyzed by an enzyme, labeled $E_i$.
-At any stage of growth, the tree can exit the reaction compartment as an output.
-Alternatively, it can be passed to a subsequent reaction compartment for further growth driven by different enzymes.
-Note that the enzymatic rule is sensitive to the two monomers being linked by a bond, as well as any branches.
-For example, enzyme $E_2$ will add a Galactose to a GalNAc only if the GlcNAc branch is present; otherwise, the reaction will not proceed (`X').
-The structures, reactions, and enzymes shown here are illustrative, and they do not correspond to any measured data set; see the following section for a real example.
-In biological experiments, the combined outputs of every compartment are measured; the underlying reactions must be inferred.
-The inference problem is the subject of this paper.}
-\end{figure}
-
 DNA and RNA are made by copying template DNA, in a process called transcription carried out by an enzyme called RNA polymerase.
@@
 
+\begin{figure}[t]
+  \centering
+  \begin{minipage}{0.54\linewidth}
+    \includegraphics[width=0.9\linewidth]{gfig1.png}    
+  \end{minipage}
+  \begin{minipage}{0.44\linewidth}
+    \caption{Biological details of glycan production.
+There are many types of sugar monomer building blocks (each represented by standard colored shapes); for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
+    \label{fig:glycan-rule}
+  \end{minipage}
+\vspace{-9mm}
+\end{figure}
+
+In Figure~\ref{fig:glycan-rule}, we present details of glycan production.
+A glycan is a tree-like sugar tree (nodes linked by edges) attached to a substrate protein at the root (labeled `R').
+Distinct edge orientations correspond to covalent bonds of distinct carbons on the sugar monomer.
+Curved boxes represent reaction compartments within cells, which are the site of glycan production.
+Each step of glycan growth (black arrows) represents the addition of a single new monomer to a specific attachment point on the tree.
+Each such step is catalyzed by an enzyme, labeled $E_i$.
+At any stage of growth, the tree can exit the reaction compartment as an output.
+Alternatively, it can be passed to a subsequent reaction compartment for further growth driven by different enzymes.
+Note that the enzymatic rule is sensitive to the two monomers being linked by a bond, as well as any branches.
+For example, enzyme $E_2$ will add a Galactose to a GalNAc only if the GlcNAc branch is present; otherwise, the reaction will not proceed (`X').
+The structures, reactions, and enzymes shown here are illustrative, and they do not correspond to any measured data set; see the following section for a real example.
+In biological experiments, the combined outputs of every compartment are measured; the underlying reactions must be inferred.
+%The inference problem is the subject of this paper.
+
 %--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

### 2020-10-14 papers/sugar-synth/experiments.tex
sadsad

@@
 and pass the input parameters of the method.
-
-\subsection{Optimizations}
-We have incorporated several optimizations to gain efficiency or reduce
-the number of iterations in the method. Let us discuss some of the remarkable
-optimizations.
-
-\paragraph{Ordered templates}
-Our method constructs a list of templates.
-Since the templates are symmetric, there is a potential of unnecessary
-iterations.
-Let us suppose the method rejects a set of learned rules.
-In the subsequent queries include the constraints that reject the set of rules,
-the solvers may choose the same set of rules again by permuting the mapping
-from the templates to the learned rule.
-To avoid these iterations, we define an arbitrary total order over the rules.
-We add ordering constraints stating that any assignments to the list of templates should produce
-ordered rules with respect to the total order.
-
-\paragraph{Quantifier instantiations}
-The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
-For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
-to ensure that the query is not unsatisfiable.
-The instantiations sufficiently slow down the solving process.
-We assist the solver by also adding an instantiation of the quantifiers that is equal
-to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
-We observe that for some inputs only adding the instantiations and not the universally quantified
-formula is more effective.
-
-\paragraph{Constraints for counterexample molecules}
-In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
-at line~\ref{line:consNewR} in each iteration.
-The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
-the construction of many terms repeatedly. In our implementation, we construct the constraints
-once outside the while loop.
-We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
-Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
-of the template variables that were returned by the solver at line~\ref{line:posModel}.
-Due to the assignments, the templates become concrete rules in the context of solving.
 

### 2020-10-14 papers/sugar-synth/intro.tex
sadsad

@@
 We also use templates to model the space of the production rules.
-The templates usually have parameters that limit the search space.
-In our case, the size and number of templates are the parameters that
-limit sizes of pieces of trees that are added in one step,
-number of the production rules, and
-the depth of surrounding that are checked to control the rules.
+% The templates usually have parameters that limit the search space.
+% In our case, the size and number of templates are the parameters that
+% limit sizes of pieces of trees that are added in one step,
+% number of the production rules, and
+% the depth of surrounding that are checked to control the rules.
 
-\begin{figure}[t]
-  \begin{tikzpicture}[thick,shorten >=1pt,node distance=2cm,on grid]
-    \node[sqloc]   (q0)                {Synthesis query};
-    \node   (qi) [xshift=-1cm, left=of q0]                {Input glycans};
-    \node[sqloc]           (q1) [xshift=2cm, right=of q0] {Counterexample query};
-    \node   (qo) [ below=of q1]                {Synthesized rules};
-    \node   (qf) [ below=of q0]                {Synthesis failed};
-    \path[->] (q0) edge [bend left=45] node [above] {Synthesized rules} node [below,xshift=-1.2cm,yshift=-3mm] {sat} (q1);
-    \path[->] (q1) edge [bend left=45] node [below] {Counterexample} node [above,xshift=1.2cm,yshift=3.5mm] {sat} (q0);
-    \path[->] (qi) edge  (q0);
-    \path[->] (q1) edge node [right,yshift=6mm] {unsat} (qo);
-    \path[->] (q0) edge node [left,yshift=6mm] {unsat} (qf);  
-  \end{tikzpicture}
-  \caption{The architecture of our synthesis method}
-  \label{fig:illus-method}
-\end{figure}
+% \begin{figure}[t]
+%   \begin{tikzpicture}[thick,shorten >=1pt,node distance=2cm,on grid]
+%     \node[sqloc]   (q0)                {Synthesis query};
+%     \node   (qi) [xshift=-1cm, left=of q0]                {Input glycans};
+%     \node[sqloc]           (q1) [xshift=2cm, right=of q0] {Counterexample query};
+%     \node   (qo) [ below=of q1]                {Synthesized rules};
+%     \node   (qf) [ below=of q0]                {Synthesis failed};
+%     \path[->] (q0) edge [bend left=45] node [above] {Synthesized rules} node [below,xshift=-1.2cm,yshift=-3mm] {sat} (q1);
+%     \path[->] (q1) edge [bend left=45] node [below] {Counterexample} node [above,xshift=1.2cm,yshift=3.5mm] {sat} (q0);
+%     \path[->] (qi) edge  (q0);
+%     \path[->] (q1) edge node [right,yshift=6mm] {unsat} (qo);
+%     \path[->] (q0) edge node [left,yshift=6mm] {unsat} (qf);  
+%   \end{tikzpicture}
+%   \caption{The architecture of our synthesis method}
+%   \label{fig:illus-method}
+% \end{figure}
 
-Our method is iterative, as illustrated in Figure~\ref{fig:illus-method}.
-Our tool first constructs quantifier-free constraints encoding that a set of unknown rules
+Our method is iterative.
+%as illustrated in Figure~\ref{fig:illus-method}.
+Our tool first constructs constraints encoding that a set of unknown rules
 defined by templates can assemble the input set of molecules.
@@
 We call the query to the solver {\em counterexample query}.
-If there is no such molecule, our method terminates and reports
+If there is no such molecule, our method reports
 the synthesized rules.
@@
 We organize the paper as follows.
-In section~\ref{sec:bio}, we introduce the biological background of the synthesis problem and present the basis of the formal model.
-If the reader wants to avoid getting into the biological details, she may skip the section.
+In section~\ref{sec:bio}, we introduce the biological background.
+%of the synthesis problem and present the basis of the formal model.
+% If the reader wants to avoid getting into the biological details, she may skip the section.
 In section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
-In section~\ref{sec:model}, we present the formal model of the glycans, their production
-rules, and the templates.
-In section~\ref{sec:algo}, we present our method of solving the synthesis problem.
-In section~\ref{sec:variations}, we present the variations of the problem and the modifications
-in our methods to support the variations.
-In section~\ref{sec:experiments}, we present our implementation of the method and experiments on
-some glycan data.
+In section~\ref{sec:model}, we present the formal model of the glycans,
+and their production rules.
+In section~\ref{sec:algo}, we present our method for the synthesis problem.
+In section~\ref{sec:variations}, we present the variations of the problem.
+%and the modifications in our methods to support the variations.
+In section~\ref{sec:experiments}, we present our implementation of the method and experiments. %on some glycan data.
 In section~\ref{sec:related} and section~\ref{sec:conclusion},
-we present the related work, future directions, and conclude the paper.
+we present the related work, and conclude the paper.
 

### 2020-10-14 papers/sugar-synth/main.tex
sadsad

@@
+\documentclass{llncs}
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
-\bibliographystyle{ACM-Reference-Format}
+% \bibliographystyle{ACM-Reference-Format}
 %% Citation style
@@
 %% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
+% \citestyle{acmauthoryear}   %% For author/year citations
 
@@
 
+\maketitle
+
 \begin{abstract}
@@
 \end{abstract}
-
-\maketitle
 
@@
 
-% % \bibliographystyle{unsrt}
+\bibliographystyle{unsrt}
 \bibliography{biblio}
 
+\appendix
+
+\section{Implementation optimizations}
+
+% \subsection{Optimizations}
+We have incorporated several optimizations to gain efficiency or reduce
+the number of iterations in the method. Let us discuss some of the remarkable
+optimizations.
+
+\paragraph{Ordered templates}
+Our method constructs a list of templates.
+Since the templates are symmetric, there is a potential of unnecessary
+iterations.
+Let us suppose the method rejects a set of learned rules.
+In the subsequent queries include the constraints that reject the set of rules,
+the solvers may choose the same set of rules again by permuting the mapping
+from the templates to the learned rule.
+To avoid these iterations, we define an arbitrary total order over the rules.
+We add ordering constraints stating that any assignments to the list of templates should produce
+ordered rules with respect to the total order.
+
+\paragraph{Quantifier instantiations}
+The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
+For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
+to ensure that the query is not unsatisfiable.
+The instantiations sufficiently slow down the solving process.
+We assist the solver by also adding an instantiation of the quantifiers that is equal
+to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
+We observe that for some inputs only adding the instantiations and not the universally quantified
+formula is more effective.
+
+\paragraph{Constraints for counterexample molecules}
+In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
+at line~\ref{line:consNewR} in each iteration.
+The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
+the construction of many terms repeatedly. In our implementation, we construct the constraints
+once outside the while loop.
+We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
+Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
+of the template variables that were returned by the solver at line~\ref{line:posModel}.
+Due to the assignments, the templates become concrete rules in the context of solving.
+
+\section{Features to support in future}
+
+\subsection{Full organism data vs single cell data}
+
+The experiments that observe the set of glycan molecules are of two kinds.
+In one kind, we isolate a single cell type of an organism and identify all the glycans present in the cells.
+In our presentation, we have assumed this kind of source of data.
+However, the experiments are difficult to conduct.
+In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
+So the information that which glycans are coming from which
+cells is unknown.
+In this situation, the synthesis has another task to map the molecules
+in $\mu$ to several cell types.
+A glycan may be present in multiple cell types.
+
+Our method is easily adapted to consider this kind of data.
+We search for a small covering set of subsets of
+$\powerset{\mu}$ such that each set in the cover allows a small
+set of production rules.
+The cover sets indicate the different types of cells in terms of
+the presence of glycans.
+This variation makes the problem particularly hard.
+There can be a potentially large number of covering subsets.
+So far, we have not encountered a large enough data set such that
+we apply the variation effectively.
+
+\subsection{Incomplete data}
+
+The experiments are imperfect. They may not detect all possible glycans
+in a cell.
+We may need to leave the possibility of allowing a few more molecules to be
+producible beyond $\mu$.
+We replace the no extra molecule constraint by a constraint that allow molecules
+that are `similar' to molecules in $\mu$.
+However, we could not imagine any measure of similarity that naturally stems from
+biological intuition.
+For now, our tool supports the count of the number of monomer differences
+as a measure of similarity.
+
+\section{Extended related work}
+
+% Success of formal methods
+Formal methods have successful in applications to a vast range of
+problems from the analysis of systems.
+%
+In particular,
+the verification of hardware designs
+using SAT solver based model-checking has been applied
+to industrial-scale problems~\cite{biere1999symbolic2}.
+%
+The verification of software systems is a much harder problem, and
+many methods like bounded model-checking~\cite{biere2003bounded},
+abstract interpretation~\cite{lattice77}, and
+counterexample guided abstraction refinement (CEGAR)~\cite{ClarkeCEGAR} have
+been successful for the problem.
+%
+
+Synthesis of programs that implements a given behavior
+has been a focus of research for a while~\cite{PnueliSynthesis}.
+%
+In recent years, the synthesis of programs using SAT/SMT solvers
+has gained momentum.
+%
+The approach encodes the search of a program that exhibits a certain
+behavior into a satisfiability problem.
+%
+The solvers attempt to find a solution of the satisfiability problem.
+%
+The solution is the synthesized program.
+%
+In~\cite{SrivastavaSynthesis,Solar-Lezama2005},
+the method has been successfully applied
+to find programs that satisfy quantified specifications like sorting
+programs, and have missing ``holes'' and need proper implementations
+for them.
+%
+In~\cite{exampleSynth},
+a set of pairs of input and output examples is the specification of a program,
+and the synthesis method searches for programs in a space defined by a
+template~\cite{sygus}.
+
+% Boolean networks
+
+% Program synthesis
+Since the formal methods have been exhibiting effectiveness in the vast
+range of problems, the
+systems biology community has been applying the methods for various
+biological problems~\cite{fisher2007executable}.
+The formal methods approach is distinctly different from the statistical
+approach traditionally used in biology.
+In formal methods, we aim to match precisely the expected behavior rather
+than learn approximate artifacts from biological data.
+The use of formal methods belongs to two broad categories:
+the analysis of biological models and the synthesis of the models.
+
+The key focus has been to model gene regulatory networks (GRNs), since
+they are the core of the central dogma of biology.
+Boolean networks are often used to model
+GRNs to find stable points or attractors, i.e., nodes in the networks
+where a system eventually reaches no matter where it starts~\cite{wang2012BooleanOverview}.
+% In~\cite{boolean-networks-stabilty}, the networks are used to evaluate
+% stability of GRNs under mutations.
+The Boolean networks are not often sufficient to adequately model the behavior
+of GRNs.
+The continuous-time Markov chains (CTMCs) are used to bring in the aspects of
+timing and probabilistic constraints to GRNs.
+The transient behaviors of GRNs are also crucial for various aspects
+of biology.
+% For example, the decision making in some cells is based on
+% response timing to stimulus~\cite{response-timing-paper}.
+In~\cite{delayedCTMC}, a methods based on
+model-checking is used to estimate the time evolution of the probability
+distributions over states of GRNs.
+
+There are many biological processes where we can observe the system behavior,
+but the exact mechanisms of the processes are not known.
+%
+Recently, the methods for formal synthesis are finding their applications in biology
+\cite{dunn2014defining,xuPluripotency,booleanModelKarp13,paoletti2014analyzing}.
+%
+In~\cite{koksal2013synthesis},
+GRNs with unknown interactions are modeled using
+a automata-like model with missing components
+and the specification is the variations of the behavior of the system under mutations.
+Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
+CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
+the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
+Otherwise, the method learns a new sample where the program violates the specifications.
+It adds the sample to the set and goes to the next iteration.
+Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
+solving the generated queries.
+%
+Our method follows a similar pattern, where
+a set of sampled observations is a specification, i.e., output molecules,
+and we need to find the governing programs, i.e., production rules.
+%
+In a subsequent work~\cite{fisher2015synthesising}, a Boolean network model is used,
+the functions attached to the nodes are considered unknown, and
+a different kind of data sets provides the behavior specification.
+In this approach, their earlier method is adopted to work in the new situation.
+%
+
+As far as we know, there has been no similar computation based analysis of glycan production rules.
+%
+However, there has been a theoretical analysis of the properties of the production
+rules in~\cite{Jaiman2018}. The work identifies the conditions for the production
+of a finite set of molecules and the cases of unique production methods.
+%
+In our work, we are taking the computational approach of synthesis from formal methods
+instead of looking for the theoretical conditions.
+
+% 
+% To resolve this paradox we borrow from the field of
+% al-gorithmic self-assembly, which explores how small buildingblocks
+% with stochastic local interactions assemble into various global
+% configurations [23]–[25].  A central inverse problemin self-assembly
+% is to design building blocks that assembleinto a target shape.  This
+% process has been studied using thetheoretical framework of Wang tiles:
+% square building blockswith colored sides that stick to one another
+% along sides withmatching colors [26].  For efficient assembly of Wang
+% tiles,their coloring must be cleverly chosen so the target shape isthe
+% unique terminal output of the stochastic assembly process[27].   That
+% is, their assembly must be algorithmic.  Glycans may be considered a
+% natural realization of the Wang construct,with monomers acting like
+% tiles whose stickiness is encodedby GTase enzymes.  If glycan
+% biosynthesis could be madealgorithmic, cells could suppress unwanted
+% byproducts andgenerate only desired glycan oligomers with high yield.
+
 \end{document}

### 2020-10-14 papers/sugar-synth/model.tex
sadsad

@@
 
-\begin{example}
-  In Figure~\ref{fig:exrule}(a), we present a rule. That has
-  two kinds of nodes.
-  The rule adds the square node.
-  Therefore, $v_e$ is the square node.
-  The circular nodes $A$ and its right child $B$ are part of a pattern.
-  The pattern must be present in the molecule to apply the rule.
-  In Figure~\ref{fig:exrule}(b), we present an application of the rule.
-  The solid tree with three nodes is the initial molecule.
-  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
-  Upon applying the rule, the rule adds a left child with label $A$ to the middle node.
-  The rule is not applicable at the root $A$, since it has a right child $A$
-  and the combination does not match the pattern.
-\end{example}
-
-\begin{figure}[t]
-  \begin{minipage}{0.48\linewidth}
-    \center
-  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
+\begin{wrapfigure}{r}{0.37\textwidth}
+  \vspace{-5mm}
+    % \begin{minipage}{0.48\linewidth}
+      \small
+    % \center
+  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
     \node[loc] (v0) {$A$};
@@
     \path[->] (v0) edge (v2);
-  \end{tikzpicture}\par
-  (a)
-  \end{minipage}
-  \begin{minipage}{0.48\linewidth}
-        \center
-  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
+  \end{tikzpicture}
+  % \par
+  % (a)
+  % \end{minipage}
+  % \begin{minipage}{0.48\linewidth}
+    % \small
+    % \center
+  \quad
+  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid]
     \node[loc] (v0) {$A$};
@@
   \end{tikzpicture}
-  \par
-  (b)
-  \end{minipage}
-
-  \caption{(a) An example of a rule. (b) An application of the rule.}
+  % \par
+  \hfill (a) \hfill (b)\hfill\mbox{}
+  % (b)
+  % \end{minipage}
+  \caption{(a) An example rule. (b) An application of the rule.}
   \label{fig:exrule}
-\end{figure}
+  \vspace{-8mm}
+\end{wrapfigure}
+\paragraph{Example:} In Figure~\ref{fig:exrule}(a), we present a rule. That has
+  two kinds of nodes.
+  The rule adds the square node.
+  Therefore, $v_e$ is the square node.
+  The circular nodes %$A$ and its right child $B$
+  are part of a pattern.
+  The pattern must be present in the molecule to apply the rule.
+  In Figure~\ref{fig:exrule}(b), we present an application of the rule.
+  The solid tree with three nodes is the initial molecule.
+  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
+  Upon applying the rule, the rule adds a left child with label $A$ to the middle node.
+  The rule is not applicable at the root $A$, since it has a right child $A$
+  and the combination does not match the pattern.
 

### 2020-10-14 papers/sugar-synth/related.tex
sadsad

@@
 Formal methods have successful in applications to a vast range of
-problems from the analysis of systems.
-%
-In particular,
-the verification of hardware designs
-using SAT solver based model-checking has been applied
-to industrial-scale problems~\cite{biere1999symbolic2}.
-%
-The verification of software systems is a much harder problem, and
-many methods like bounded model-checking~\cite{biere2003bounded},
-abstract interpretation~\cite{lattice77}, and
-counterexample guided abstraction refinement (CEGAR)~\cite{ClarkeCEGAR} have
-been successful for the problem.
-%
-
+problems from the analysis of
+systems~\cite{biere1999symbolic2,biere2003bounded,lattice77,ClarkeCEGAR}.
 Synthesis of programs that implements a given behavior
 has been a focus of research for a while~\cite{PnueliSynthesis}.
-%
 In recent years, the synthesis of programs using SAT/SMT solvers
-has gained momentum.
+has gained momentum~\cite{SrivastavaSynthesis,Solar-Lezama2005,sygus}.
 %
@@
 The solution is the synthesized program.
-%
-In~\cite{SrivastavaSynthesis,Solar-Lezama2005},
-the method has been successfully applied
-to find programs that satisfy quantified specifications like sorting
-programs, and have missing ``holes'' and need proper implementations
-for them.
-%
-In~\cite{exampleSynth},
-a set of pairs of input and output examples is the specification of a program,
-and the synthesis method searches for programs in a space defined by a
-template~\cite{sygus}.
 
@@
 biological problems~\cite{fisher2007executable}.
-The formal methods approach is distinctly different from the statistical
-approach traditionally used in biology.
-In formal methods, we aim to match precisely the expected behavior rather
-than learn approximate artifacts from biological data.
-The use of formal methods belongs to two broad categories:
-the analysis of biological models and the synthesis of the models.
-
-The key focus has been to model gene regulatory networks (GRNs), since
-they are the core of the central dogma of biology.
-Boolean networks are often used to model
-GRNs to find stable points or attractors, i.e., nodes in the networks
-where a system eventually reaches no matter where it starts~\cite{wang2012BooleanOverview}.
-% In~\cite{boolean-networks-stabilty}, the networks are used to evaluate
-% stability of GRNs under mutations.
-The Boolean networks are not often sufficient to adequately model the behavior
-of GRNs.
-The continuous-time Markov chains (CTMCs) are used to bring in the aspects of
-timing and probabilistic constraints to GRNs.
-The transient behaviors of GRNs are also crucial for various aspects
-of biology.
-% For example, the decision making in some cells is based on
-% response timing to stimulus~\cite{response-timing-paper}.
-In~\cite{delayedCTMC}, a methods based on
-model-checking is used to estimate the time evolution of the probability
-distributions over states of GRNs.
+The key focus has been to model gene regulatory networks (GRNs)~\cite{wang2012BooleanOverview,delayedCTMC}.
 
@@
 %
-Recently, the methods for formal synthesis are finding their applications in biology
-\cite{dunn2014defining, xuPluripotency, booleanModelKarp13, paoletti2014analyzing}.
+Recently, the methods for formal synthesis are finding their applications in biology~\cite{dunn2014defining,xuPluripotency,booleanModelKarp13,paoletti2014analyzing,koksal2013synthesis,fisher2015synthesising}.
 %
-In~\cite{koksal2013synthesis},
-GRNs with unknown interactions are modeled using
-a automata-like model with missing components
-and the specification is the variations of the behavior of the system under mutations.
-Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
-CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
-the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
-Otherwise, the method learns a new sample where the program violates the specifications.
-It adds the sample to the set and goes to the next iteration.
-Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
-solving the generated queries.
-%
-Our method follows a similar pattern, where
-a set of sampled observations is a specification, i.e., output molecules,
-and we need to find the governing programs, i.e., production rules.
-%
-In a subsequent work~\cite{fisher2015synthesising}, a Boolean network model is used,
-the functions attached to the nodes are considered unknown, and
-a different kind of data sets provides the behavior specification.
-In this approach, their earlier method is adopted to work in the new situation.
-%
+As far as we know, there has been no similar computation based analysis of glycan production rules.
 
-As far as we know, there has been no similar computation based analysis of glycan production rules.
-%
-However, there has been a theoretical analysis of the properties of the production
-rules in~\cite{Jaiman2018}. The work identifies the conditions for the production
-of a finite set of molecules and the cases of unique production methods.
-%
-In our work, we are taking the computational approach of synthesis from formal methods
-instead of looking for the theoretical conditions.
-
-% 
-% To resolve this paradox we borrow from the field of
-% al-gorithmic self-assembly, which explores how small buildingblocks
-% with stochastic local interactions assemble into various global
-% configurations [23]–[25].  A central inverse problemin self-assembly
-% is to design building blocks that assembleinto a target shape.  This
-% process has been studied using thetheoretical framework of Wang tiles:
-% square building blockswith colored sides that stick to one another
-% along sides withmatching colors [26].  For efficient assembly of Wang
-% tiles,their coloring must be cleverly chosen so the target shape isthe
-% unique terminal output of the stochastic assembly process[27].   That
-% is, their assembly must be algorithmic.  Glycans may be considered a
-% natural realization of the Wang construct,with monomers acting like
-% tiles whose stickiness is encodedby GTase enzymes.  If glycan
-% biosynthesis could be madealgorithmic, cells could suppress unwanted
-% byproducts andgenerate only desired glycan oligomers with high yield.
+% In~\cite{},
+% GRNs with unknown interactions are modeled using
+% a automata-like model with missing components
+% and the specification is the variations of the behavior of the system under mutations.
+% Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
+% CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
+% the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
+% Otherwise, the method learns a new sample where the program violates the specifications.
+% It adds the sample to the set and goes to the next iteration.
+% Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
+% solving the generated queries.
+% %
+% Our method follows a similar pattern, where
+% a set of sampled observations is a specification, i.e., output molecules,
+% and we need to find the governing programs, i.e., production rules.
+% %
+% In a subsequent work~\cite{}, a Boolean network model is used,
+% the functions attached to the nodes are considered unknown, and
+% a different kind of data sets provides the behavior specification.
+% In this approach, their earlier method is adopted to work in the new situation.
+% %
 

### 2020-10-14 papers/sugar-synth/variations.tex
sadsad

@@
 However, the biological problem has more details.
-There are unsettled modeling choices.
-We have developed and implemented several variations of the
+% There are unsettled modeling choices.
+We have developed % and implemented
+several variations of the
 above method.
-The user may choose the variation depending on the choice of model.
+% A user may choose the variation depending on the choice of model.
 
@@
 \end{algorithmic}
-After adding the above constraints, we can divide the learned rules
+Using the above constraints, we can divide the learned rules
 into compartments.
@@
 
-\subsection{Full organism data vs single cell data}
-
-The experiments that observe the set of glycan molecules are of two kinds.
-In one kind, we isolate a single cell type of an organism and identify all the glycans present in the cells.
-In our presentation, we have assumed this kind of source of data.
-However, the experiments are difficult to conduct.
-In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
-So the information that which glycans are coming from which
-cells is unknown.
-In this situation, the synthesis has another task to map the molecules
-in $\mu$ to several cell types.
-A glycan may be present in multiple cell types.
-
-Our method is easily adapted to consider this kind of data.
-We search for a small covering set of subsets of
-$\powerset{\mu}$ such that each set in the cover allows a small
-set of production rules.
-The cover sets indicate the different types of cells in terms of
-the presence of glycans.
-This variation makes the problem particularly hard.
-There can be a potentially large number of covering subsets.
-So far, we have not encountered a large enough data set such that
-we apply the variation effectively.
-
-\subsection{Incomplete data}
-
-The experiments are imperfect. They may not detect all possible glycans
-in a cell.
-We may need to leave the possibility of allowing a few more molecules to be
-producible beyond $\mu$.
-We replace the no extra molecule constraint by a constraint that allow molecules
-that are `similar' to molecules in $\mu$.
-However, we could not imagine any measure of similarity that naturally stems from
-biological intuition.
-For now, our tool supports the count of the number of monomer differences
-as a measure of similarity.
-
 %--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

### 2020-10-14 papers/sugar-synth/algo.tex
many changes

@@
 Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
-At a high level, our method is iterative.
-The method first synthesizes a set of rules $R$ that can produce the molecules in input set $\mu$, i.e., $\mu \subseteq P(S,R)$.
-However, there is also an additional requirement that a molecule
-that is not in $\mu$ is not producible by the synthesized rules $R$.
-Therefore, the method looks for the producible molecules that are not in $\mu$, called
-{\em counterexample molecule}.
-If there is no such molecule, our method terminates and reports
-the synthesized rules.
-If there is a molecule in $P(S,R) - \mu$, we add constraints that says
-that look for rules
-that do not generate
-the molecule and go for another iteration.
-Now let us look at the details of the method.
+
+% At a high level, our method is iterative.
+% The method first synthesizes a set of rules $R$ that can produce the molecules in input set $\mu$, i.e., $\mu \subseteq P(S,R)$.
+% However, there is also an additional requirement that a molecule
+% that is not in $\mu$ is not producible by the synthesized rules $R$.
+% Therefore, the method looks for the producible molecules that are not in $\mu$, called
+% {\em counterexample molecule}.
+% If there is no such molecule, our method terminates and reports
+% the synthesized rules.
+% If there is a molecule in $P(S,R) - \mu$, we add constraints that says
+% that look for rules
+% that do not generate
+% the molecule and go for another iteration.
+% Now let us look at the details of the method.
 
@@
 
-The method assumes that the input set $\mu$ is finite, which a
-natural assumption.
+The method assumes that the input set $\mu$ is finite.
 Even if a set of rules can produce
-an unboundedly large number of molecules.
-No real-life biology will exhibit such a set in a cell.
-Since we are synthesizing production rules, our method
-needs to bound the search space.
+an unbounded number of molecules.
+No biology will exhibit such a set in a cell.
+Our method
+bounds the search space of production rules.
 The method also takes two numbers as input:
@@
 and 
-$n$ is the maximum number of the learned rules.
-The parameters eliminate potential
-unbounded search space.
+$n$ is the maximum number of them.
+% The parameters eliminate potential
+% unbounded search space.
 If the method fails to find production rules,
 the user can always call the method with larger parameters.
-However, we do not expect to find a large number of or sized rules
-because chemical reactions do not usually have many conditional
-on the precursors and do not apply large changes in a single step.
+% However, we do not expect to find a large number of or sized rules
+% because chemical reactions do not usually have many conditional
+% on the precursors and do not apply large changes in a single step.
 First, the method initializes $S$ with the set of monomers occurring
@@
 We use templates to model the search space of rules.
-A template is also a tree that has a certain depth and
+A template is also a tree that has a depth and
 the internal nodes of the tree have the same number of children.
@@
 the `situation' of the node.
-The first variable at the node encodes choices of
-sugar monomers or their absence.
-The domain of the variables is $S \union \{\bot\}$.
+% The first variable at the node encodes choices of
+% sugar monomers or their absence.
+The domain of the first variables is $S \union \{\bot\}$.
 Let $SVars$ be the set of uncountable variables with the domain.
@@
 
+\begin{wrapfigure}{r}{0.37\textwidth}
+    \vspace{-6mm}
+        \center
+  \begin{tikzpicture}[shorten >=1pt,thick,node distance=1cm,on grid,scale=0.5]
+    \node[loc] (v0) {$v_0$};
+    \node[sqloc, below right of=v0] (v1) {$v_e$};
+    \draw [->] plot [smooth] coordinates { (v0.south)
+          ($(v0.south)+(3mm,-2mm)$)
+          ($(v0.south)+(10mm,1mm)$)
+          % ($(v0.south)+(9mm,-6mm)$)
+          % ($(v0.east)+(8mm,-8mm)$)
+          (v1.north) };
+    % \path[->] (v0) edge (v1);
+    % \path[->,dashed] (v1) edge (v2);
+    % \path[->] (v1) edge (v3);
+    \fill[fill=gray,opacity=0.2] 
+    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
+    \fill[fill=gray,opacity=0.4] 
+    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
+    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
+  \end{tikzpicture}
+  \caption{Parts of production rules in the rule templates.}
+  \label{fig:rule-parts}
+  \vspace{-8mm}
+\end{wrapfigure}
 A node in a production rule can be in four situations.
-In Figure~\ref{fig:rule-parts}, we illustrate the situations of the nodes in a rule.
+In Figure~\ref{fig:rule-parts}, we illustrate the situations.
 The first situation is when a node position may be in the expanding part.
-The nodes in the dark gray area are the expanding nodes.
+The dark gray nodes are the expanding nodes.
 The second situation is when a node position may not be present in the rule.
@@
 Formally, the templates are defined as follows.
-
-\begin{figure}[t]
-        \center
-  \begin{tikzpicture}[shorten >=1pt,thick,node distance=2cm,on grid]
-    \node[loc] (v0) {$v_0$};
-    \node[sqloc, below right of=v0] (v1) {$v_e$};
-    \draw [->] plot [smooth] coordinates { (v0.south)
-          ($(v0.south)+(3mm,-2mm)$)
-          ($(v0.south)+(2mm,-6mm)$)
-          ($(v0.south)+(9mm,-6mm)$)
-          % ($(v0.east)+(8mm,-8mm)$)
-          (v1.north) };
-    % \path[->] (v0) edge (v1);
-    % \path[->,dashed] (v1) edge (v2);
-    % \path[->] (v1) edge (v3);
-    \fill[fill=gray,opacity=0.2] 
-    (0,0) -- ++(-1.5cm,-1.7cm) -- ++(1.2cm,-1.2cm) -- ++(6cm,0cm) -- cycle;
-    \fill[fill=gray,opacity=0.4] 
-    (v1.center) -- ++(-1.5cm,-1.5cm) -- ++(3cm,0cm) -- cycle;
-    \draw[dashed] (-1.5cm,-1.7cm) -- ++(-1.2cm,-1.2cm) -- ++(2.4cm,0cm) -- cycle;
-  \end{tikzpicture}
-
-  \caption{Parts of production rules in the rule templates.}
-  \label{fig:rule-parts}
-\end{figure}
 
@@
 Let $v'$ be the $i$th ancestor at some iteration of the loop.
-At lines~\ref{line:encodep-ans-match} and~\ref{line:encodep-ans-order}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
+At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
 declared $MatchAns$, sugar matches in $v_r$ and $v'$, and

### 2020-10-14 papers/sugar-synth/bio.tex
many changes

@@
   \begin{minipage}{0.44\linewidth}
-    \caption{Biological details of glycan production.
-There are many types of sugar monomer building blocks (each represented by standard colored shapes); for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
+    \caption{Biological details of glycan production. There are many types of sugar monomer building blocks; %(each represented by standard colored shapes);
+      for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
     \label{fig:glycan-rule}

### 2020-10-14 papers/sugar-synth/experiments.tex
many changes

@@
 
-\subsection{Implementation} 
+{\em Implementation:} 
 We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool} for
@@
 
-\subsection{Benchmark}
-
+{\em Benchmark: }
 We have applied our tool on four sets of molecules form
@@
 
-\subsection{Results}
 \input{tbl-results}
 
-We have applied \ourtool~on the data set. For each data set, we choose several
+{\em Results:} We have applied \ourtool~on the data set. For each data set, we choose several
 parameter combinations to illustrate the relative performance of the tool.

### 2020-10-14 papers/sugar-synth/main.tex
many changes

@@
 
-\title{Automated inference of production rules for glycans}
-\author{Ashutosh Gupta and Mukund Thattai and Ansuman Biswas}
+\title{%Automated inference
+  Synthesis of production rules for glycans}
+\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
 %\institute{IITB and NCBS}
@@
 Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks.
-The rules by which these enzymes operate are not fully understood.
-In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans found in a cell.
-We have build a tool based on this method and have used it to infer rules based on published glycan data sets.
+The rules by which the enzymes operate are not fully understood.
+In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell.
+We have build a tool based on the method and applied it to infer rules based on published glycan data.
 \end{abstract}

### 2020-10-14 papers/sugar-synth/model.tex
many changes

@@
 In this section, we present the formal model for the synthesis problem.
-We model glycan molecules as labeled trees and production rules are also modeled using small pieces of trees. The glycan molecules are assembled by applying the production rules repeatedly.
+We model glycan molecules and production rules as labeled trees. The glycan molecules are assembled by applying the production rules repeatedly.
 Our synthesis problem reduces into finding the pieces of trees that represent the production rules.
@@
 the oligomer molecules. Each $s \in S$ is associated with arity $m$ 
-(written $arity(s) = m$), {\em i.e.}, the maximum number of children for $s$. The children of the monomers are indexed, which means we can refer to the $k$th child of $s$ for some $k \leq arity(s)$.
-In the monomers, they correspond to bonds at specific positions from where children are connected.
+(written $arity(s) = m$).
+% , {\em i.e.}, the maximum number of children for $s$.
+The children of the monomers are indexed. We refer to the $k$th child of $s$ for some $k \leq arity(s)$.
+They correspond to bonds at specific positions in the monomers where children are connected.
 Now we define the glycan molecules as labeled trees. Now onward we refer to the glycans simply as molecules.
@@
 $C : V \times \naturals \pmaps V$ maps the indexed children of nodes, and
-$v_0 \in V$ is the root of the tree.
-Furthermore, a molecule must satisfy the conditions of the arity of monomers.
-Formally, let $M(v) = s$, if $C(v,n) = v'$ then $n \leq arity(s)$.
+$v_0 \in V$ is the root of $m$.
+A molecule must respect the arity of monomers, i.e., if $M(v) = s$ and $C(v,n) = v'$ then $n \leq arity(s)$.
 \end{df}
@@
 For each $v \in m$, if $(v,n)$ is not in the domain of $C$, we write $C(v,n) = \bot$.
-We assume that $C(v,0) = \bot$ and if $n > n' > 0$ and $C(v,n) \neq \bot$, then $C(v,n') \neq \bot$.
+We assume that $C(v,0) = \bot$.
+%and if $n > n' > 0$ and $C(v,n) \neq \bot$, then $C(v,n') \neq \bot$.
 Let $NumberOfChildren(v)$ be equal to the number of $n$s such that $C(v,n) \neq \bot$. 
@@
 
-Now let us define the model of the production process of the molecules.
+Now we define the model of the production process of the molecules.
 A production rule expands a molecule $m$ by attaching a new piece of tree at a node that has a vacant spot among its children if the surroundings of the
@@
   A {\em production rule} $r = (V, M, C, v_0, v_e)$ is a labeled tree, where
-  $V$ is a set of nodes in the tree,
-  $M : V \maps S$ maps nodes to their label, 
+  $V$ is a set of nodes,
+  $M : V \maps S$ maps nodes to labels, 
   $C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
-  $v_0 \in V$ is the root of the tree, and
+  $v_0 \in V$ is the root, and
   $v_e \in V$ is the root of expanding part of the rule.
@@
 If we {\em apply}, a rule $r$ on a molecule $m$, then it is extended at some node
-$v \in m$. A copy of the descendants of $v_e$ will be be attached to
+$v \in m$. A copy of the descendants of $v_e$ will be attached to
 $v$ in $m$, and the rest of the nodes in the rule have to match $v$ and above.
-We will refer to the descendants of $v_e$ as {\em expanding nodes}
+We call the descendants of $v_e$ as {\em expanding nodes}
 and all the other nodes as {\em matching nodes}.
@@
 \begin{wrapfigure}{r}{0.37\textwidth}
-  \vspace{-5mm}
+  \vspace{-12mm}
     % \begin{minipage}{0.48\linewidth}
@@
   % \end{minipage}
-  \caption{(a) An example rule. (b) An application of the rule.}
+  \caption{(a) A rule. (b) An application of the rule.}
   \label{fig:exrule}
@@
 \end{wrapfigure}
-\paragraph{Example:} In Figure~\ref{fig:exrule}(a), we present a rule. That has
+\paragraph{Example:} In Figure~\ref{fig:exrule}(a), we present a rule. It has
   two kinds of nodes.
-  The rule adds the square node.
-  Therefore, $v_e$ is the square node.
+  The rule adds the square node ($v_e$).
   The circular nodes %$A$ and its right child $B$
-  are part of a pattern.
-  The pattern must be present in the molecule to apply the rule.
+  are the pattern, which must be present in the molecule to apply the rule.
   In Figure~\ref{fig:exrule}(b), we present an application of the rule.
@@
   The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
-  Upon applying the rule, the rule adds a left child with label $A$ to the middle node.
-  The rule is not applicable at the root $A$, since it has a right child $A$
-  and the combination does not match the pattern.
+  %Upon applying the rule,
+  The rule adds a left child with label $A$ to the middle node.
+  The rule is not applicable at the root $A$ due to pattern mismatch.
+  % , since it has a right child $A$
+  % and the combination does not match the pattern.
 
@@
 \subsection{Synthesis problem}
-As discussed in section~\ref{sec:bio},
-we observe a set of glycan molecules present in a cell.
+We observe a set of glycan molecules $\mu$ present in a cell.
 %
@@
 %
-Let us suppose we are given a set of molecules $\mu$ produced by some
-unknown set of production rules.
-%
 The {\em synthesis problem} is to find a set $R$ of production rules
@@
 
-%
-We can define a more general form of the synthesis problem.
-Let us suppose we are also given $k$ compartments with unknown rules.
-The goal of synthesis is to find  $R_1,...,R_k$ such that
-$\mu = P(S, R_1,..,R_k)$.
-We may relax the requirement of exactly producing $\mu$.
-We may say that molecules in $\mu$ have to be produced, but we are ok
-if subtrees of the molecules of $\mu$ are also produced.
-Formally, we may weaken the requirement to $\mu \subseteq P(Q, R_1,..,R_k) \subseteq subtree(\mu)$.
+% %
+% We can define a more general form of the synthesis problem.
+% Let us suppose we are also given $k$ compartments with unknown rules.
+% The goal of synthesis is to find  $R_1,...,R_k$ such that
+% $\mu = P(S, R_1,..,R_k)$.
+% We may relax the requirement of exactly producing $\mu$.
+% We may say that molecules in $\mu$ have to be produced, but we are ok
+% if subtrees of the molecules of $\mu$ are also produced.
+% Formally, we may weaken the requirement to $\mu \subseteq P(Q, R_1,..,R_k) \subseteq subtree(\mu)$.
 
-The above are a few simplified versions of the biological problem.
-We may generalize the problem further where
-%rules are partitioned and the partitions are applied in phases one after another,
-rules are not applied exhaustively due to time constraints,
-the given $\mu$ is finite and may not be exhaustive,
-and we only know the weights of the parts of molecules in $\mu$.
-We will first present a method for solving a simplified version of the problem.
-However, our tool handles some of the above variations.
-We will discuss the variations in section~\ref{sec:variations}.
+% The above are a few simplified versions of the biological problem.
+% We may generalize the problem further where
+% %rules are partitioned and the partitions are applied in phases one after another,
+% rules are not applied exhaustively due to time constraints,
+% the given $\mu$ is finite and may not be exhaustive,
+% and we only know the weights of the parts of molecules in $\mu$.
+% We will first present a method for solving a simplified version of the problem.
+% However, our tool handles some of the above variations.
+% We will discuss the variations in section~\ref{sec:variations}.
 

### 2020-10-14 papers/sugar-synth/motivation.tex
sadsad
many changes

@@
 \includegraphics[width=0.8\linewidth]{gfig2.png}
-\caption{An example of a glycan data set. Figure credit: Anjali Jaiman, PhD thesis.}
+\caption{A glycan data set. Figure credit: Anjali Jaiman, PhD thesis.}
 \label{fig:dataset-gly}
+\vspace{-6mm}
 \end{figure}
@@
 In this section, we first present a motivating example to illustrate our method.
-In Figure~\ref{fig:dataset-gly}, we show the glycan oligomers associated with human chorionic
+In Figure~\ref{fig:dataset-gly},
+we consider the glycan oligomers associated with human chorionic
 gonadotropin~\cite{Harrd1992}.
@@
 
-Now we may consider the abstract general problem. In Figure~\ref{fig:ex-gly}(a), we present a set of
-glycan molecules present in a cell consists of three molecules, which are structurally similar to the
-three glycan molecules in Figure~\ref{fig:dataset-gly}.
-To keep things simple for illustration, we have dropped the third glycan molecule
-(inside light blue square) from the four in
-Figure~\ref{fig:dataset-gly}.
-%
+Now we consider the synthesis problem. In Figure~\ref{fig:ex-gly}(a), we present a set of
+glycan molecules present in a cell consists of three molecules, which are structurally similar to the three glycan molecules in Figure~\ref{fig:dataset-gly}.
+To keep illustration simple, we dropped the third glycan molecule.
 The molecules contain four types of monomers.
@@
 
-Before considering the synthesis problem that produces exactly
-the three molecules,
-let us consider a set of rules in Figure~\ref{fig:ex-gly}(b)
+% Before considering the synthesis problem that produces exactly
+% the three molecules,
+Let us first consider six rules in Figure~\ref{fig:ex-gly}(b)
 that produce the molecules.
-There are six rules.
 All the rules are in the same compartment.
-% The first compartment has four rules and the second compartment has three rules.
 The rules have two kinds of nodes.
-If the circular nodes are present around a leaf,
-the rule is enabled and appends the molecule at the leaf with the square nodes.
-In Figure~\ref{fig:ex-gly}(c), we show the steps of generating the middle
+If the circular nodes are present around a node,
+the rule is enabled and may append the molecule at the
+node with the square nodes.
+In Figure~\ref{fig:ex-gly}(c), we show the steps of generating the last
 glycan molecule.
@@
 The last step looks at the two ancestors before adding a single node.
-We may avoid the need for compartments if the operations can add multiple nodes at
-a time or look for a deep context for being enabled.
+% We may avoid the need for compartments if the operations can add multiple nodes at
+% a time or look for a deep context for being enabled.
 
@@
 
-Furthermore, there is another interesting case occurs in this example.
-Please look at the last two rules.
-The second last rule adds $C\rightarrow B$ branch.
-The last rule checks if $C\rightarrow B$ nodes are present then adds $C$.
-If both rules were adding one node at a time without checking deep context,
-then the undesired molecule pattern in Figure~\ref{fig:ex-gly}(e) is
-also producible.
-We would have needed compartments to avoid the undesirable molecule.
-Therefore, there is a tread-off between rule sizes and the number of compartments.
+% Furthermore, there is another interesting case occurs in this example.
+% Please look at the last two rules.
+% The second last rule adds $C\rightarrow B$ branch.
+% The last rule checks if $C\rightarrow B$ nodes are present then adds $C$.
+% If both rules were adding one node at a time without checking deep context,
+% then the undesired molecule pattern in Figure~\ref{fig:ex-gly}(e) is
+% also producible.
+% We would have needed compartments to avoid the undesirable molecule.
+% Therefore, there is a tread-off between rule sizes and the number of compartments.
 

### 2020-10-14 papers/sugar-synth/variations.tex
many changes

@@
 Similarly, we also need to modify the function
-$\textsc{EncodeP}$ by replacing line~\ref{line:encodep-ans-order}
-by the following line, which also inserts extra constraints.
+$\textsc{EncodeP}$ by inserting the following line after~\ref{line:encodep-ans-match}.
 \begin{algorithmic}[1]
@@
   \setcounterref{ALG@line}{line:encodep-ans-match}
-  \State $c \landplus \tau(v') < mark \land compart(v_r) \geq compart(v')$
+  \State $c \landplus compart(v_r) \geq compart(v')$
   \vspace{1ex}

### 2020-10-15 src/sugar/sugar-encode.cpp
variations

@@ -357,7 +357,7 @@ z3::expr sugar_encoding::match( const sugar_mol_ptr& root1, const sugar_mol_ptr&
     auto m1_rule_bits = root1->get_cons()->get_rule_bits();
     auto m2_rule_bits = root2->get_cons()->get_rule_bits();
     for( unsigned i=0 ; i < m1_rule_bits.size(); i++ ) {
-        v1.push_back( m1_rule_bits[i] == m2_rule_bits[i] ); // rule same for runaway, comparrtment same for runaway, compare at each level
+        v1.push_back( (m1_rule_bits[i] == m2_rule_bits[i]) && !rs.at(i)->get_is_fast() ); // rule same for runaway, comparrtment same for runaway, compare at each level
       }
     for(unsigned i = 0; i < uu->get_num_sugar_bits(); i++ ){
       v1.push_back(uu->get_sugar_bit(i)==uv->get_sugar_bit(i));
@@ -476,7 +476,7 @@ z3::expr sugar_encoding::compare(const sugar_mol_ptr& m1, const sugar_mol_ptr& m
           if(j!=d-1){s.push(mp->get_sibling_num());}          
           rep.push_back(m4->get_cons()->get_compartment()==mp->get_cons()->get_compartment());
           if(j!=d-1){mp=mp->get_parent(); }
-      }    
+      }  
       rep.push_back(match( std::make_shared<sugar_mol>(*mp),m3,m4));     
       while(!s.empty()){
       if(m4==nullptr){return mk_false(ctx);}
@@ -796,7 +796,7 @@ match_tree(  rule_ptr r, sugar_mol_ptr m, z3::expr mark, bool is_match_part,
   match.push_back( z3::implies( r->get_occur_cons(), s_bit && tcons && ccons));
   if(b){
     if (mark){
-      match.push_back( z3::implies(  !r->get_occur_cons() && r_parent->get_occur_cons() && r->get_is_hard_end(),  m->get_cons()->get_tstamp() > mark  ));  
+    match.push_back( z3::implies(  !r->get_occur_cons() && r_parent->get_occur_cons() && r->get_is_hard_end(),  m->get_cons()->get_tstamp() > mark  ));  
     }    
   }  
   for( unsigned i = 0; i < m->get_children_num(); i++ ) {    
@@ -901,7 +901,7 @@ no_fast_match_cons( sugar_mol* m_parent, unsigned m_child_num,
[…]

### 2020-10-15 src/sugar/sugar-encode.h

@@ -159,7 +159,8 @@ private:
   z3::expr no_match_cons( const sugar_mol_ptr& m  );
   void no_fast_can_extend_unknown(const sugar_mol_ptr& m,VecExpr& no_matches);
   void no_fast_can_extend(const sugar_mol_ptr& m,VecExpr& no_matches);
-
+  unsigned int depth_of(const sugar_mol_ptr& m);
+  bool partial(const sugar_mol_ptr& m1, const sugar_mol_ptr& m2);
   // reading solutions
   sugar_mol_ptr read_neg_mol( sugar_mol_ptr&, z3::model&, expr_set& );
   sugar_t_ptr read_sugar( const VecExpr& sugar_bits, z3::model& m );

### 2020-10-15 papers/sugar-synth/intro.tex
many changes
17 pages

@@
 We have applied the tool on data sets from published sources (these are available in the database UniCarbKB \cite{Campbell2013}).
-We have obtained the rules for the data sets.
-The outputs are within the expectations of biological
+% We have obtained the rules for the data sets.
+The output rules are within the expectations of biological
 intuition.
@@
 %and the modifications in our methods to support the variations.
-In section~\ref{sec:experiments}, we present our implementation of the method and experiments. %on some glycan data.
-In section~\ref{sec:related} and section~\ref{sec:conclusion},
-we present the related work, and conclude the paper.
+In section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
+% In section~\ref{sec:related} and section~\ref{sec:conclusion},
+% we present the related work, and conclude the paper.
+In section~\ref{sec:conclusion}, we conclude the paper.
 

### 2020-10-15 papers/sugar-synth/motivation.tex
17 pages

@@
 If we allow an arbitrary number of rules,
-the rules to look at their context up to
-an arbitrary depth, and an arbitrarily large number of compartments,
+and the rules to look at their context up to
+an arbitrary depth, % and an arbitrarily large number of compartments,
 then there is a trivial solution.
-Therefore, our method needs three parameters.
-The first parameter limits the number of rules.
-The second parameter limits the maximum height of the rule trees.
-The third parameter limits the maximum number of compartments.
-If we give larger values to the parameters, then we have more chances to
-find the production rules.
-For this illustration, we searched for the production rules with
-seven as the limit on the number of rules, three as the limit on the rule heights,
-and a single compartment.
+Therefore, our method limits the number and size of rules.
+% parameters.
+% The first parameter limits the number of rules.
+% The second parameter limits the maximum height of the rule trees.
+% The third parameter limits the maximum number of compartments.
+% If we give larger values to the parameters, then we have more chances to
+% find the production rules.
+For this illustration, we searched for the seven production rules with
+three as the limit on the rule heights.
+All rules are in single compartment.
 
@@
 We call a solver to solve the synthesis query.
-After the first query, we obtain the rules presented in Figure~\ref{fig:ex-gly}(f).
-Next, we look for another molecule that may be produced by the rule.
-Using another query to the solver, we find a counterexample molecule that is
-producible by the rules but is not in the input set.
-Coincidently, we have already presented the counterexample molecule in Figure~\ref{fig:ex-gly}(e).
-Afterword, we add constraints that the molecule must not
-be producible by the synthesized rules in our synthesis query and go for the next iteration.
-After 19 iterations, our tool synthesizes a set of rules that satisfies the requirements.
+After the first query, we obtain the rules presented in Figure~\ref{fig:ex-gly}(e).
+The rule set can produce molecules that are not are not in input.
+% Next, we look for another molecule that may be produced by the rule.
+% Using another query to the solver, we find a counterexample molecule that is
+% producible by the rules but is not in the input set.
+% Coincidently, we have already presented the counterexample molecule in Figure~\ref{fig:ex-gly}(e).
+% Afterword, we add constraints that the molecule must not
+% be producible by the synthesized rules in our synthesis query and
+% go for the next iteration.
+We need to iterate further.
+After 19 iterations, our tool synthesizes a set of rules that
+satisfies the requirements.
 

### 2020-10-15 papers/sugar-synth/variations.tex
variations.tex

@@
 
-\subsection{Compartment stay model}
+% \subsection{Compartment stay model}
 
-When molecules pass through compartments, they are expected to stay
-for a while in the compartments.
-%
-The length of the stay determines the number of rule applications
-occurring on the molecule.
-%
-In our method, we assume that the molecules may stay any length of time
-in a compartment, which may result in any number of applications of the rules.
-%
-% This allows the maximum number of molecules are produced for a given set of rules.
-%
-We can assume a whole range of {\em stay models} for the synthesis.
-%
-We have considered one more stay model, where a molecule stays
-in the compartment until no more rule applications are possible.
-%
-We call the stay model as {\em infinite stay}.
+% When molecules pass through compartments, they are expected to stay
+% for a while in the compartments.
+% %
+% The length of the stay determines the number of rule applications
+% occurring on the molecule.
+% %
+% In our method, we assume that the molecules may stay any length of time
+% in a compartment, which may result in any number of applications of the rules.
+% %
+% % This allows the maximum number of molecules are produced for a given set of rules.
+% %
+% We can assume a whole range of {\em stay models} for the synthesis.
+% %
+% We have considered one more stay model, where a molecule stays
+% in the compartment until no more rule applications are possible.
+% %
+% We call the stay model as {\em infinite stay}.
 
-To encode the infinite stay model, we need to add constraints that
-say when a molecule goes to the next compartment, no rule of
-the current compartment is applicable.
-%
-We add the following constraints in \textsc{EncodeProduce}.
-$$
-\Land_{v\in m}\Land_{i \text{ such that } C(v,i) \neq \bot}\Land_{t \in T}
-compart( v ) \leq compart(t) < compart(C(v,i)) 
-\limplies \Land noMatch(C(v,i),t)
-$$
-where constraint $noMatch(v,t)$ is collected in \textsc{EncodeP}
-by inserting the following code after the while loop and an update on $c$ before making the last two updates on $c$.
-$noMatch(v,t)$ is $\ltrue$
-before the call to \textsc{EncodeP}.
+% To encode the infinite stay model, we need to add constraints that
+% say when a molecule goes to the next compartment, no rule of
+% the current compartment is applicable.
+% %
+% We add the following constraints in \textsc{EncodeProduce}.
+% $$
+% \Land_{v\in m}\Land_{i \text{ such that } C(v,i) \neq \bot}\Land_{t \in T}
+% compart( v ) \leq compart(t) < compart(C(v,i)) 
+% \limplies \Land noMatch(C(v,i),t)
+% $$
+% where constraint $noMatch(v,t)$ is collected in \textsc{EncodeP}
+% by inserting the following code after the while loop and an update on $c$ before making the last two updates on $c$.
+% $noMatch(v,t)$ is $\ltrue$
+% before the call to \textsc{EncodeP}.
+% \begin{algorithmic}[1]
+%   \vspace{1ex}
+%   \setcounterref{ALG@line}{line:encodep-vr-expand}
+%   \State $nomatch(v,t) \landplus \lnot c$
+%   \vspace{1ex}
+% \end{algorithmic}
+% The constraints state that for each node $v$ and template rule $t$
+% if a rule adds node $v$ before or at the compartment of $t$
+% and its $i$th child after the compartment of $t$,
+% then $t$ must not be applicable at the $i$th child.
+% The above constraints effectively encode that the molecule
+% cannot expand at some node until all the relevant rules on
+% earlier compartments are disabled.
+% Note that if there is no child at the $i$th node, we are adding no constraints of disabling rules of future compartments.
+% One may also add those as a requirement, depending on
+% the interpretation of the stay model.
+
+\subsection{Fast and slow reactions}
+During the stay of a molecule in a compartment, it could be observed that some reactions were dominating than others.
+This can be explained by characterizing reactions as either slow or fast.
+A slow reaction can occur only when no other fast reaction is able to extend the molecule in that compartment.
+We can now define \textsc{Extend} recursively as follows:
+\begin{align*}
+  \textsc{Extends}(r, m) :=\;&   Apply(m,r) \lor
+  \Lor_{\mathclap{i \in [1,w]}} \textsc{Extends}( r, C(m,i ) )\\
+    \textsc{Extends}(r, \bot) :=\;&  \nu(v) \neq \bot
+  \end{align*}
+We now modify $\textsc{EncodeP}$ after~\ref{line:mcut}:
 \begin{algorithmic}[1]
   \vspace{1ex}
-  \setcounterref{ALG@line}{line:encodep-vr-expand}
-  \State $nomatch(v,t) \landplus \lnot c$
+  \setcounterref{ALG@line}{line:mcut}
+  \State $c \landplus \lnot \textsc{Fast}(v_r) \implies 
+  \Land_{t \in T} (\textsc{Fast}(t) \implies \lnot\textsc{Extends}(t,v))$
   \vspace{1ex}
 \end{algorithmic}
-The constraints state that for each node $v$ and template rule $t$
-if a rule adds node $v$ before or at the compartment of $t$
-and its $i$th child after the compartment of $t$,
-then $t$ must not be applicable at the $i$th child.
-The above constraints effectively encode that the molecule
-cannot expand at some node until all the relevant rules on
-earlier compartments are disabled.
-Note that if there is no child at the $i$th node, we are adding no constraints of disabling rules of future compartments.
-One may also add those as a requirement, depending on
-the interpretation of the stay model.
+Here, \textsc{Fast} sets the constraint on the rule to be fast.
+A negative molecule which is a proper subtree of an input molecule will cease to be negative if there is any fast reaction that is able to extend it as fast reactions happen aggressively and can make partial molecules complete.
+Therefore, we modify the function $\textsc{SugarSynth}$ after~\ref{line:encode-neg-mol}:
+\begin{algorithmic}[1]
+  \vspace{1ex}
+  \setcounterref{ALG@line}{line:encode-neg-mol}
+  \State $nCons \landplus
+  \lnot  \Lor_{r \in R} \textsc{Fast}(r) \land \textsc{Extends}(r,m')$
+  \vspace{1ex}
+\end{algorithmic}
+Here, \textsc{Fast} constraints the rule to be fast. The constraints state that none of the rules should be both fast and able to extend the negative molecule received from the model.
 
 \subsection{Unbounded molecules}
-
 We always observe a finite set of molecules. However,
@@
 an unboundedly large size of molecules.
-If due to some biological intuition, we believe that a subtree in a molecule
-will form a repeating pattern by attaching a copy of itself to
-one of its leaves.
-We pass such information to our method.
-We add constraints to our method to encode that
-the last compartment consists of rules for the repeating pattern,
-the production rules add a new copy of the subtree only after looking
-within the context of the previous copy of the subtree,
-and once the rules for repeating pattern are active
-all the other rules of the last compartment must not be applicable.
+In such cases, rules are allowed to form molecules having repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules.
+Constraints are added to encode that those rules are all applied in the same compartment and are slow reactions.
+We put a limit on the depth of such repeating subtrees ($d_0$) and the number of repetitions ($r_0$) .
+Then, for a particular depth $d$ and repetitions $r$, we recursively define \textsc{Repeat} and \textsc{ExactMatch} as follows:
+\begin{align*}
+  \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
+    \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
+    \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
+    \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
+    \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
+    \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
+    \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
+    \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
+    \textsc{ExactMatch}(m,\bot) :=\;& \bot 
+  \end{align*}
+  where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
+The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.
+\begin{algorithmic}[1]
+  \vspace{1ex}
+  \setcounterref{ALG@line}{line:consNewR}
+  \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
+  \vspace{1ex}
+\end{algorithmic}
+where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
+
+\subsection{Non-monotonic rules}
+Some molecules produced dictate that after a certain rule is applied to them in a compartment, it restricts the application of any other rule from that point of time.
+Such a rule is said to have \textsc{HardEnds}.
+This feature of rules help in producing an exact set of desired molecules.
+We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
+\begin{algorithmic}[1]
+  \vspace{1ex}
+  \setcounterref{ALG@line}{line:matchtree-cons}
+  \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
+  \vspace{1ex}
+\end{algorithmic}
+In our method, \textsc{HardEnds} sets a constraint on the given rule to have hard ends.
+The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.
 

### 2020-10-15 papers/sugar-synth/algo.tex

@@
 \begin{wrapfigure}{r}{0.37\textwidth}
-    \vspace{-6mm}
+    \vspace{-12mm}
         \center
@@
 In Figure~\ref{fig:rule-parts}, we illustrate the situations.
-The first situation is when a node position may be in the expanding part.
+The first situation is when a node position is in the expanding part.
 The dark gray nodes are the expanding nodes.
-The second situation is when a node position may not be present in the rule.
-The dashed area includes absent nodes
-from the full tree.
-Among the matching nodes, we make two cases, namely if the expanding
-nodes are the descendants of the matching node or not. 
+The second situation is when a node position is not in the rule.
+The dashed area are the absent nodes.
+Among the matching nodes, we have two cases.
+% , namely if the expanding
+% nodes are the descendants of the matching node or not.
 The third situation is when a node position is in the matching part
 and has expanding decadents.
-The solid path from $v_0$ to the root $v_e$ of the expanding part includes
-the nodes that are ancestors of expanding part among the matching nodes.
-Finally, the fourth situation is when a node position is in the matching part and
-has no expanding decedents.
-The nodes in the rest of the light gray area are in the fourth situation.
-Another variable is mapped to a node to encode the four situations.
+The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
+% the nodes that are ancestors of expanding part among the matching nodes.
+Finally, the fourth situation is the rest of node positions
+in the matching part, which is the light gray area.
+% The nodes in the rest of the light gray area are in the fourth situation.
+A variable is mapped to a node to encode the four situations.
 Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
@@
 Let $KVars$ be the set of uncountable variables with domain $K$. 
-We will use the pool of $KVars$ to add variables to the templates.
-Formally, the templates are defined as follows.
+% We will use the pool of $KVars$ to add variables to the templates.
+The templates are defined as follows.
 
@@
 \end{df}
-In the method at line~\ref{line:createRtemp}, we use the
-call to \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
-height $d$ and children width $w$. We get $d$ and $n$ as input, and
-we let $w$ be the maximum arity of any sugar that occurs in $\mu$.
-This allows us to map any node to any sugar.
-We also assume that templates
-returned by $\textsc{MakeTemplatesRule}$ do not share variables.
-We obtain a production rule from a template by assigning values
-to the variables.
-If for a node $v$ in a template if we assign $\kappa(v) = Absent$,
+In the algorithm at line~\ref{line:createRtemp}, we call \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
+height $d$ and children width $w$.
+% We get $d$ and $n$ as input, and
+% we let $w$ be the maximum arity of any sugar that occurs in $\mu$.
+Since $w$ is the maximum arity of any sugar, we can map any node to any sugar.
+% We also assume that templates
+% returned by $\textsc{MakeTemplatesRule}$ do not share variables.
+% We obtain a production rule from a template by assigning values
+% to the variables of the template.
+For a node $v$ in a template if we assign $\kappa(v) = Absent$,
 we call the node {\em absent}. Otherwise, we call the node {\em present}.
-Not all the assignments may result in a valid rule.
+% Not all the assignments may result in a valid rule.
 Next at line~\ref{line:ruleCorr}, we will construct constraints
 that encode the set of valid rules. 
-
 A valid assignment to the variables in
@@
 satisfy the following six conditions.
-The first two encode the basic agreement among variables.
-The third encodes the tree structure of the rule.
-The last three encode the relative positions of the four
-parts of the rule.
+% The first two encode the basic agreement among variables.
+% The third encodes the tree structure of the rule.
+% The last three encode the relative positions of the four
+% parts of the rule.
 \begin{enumerate}
-\item If a node is present, then it is labeled with a sugar.
+\item If a node is present, then it is labeled with a sugar.\\
   %Otherwise, we ignore the label.
-$$
+$
 \Land_{s \in S} \Land_{v \in V} (\kappa(v) \neq Absent \limplies \nu(v) \neq \bot )
-$$
-\item  The number of children of a node should match with the arity of the labeled sugar.
-  The following formula states that the children that are at greater arity than that of the label
-  must be absent.
-$$
+$
+\item
+  % The number of children of a node should match with the arity of the labeled sugar.
+  % The following formula states that
+
+  The children that are at greater arity than that of the label
+  are absent.\\
+$
 \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
-$$
-\item If a node is present, then the parent of the node is also present.
-  $$
+$
+\item If a node is present, then the parent of the node is also present.\\
+  $
   \Land_{{\text{internal node } v \in V}}
   \Land_{i\in[1,w]} (\kappa(C(v,i) ) \neq Absent \limplies \kappa( v ) \neq Absent )  
-  $$
-\item If a node is $Expand$, then its children are also $Expand$ if present. The following constraints state that once one reached the expanding part, there
-are no descendants that are matching.
-  $$
-\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Expand \limplies \kappa(C(v,i)) \in \{Expand,Absent\}   )
-$$
-\item If a node is $Match$, then its children are also $Match$ if present.
-  Similarly, the following constraints state that once one reached the matching part without expanding descendants, there are no descendants that are expanding.
-$$
-\Land_{i\in[1,w]} \Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = Match \limplies \kappa(C(v,i)) \in \{Match,Absent\}   )
-$$
-\item If a node is $MatchAns$, then exactly one child is $MatchAns$ or $Expand$.
-  The following constraints encode that there is exactly one subtree that is expanding part
-  by stating that there is no branching in $MatchAns$ nodes and they form a path.  
-$$
-\Land_{\mathrlap{\text{internal node } v \in V}} (\kappa(v) = MatchAns \limplies \sum_{i\in[1,w]}( \kappa(C(v,i)) \in \{MatchAns,Expand\}) = 1
-$$
+  $
+\item If a node is $Expand$, then its children are also $Expand$ if present.\\
+  % The following constraints state that once one reached the expanding part, there are no descendants that are matching.
+  $
+\Land_{i\in[1,w]} \Land_{{\text{internal node } v \in V}} (\kappa(v) = Expand \limplies \kappa(C(v,i)) \in \{Expand,Absent\}   )
+$
+\item If a node is $Match$, then its children are also $Match$ if present.\\
+  % Similarly, the following constraints state that once one reached the matching part without expanding descendants, there are no descendants that are expanding.
+$
+\Land_{i\in[1,w]} \Land_{{\text{internal node } v \in V}} (\kappa(v) = Match \limplies \kappa(C(v,i)) \in \{Match,Absent\}   )
+$
+\item If a node is $MatchAns$, then exactly one child is $MatchAns$ or $Expand$.\\
+  % The following constraints encode that there is exactly one subtree that is expanding part
+  % by stating that there is no branching in $MatchAns$ nodes and they form a path.
+\mbox{}\hspace{-5mm}  
+$
+\Land_{v \in V} (\kappa(v) = MatchAns \limplies \sum_{i\in[1,w]}( \kappa(C(v,i)) \in \{MatchAns,Expand\}) = 1
+$
 \end{enumerate}
 
-In the method, the call to $\textsc{RuleTemplateCorrectness}$ at
+The call to $\textsc{RuleTemplateCorrectness}$ at
 line~\ref{line:ruleCorr} creates the above constraints
@@
 Therefore, we need to define the search space for the molecules.
-We use similar templates for defining the search space.
-Since we always search in bounded space, we limit the template size using a parameter,
+We use templates for defining the search space.
+We limit the template size using a parameter,
 namely the height of the template.
@@
 a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
-every internal node has $w$ children, where 
+each internal node has $w$ children, where 
 $V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
@@
 and
-$v_{0m} \in V$ is the root of the tree.
+$v_{0m} \in V$ is the root.
 \end{df}
@@
 Our method searches for unwanted producible molecules up to the height of $h$,
-which we choose to be the maximum of the heights of molecules in $\mu$.
+which we set to the maximum of the heights of molecules in $\mu$.
 %
-The choice of $h$ is somewhat arbitrary.
-%
-We assume if the rules could have produced a molecule with height less than $h$,
-then it must be in $\mu$.
-On the other hand, we ignore the potential formation of the large size of molecules.
-%
-Since we observe only finite molecules in any wet experiment, we may assume
-there may be large producible molecules, but we did not observe them.
-However, this point is debatable and $h$ may be viewed as a parameter of the whole
-method.
+The choice of $h$ is arbitrary.
+% %
+% We assume if the rules could have produced a molecule with height less than $h$,
+% then it must be in $\mu$.
+% On the other hand, 
+% We ignore the potential formation of the large size of molecules.
+% %
+% Since we observe only finite molecules in any wet experiment, we may assume
+% there may be large producible molecules, but we did not observe them.
+% However, this point is debatable and $h$ may be viewed as
+% a parameter of the whole method.
 
@@
 are valid.
-We add the following four conditions for
+We add the following conditions for
 valid assignments for molecule template $\hat{m} = (V, \nu, C, \hat{v_0})$.
-The first  two conditions encode the tree structure of the molecule and
-matching the number of children with label arity for each node.
-The second encodes that the height of the molecule is large enough.
-The last encodes that new molecule must not be the same as any molecule in $\mu$.
+% The first  two conditions encode the tree structure of the molecule and
+% matching the number of children with label arity for each node.
+% The second encodes that the height of the molecule is large enough.
+% The last encodes that new molecule must not be the same as any molecule in $\mu$.
 \begin{enumerate}
-\item If a node is present, then the parent of the nodes is also present.
-  $$
-  \Land_{i\in[1,w]}\Land_{\mathrlap{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
-  $$
-\item The children count of a node matches with the arity of the labeled sugar.
-  $$
+\item If a node is present, then the parent of the nodes is also present.\\
+  $
+  \Land_{i\in[1,w]}\Land_{{\text{internal node } v \in V}} (\nu(C(v,i) ) \neq \bot \limplies \nu( v ) \neq \bot )  
+  $
+\item The children count of a node matches with the arity of the labeled sugar.\\
+  $
   \Land_{s \in S} \Land_{i \in (arity(s),w] }\Land_{v \in V} (\nu(v) = s \limplies \kappa( C(v,i) ) = Absent )
-  $$
-\item At least one of the leaves is present. This is for
-  avoiding the trivial solution.
-  $$
-  \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
-  $$
-  where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
-  This allows us to look for all molecules that are close
-  to height $h$.
-  We choose $d' = h-d$, where $d$ is the height of the rule templates.
-  This $d'$ ensures that we do not miss producible molecules that
-  can not have exact height $h$ but is slightly larger than $h$.
-  Then, there must be another producible
-  molecule that is a subtree of the molecule and has height in
-  between $h$ and $h-d$
-  because the maximum depth of a rule is $d$.
+  $
+% \item At least one of the leaves is present. This is for
+%   avoiding the trivial solution.\\
+%   $
+%   \Lor_{ v \in V \text{ and } depth(v) > d' } \nu( v ) \neq \bot,
+%   $
+%   where $d'$ is a number close to and smaller than $height(\hat{v_0})$.
+%   This allows us to look for all molecules that are close
+%   to height $h$.
+%   We choose $d' = h-d$, where $d$ is the height of the rule templates.
+  % This $d'$ ensures that we do not miss producible molecules that
+  % can not have exact height $h$ but is slightly larger than $h$.
+  % Then, there must be another producible
+  % molecule that is a subtree of the molecule and has height in
+  % between $h$ and $h-d$
+  % because the maximum depth of a rule is $d$.
 \item We find a molecule that is not in $\mu$.
-  To encode the constraints, let predicate $Neq$ be recursively defined as follows.
-  \begin{align*}
-  Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
-  \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
-    Neq(v,\bot) :=\;&  \nu(v) \neq \bot
-  \end{align*}
-  The constraint is
+  We encode $\Lor_{{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$, where
+   predicate $Neq$ be recursively defined as follows.\\
+  $Neq(v, v') :=\; \nu(v) \neq M(v') \lor
+  \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) )$\\
+  $Neq(v,\bot) :=\;  \nu(v) \neq \bot$
   % \begin{align*}
-  $\Lor_{%\mathclap
-    {m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$
+  % Neq(v, v') :=\;&   \nu(v) \neq M(v') \lor
+  % \Lor_{\mathclap{i \in [1,w]}} Neq( C(v,i), C(v',i ) )\\
+  %   Neq(v,\bot) :=\;&  \nu(v) \neq \bot
   % \end{align*}
-  In our implementation, 
-  we have an option to also ignore the subtrees of
-  the molecules of $\mu$.
+  % The constraint is
+  % % \begin{align*}
+  % % \end{align*}
+  % In our implementation, 
+  % we have an option to also ignore the subtrees of
+  % the molecules of $\mu$.
 \end{enumerate}
@@
 Let us continue with~\textsc{SugarSynth}.
-At~\ref{line:molenc} line, we call $\textsc{EncodeProduce}$ for
+At line~\ref{line:molenc}, we call $\textsc{EncodeProduce}$ for
 each molecule in $\mu$ and generate constraints $pCons$ stating that
@@
 It first checks the satisfiability of conjunction
-$tCons \land pCons \land nCons$
-using some SMT solver, where $nCons$ is $\ltrue$ in the first iteration
+$tCons \land pCons \land nCons$, where $nCons$ is $\ltrue$ in the first iteration
 and will encode constraints related to counterexample molecules.
@@
 If it is satisfiable, it constructs the rules at line~\ref{line:getR} from
-the satisfying assignment $a$
+%the satisfying assignment $a$
 and stores them in $R$.
@@
 If it is not satisfiable, we have found the rules that generate exactly
-the molecules in $\mu$ within our search range of counterexample molecules,
-and the loop terminates.
+the molecules in $\mu$ and the loop terminates.
 Otherwise, we use the satisfying assignment to create a
@@
 At line~\ref{line:negCons}, we add constraints to $nCons$ stating that
-all possible applications of templates rules in $T$ must not produce $m'$.
+all possible applications of template rules in $T$ must not produce $m'$.
 We use shorthand $F \landplus G$ for $ F := F \land G$.
@@
 Since we negate the returned formula by $\textsc{EnocdeProduce}$ 
-and then we check the satisfiability, the encoding will not be correct
-due to the fresh variables.
-Therefore, we will need to introduce universal quantifiers to the
+and then we check the satisfiability.
+% the encoding will not be correct
+% due to the fresh variables.
+Therefore, we need to introduce universal quantifiers to the
 fresh variables.
@@
 variables before adding to $nCons$.
-Intuitively, the universal quantifiers say that all possible
-application patterns of the learned rules do not produce $m'$.
-Afterword, the loop goes to the next iteration.
+% Intuitively, the universal quantifiers say that all possible
+% application patterns of the learned rules do not produce $m'$.
+Next, the loop goes to the next iteration.
 Since the domain of all the variables is finite, eventually the loop must
@@
 
-Now let us look at the encoding generated by $\textsc{EncodeProduce}$
-in detail.
+Now let us look at the encoding generated by $\textsc{EncodeProduce}$.
 The process of production of molecules adds pieces of trees of nodes
@@
 the order of the application of the rules.
-The encoding is like encoding bounded executions of programs.
-
+% The encoding is like encoding bounded executions of programs.
 To model the production due to the application of the rules,
-we need to attach three maps to the nodes of the molecules.
+we attach three maps to the molecule nodes.
 \begin{itemize}
 \item Let $cuts$ map each node to a Boolean variable
-indicating the node is the point where a rule is applied and the
-molecule was expanded.
+indicating the node is the point where a rule is applied to
+expand the molecule.
 We say points of the applications of the production
 rules as {\em cuts} of the tree.
-All the nodes that are in between the cuts are added in a single step.
+% All the nodes that are in between the cuts are added in a single step.
 
@@
   indicating the time point when the node was added to the molecule.
-  The order of the addition is an important constraint since already
-  existing nodes in a molecule decide what can be added in the future.
-  If we do not capture the order in our encoding, we may wrongly declare
-  that a molecule is producible while it is not due to the matching
-  constraints.
+  Since already added nodes in a molecule decide what can be added later,
+  we need to record the order of the addition.
+  % If we do not capture the order in our encoding, we may wrongly declare
+  % that a molecule is producible while it is not due to the matching
+  % constraints.
 \end{itemize}
@@
 Both the inputs can be template or concrete.
-The function is tolerant of the variations, but our presentation
-assumes that the molecule is concrete and the rules are templates.
+% The function is tolerant of the variations, but
+Our presentation assumes that the molecule is concrete and
+the rules are templates.
 This will cover the case when the $\textsc{EncodeProduce}$ is called at
@@
 However, at line \ref{line:consNewR} the molecule is a template
-and the rules are concrete.
-We will discuss the handling of this situation at the end of this subsection.
-$\textsc{EncodeProduce}$ is implemented with the help of three other functions,
+and the rules are concrete, which we will discuss later.
+% We will discuss the handling of this situation at the end of this subsection.
+$\textsc{EncodeProduce}$ uses help of three other supporting functions,
 $\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
-All of them are presented in algorithm~\ref{alg:produce}.
+% All of them are presented in algorithm~\ref{alg:produce}.
 %
@@
 $\textsc{EncodeProduce}$ returns constraints stating for each node $v$
-and rule template $t$ that
+and rule template $t$,
 if $v$ is at a cut and $t$ is applied at $v$, then $t$ must match
@@
 We require $rmatch(v)$ to be equal to some rule.
-Therefore, if $v$ is a cut node, some template must be applied. 
+% Therefore, if $v$ is a cut node, some template must be applied. 
 Since we are matching with template rule, we do not know the
-position of the head of expanding nodes.
+position of the root of expanding nodes.
 We enumerate to all possible depths from $1$ to $d-1$ for finding
-the head.
+the root.
 For each $\ell \in [1,d)$, we call $\textsc{EncodeP}( v, t, \ell)$
@@
 
-In $\textsc{EncodeP}$,
-we need to match rule template $t$ such that the root of the expanding
-nodes matches with $v$.
-Since we are given a guess $\ell$ for the depth of the
-node.
-We can traverse up from $v$ for $\ell$ steps to find the node to match
-the root of $t$.
-Naturally, the $\textsc{EncodeP}$ ignores returns $\lfalse$ if there
-is no $\ell$th ancestor of $v$, i.e., there is no possibility of a match.
-The variable $mark$ is assigned the timestamp for node $v$.
-During matching, we will add constraints stating that the matching part of the
-template should match to the nodes added before $mark$
-and expanding part of the template should match with nodes that
-are added at or after $mark$.
+In $\textsc{EncodeP}$, we can traverse up from $v$ for $\ell$ steps to
+find the node to match the root of $t$.
+Naturally, it returns $\lfalse$ if there
+is no $\ell$th ancestor of $v$. % i.e., there is no match.
+The variable $mark$ is the timestamp for node $v$.
+% During matching, we will add constraints stating that the matching part of the
+% template should match to the nodes added before $mark$
+% and expanding part of the template should match with nodes that
+% are added at or after $mark$.
 The local variable $c$ collects a conjunction of constraints as they are
-generated in the function and $c$ is initially $\ltrue$.
+generated. % and $c$ is initially $\ltrue$.
 The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
@@
 matches all the ancestors up to the parent of $v$.
-As loop traverses down, it also traverses $t$ along the matching
+As the loop traverses down, it also traverses $t$ along the matching
 using variable $v_r$.
@@
 \ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
-if the ancestors of $v$ also traverse along with the $j$th child
-(see if condition at line \ref{line:encodep-jth-child-cond}).
+if the ancestors of $v$ also traverse along with the $j$th child.
+% (see if condition at line \ref{line:encodep-jth-child-cond}).
 
-Let us consider the body of the loop in detail.
+% Let us consider the body of the loop.
 Let $v'$ be the $i$th ancestor at some iteration of the loop.
-At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that the corresponding node $v_r$ in the template rule is
-declared $MatchAns$, sugar matches in $v_r$ and $v'$, and
+At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that
+the corresponding node $v_r$ in the template rule is
+$MatchAns$, sugar matches in $v_r$ and $v'$, and
 node $v'$ is added before $mark$.
-The for loop at \ref{line:encodep-sub-for-loop}, iterates over
+The loop at line~\ref{line:encodep-sub-for-loop}, iterates over
 children of $v'$.
@@
 nodes match.
-$\textsc{MatchTree}$ is also called with a third parameter
-$mark$ and a Boolean value as the fourth parameter.
-We will discuss the procedure shortly.
+We will discuss $\textsc{MatchTree}$ shortly.
+% $\textsc{MatchTree}$ is also called with a third parameter
+% $mark$ and a Boolean value as the fourth parameter.
+% We will discuss the procedure shortly.
 If the $j$th child of $v'$ is along the path to $v$,
-we get the node name $v'_r$ for updating
+we get node $v'_r$ for updating
 $v_r$ for the next iteration of the while loop.
-After the while loop in $\textsc{EncodeP}$, $v_r$ is the
-candidate node for the root of expanding nodes.
-At line \ref{line:encodep-vr-expand}, we declare that $v_r$
-is an $Expand$ node.
-Afterword, we match the node $v$
+% After the while loop in $\textsc{EncodeP}$, $v_r$ is the
+% candidate node for the root of expanding nodes.
+After the while loop in $\textsc{EncodeP}$ at line \ref{line:encodep-vr-expand}, we declare $v_r$
+is $Expand$,  match the node $v$
 with the corresponding node $v_r$ in the template rule by calling
-$\textsc{MatchTree}$ again at line~\ref{line:encodep-mtree-expand}.
-At line~\ref{line:mcut}, we also call to match the cut pattern at
+$\textsc{MatchTree}$, and also call to match the cut pattern at
 the subtree of $v$ with the template rule.
@@
 line~\ref{line:matchtree-vr-absent}.
-If the molecule node $v$ does not exist, then the rule node must also
-be flagged absent. Otherwise, the match will fail.
-The corresponding constraints $\kappa(v_r) = Absent$ is added
-at line~\ref{line:matchtree-v-absent}.
-If both the nodes are present, then we add constraints
-that if $v_r$ is not flagged to absent in the template, then
+At line~\ref{line:matchtree-v-absent}, we encode if the molecule node $v$ does not exist, then the rule node must also
+be flagged absent. %Otherwise, the match will fail.
+Otherwise, we add constraints
+that if $v_r$ is not absent in the template, then
 the labels of $v$ and $v_r$ must match at line~\ref{line:matchtree-cons}.
-At the same line, $\textsc{MatchTree}$ also adds constraints that
+At the same line, $\textsc{MatchTree}$ also adds constraints $tCons$ that
 nodes are added in the molecule in the correct order.
-The third input of the function is $mark$, which is the time when
-node $v$ was added. 
-The function also takes $isExpand$ bit as the fourth input if
-$isExpand$ is true then the subtree must be added after
-or with $v$, and is
-false then nodes must be added before $v$.
+The last two inputs of the function is timestamp $mark$ and a bit $isExpand$,
+which tells us that the matching nodes should be added before or after $mark$.
 The calls to $\textsc{MatchTree}$ from \textsc{EncodeP} use the
-$isExpand$ appropriately to indicate that nodes correspond to the
-matching part of the rule are added before and expanding part nodes
-are added later.
+$isExpand$ appropriately.
+% to indicate that nodes correspond to the
+% matching part of the rule are added before and expanding part nodes
+% are added later.
 Afterwords at line~\ref{line:matchtree-recurse},
-the procedure calls itself for the children of $v$ and $v_r$ to
-generate the same constraints for the subtree.
+the procedure calls itself for the children of $v$ and $v_r$.
+% to generate the same constraints for the subtree.
 
@@
 is absent or not.
-If $v$ does not exist, then there is no constraint to add therefore
-the procedure returns $\ltrue$ at line~\ref{line:mcut-no-v}.
-If $v_r$ does not exist, then it returns constraints encoding
-that if the parent of $v_r$ was not absent, then $v$ must be at a cut at line~\ref{line:mcut-no-vr}.
-If both $v$ and $v_r$ exist, then we add constraints
-that if the parent of $v_r$ is not absent, $v_r$ is
-flagged to be absent in the learned rule
-if and only if $v$ is the cut points at line~\ref{line:mcut-cons}.
-$\textsc{MatchCuts}$ recursively traverses the children of $v$ and $v_r$
-and generates the constraints for the cut pattern.
-The constraints added by the procedure are guarded by the presence of parents
-because we need to enable constraints only when there is a possibility of transition
-from presence to absence.
-However, the call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
-Even if the parent of $v_r$ at line \ref{line:mcut} in \textsc{EncodeP} is present,
+% If $v$ does not exist, then there is no constraint to add therefore
+% the procedure returns $\ltrue$ at line~\ref{line:mcut-no-v}.
+% If $v_r$ does not exist, then it returns constraints encoding
+% that if the parent of $v_r$ was not absent, then $v$ must be at a cut at line~\ref{line:mcut-no-vr}.
+% If both $v$ and $v_r$ exist, then we add constraints
+% that if the parent of $v_r$ is not absent, $v_r$ is
+% flagged to be absent in the learned rule
+% if and only if $v$ is the cut points at line~\ref{line:mcut-cons}.
+% $\textsc{MatchCuts}$ recursively traverses the children of $v$ and $v_r$
+% and generates the constraints for the cut pattern.
+% The constraints added by the procedure are guarded by the presence of parents
+% because we need to enable constraints only when there is a possibility of transition
+% from presence to absence.
+The call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
+Even if the parent of $v_r$ in \textsc{EncodeP} is present,
 $v_r$ is a cut point adding cut pattern constraints for the node will create inconsistency.

### 2020-10-15 papers/sugar-synth/algo.tex
variations first cut

@@
 \begin{wrapfigure}{r}{0.37\textwidth}
-    \vspace{-12mm}
+    \vspace{-3mm}
         \center
@@
 \item We find a molecule that is not in $\mu$.
-  We encode $\Lor_{{m = (\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v).$, where
+  We encode $\Land_{{(\_,\_,\_, v_0)\in \mu}}  Neq(\hat{v_0},v_0)$, where
    predicate $Neq$ be recursively defined as follows.\\
@@
 The call to $\textsc{MatchCuts}$ in \textsc{EncodeP} passes $\lfalse$ as the third input.
-Even if the parent of $v_r$ in \textsc{EncodeP} is present,
+Even if the parent of $v_r$ is present,
 $v_r$ is a cut point adding cut pattern constraints for the node will create inconsistency.

### 2020-10-15 papers/sugar-synth/variations.tex
variations first cut

@@
-In the previous section, we have presented a simplified version of
-the synthesis problem.
-However, the biological problem has more details.
-% There are unsettled modeling choices.
-We have developed % and implemented
-several variations of the
-above method.
-% A user may choose the variation depending on the choice of model.
+We have presented a simplified version of
+the biological problem.
+% However, the biological problem has more details.
+We have developed variations of the above method to support more realistic problems.
 
-\subsection{Compartments}
+% \subsection{Compartments}
 
-The production rules may live in different compartments.
-The compartments are ordered.
+{\em Compartments : } The production rules may live in different ordered compartments.
+% The compartments are ordered.
 A molecule moves from one compartment to the next.
 The rules of the current compartment apply to the molecule.
-We have presented the formal description of the compartments in
-section~\ref{sec:model}.
+% We have presented the formal description of the compartments in
+% section~\ref{sec:model}.
 To support the compartments, we take an additional integer
@@
 of compartments.
-We construct each template rule with a new variable with domain $[1,k]$
-indicating the compartment of the rule.
-Let $v_r$ be a node of some template rule. We will write $compart(v_r)$
-for the compartment variable for the template rule.
+We construct each template rule with a new variable with domain $[1,k]$.
+Let $v_r$ be a node of some template rule. We write $compart(v_r)$
+for the compartment variable for the template.
 
@@
 ensure that they enforce the compartment order.
-Essentially, we need to encode that a rule is applied when all the
+We need to encode that a rule is applied when all the
 pattern nodes were added in the current or earlier compartment.
-For the encoding, we add a new map $compart$ that maps
-nodes to variables with domain $[1,k]$.
+% For the encoding, we add a new map $compart$ that maps
+% nodes to variables with domain $[1,k]$.
 We modify the function $\textsc{MatchTree}$ by replacing
-the {\bf if} by the following code.
+$tCons$ assignment by the following code.\\
+\begin{minipage}{1.0\linewidth}
 \begin{algorithmic}[1]
-  \vspace{1ex}
   \setcounterref{ALG@line}{line:matchtree-v-absent}
-  \If{$isExpand$}
-   \State $tCons := ( mark \leq \tau(v)  \land compart(v_r) = compart(v) )$
-  \Else
-   \State $tCons := ( \tau(v) < mark  \land compart(v_r) \geq compart(v) )$
-   \EndIf
-   \vspace{1ex}
+  \State $tCons := isExpand \;?\; ( mark \leq \tau(v)  \land compart(v_r) = compart(v) )$\par
+  \mbox{}\qquad\qquad\hspace{10mm} $:( \tau(v) < mark  \land compart(v_r) \geq compart(v) )$
 \end{algorithmic}
-In the above code, we have inserted constraints related to compartments
-as discussed earlier.
-Similarly, we also need to modify the function
-$\textsc{EncodeP}$ by inserting the following line after~\ref{line:encodep-ans-match}.
+\end{minipage}
+Similarly, we modify the function
+$\textsc{EncodeP}$ by inserting the following line after~\ref{line:encodep-ans-match}.\\
+\begin{minipage}{1.0\linewidth}
 \begin{algorithmic}[1]
-  \vspace{1ex}
   \setcounterref{ALG@line}{line:encodep-ans-match}
   \State $c \landplus compart(v_r) \geq compart(v')$
-  \vspace{1ex}
 \end{algorithmic}  
-Using the above constraints, we can divide the learned rules
-into compartments.
+\end{minipage}
+% Using the above constraints, we can divide the learned rules
+% into compartments.
 
@@
 
-\subsection{Fast and slow reactions}
-During the stay of a molecule in a compartment, it could be observed that some reactions were dominating than others.
-This can be explained by characterizing reactions as either slow or fast.
-A slow reaction can occur only when no other fast reaction is able to extend the molecule in that compartment.
-We can now define \textsc{Extend} recursively as follows:
-\begin{align*}
-  \textsc{Extends}(r, m) :=\;&   Apply(m,r) \lor
-  \Lor_{\mathclap{i \in [1,w]}} \textsc{Extends}( r, C(m,i ) )\\
-    \textsc{Extends}(r, \bot) :=\;&  \nu(v) \neq \bot
-  \end{align*}
-We now modify $\textsc{EncodeP}$ after~\ref{line:mcut}:
+\paragraph{Fast and slow reactions:}
+There is rate associated with chemical reactions. We abstract this by defining slow and fast rules.
+The fast rules dominate the slow rule.
+A slow rule can occur only when no other fast rule is able to extend the molecule in that compartment.
+Let us define function $\textsc{EncodeP}^-$, which generates constraints as $\textsc{EncodeP}$ expect
+line~\ref{line:encodep-vr-expand} is missing, i.e., we do not analyze expand part.
+% During the stay of a molecule in a compartment, it could be observed that some reactions were dominating than others. This can be explained by characterizing reactions as either slow or fast.
+% We can now define \textsc{Extend} recursively as follows:
+% $  \textsc{Extends}(r, m) :=\;   Apply(m,r) \lor \Lor_{{i \in [1,w]}} \textsc{Extends}( r, C(m,i ) )$\\
+% $\textsc{Extends}(r, \bot) :=\;  \nu(v) \neq \bot$\\
+We now modify $\textsc{EncodeP}$ after~\ref{line:encodep-vr-expand} to support fast reactions:\\
+\begin{minipage}{1.0\linewidth}
 \begin{algorithmic}[1]
-  \vspace{1ex}
-  \setcounterref{ALG@line}{line:mcut}
+  % \vspace{1ex}
+  \setcounterref{ALG@line}{line:encodep-vr-expand}
   \State $c \landplus \lnot \textsc{Fast}(v_r) \implies 
-  \Land_{t \in T} (\textsc{Fast}(t) \implies \lnot\textsc{Extends}(t,v))$
-  \vspace{1ex}
+  \Land_{t \in T} (\textsc{Fast}(t) \implies \lnot\lor_{ {\ell \in [1,d)}} \textsc{EncodeP}^-( v, t, \ell))$
+  % \vspace{1ex}
 \end{algorithmic}
+  % \vspace{-1mm}
+\end{minipage}
 Here, \textsc{Fast} sets the constraint on the rule to be fast.
 A negative molecule which is a proper subtree of an input molecule will cease to be negative if there is any fast reaction that is able to extend it as fast reactions happen aggressively and can make partial molecules complete.
-Therefore, we modify the function $\textsc{SugarSynth}$ after~\ref{line:encode-neg-mol}:
-\begin{algorithmic}[1]
-  \vspace{1ex}
-  \setcounterref{ALG@line}{line:encode-neg-mol}
-  \State $nCons \landplus
-  \lnot  \Lor_{r \in R} \textsc{Fast}(r) \land \textsc{Extends}(r,m')$
-  \vspace{1ex}
-\end{algorithmic}
-Here, \textsc{Fast} constraints the rule to be fast. The constraints state that none of the rules should be both fast and able to extend the negative molecule received from the model.
+Due to limited space, we will not present the exact constraints.
 
-\subsection{Unbounded molecules}
-We always observe a finite set of molecules. However,
-a set of production rules is capable of producing
-an unboundedly large size of molecules.
-In such cases, rules are allowed to form molecules having repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules.
-Constraints are added to encode that those rules are all applied in the same compartment and are slow reactions.
-We put a limit on the depth of such repeating subtrees ($d_0$) and the number of repetitions ($r_0$) .
-Then, for a particular depth $d$ and repetitions $r$, we recursively define \textsc{Repeat} and \textsc{ExactMatch} as follows:
-\begin{align*}
-  \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
-    \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
-    \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
-    \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
-    \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
-    \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
-    \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
-    \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
-    \textsc{ExactMatch}(m,\bot) :=\;& \bot 
-  \end{align*}
-  where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
-The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.
-\begin{algorithmic}[1]
-  \vspace{1ex}
-  \setcounterref{ALG@line}{line:consNewR}
-  \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
-  \vspace{1ex}
-\end{algorithmic}
-where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
+% Therefore, we modify the function $\textsc{SugarSynth}$ after~\ref{line:encode-neg-mol}:
+% \begin{algorithmic}[1]
+%   \setcounterref{ALG@line}{line:encode-neg-mol}
+%   \State $nCons \landplus
+%   \lnot  \Lor_{r \in R} \textsc{Fast}(r) \land \textsc{Extends}(r,m')$
+% \end{algorithmic}
+% Here, \textsc{Fast} constraints the rule to be fast. The constraints state that none of the rules should be both fast and able to extend the negative molecule received from the model.
 
-\subsection{Non-monotonic rules}
-Some molecules produced dictate that after a certain rule is applied to them in a compartment, it restricts the application of any other rule from that point of time.
-Such a rule is said to have \textsc{HardEnds}.
+{\em Unbounded molecules: }
+We only observe a finite set of molecules in cells.
+However, a set of production rules may be capable of producing an unboundedly large number of molecules.
+In such cases, rules produce molecules the have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules.
+The rules may be acceptable in some biological settings.
+We modify constraints to not declare such molecules as negative.
+Let us suppose we have a repeat pattern of depth $d$ with $r$ repetitions.
+Let $v$ be the node in template molecule where repetition has begin.
+We define $\textsc{RepeatHeads}(v,r,d)$ that returns nodes $v_0,.... v_r$ such that
+$v_1 = v$, $v_i$ is the $d$th ancestor of $v_{i+1}$.
+% Those rules should in the same compartment and are slow.
+Let us define \textsc{Repeat} and \textsc{Exact}, which encodes that the trees rooted at $v_{i}$s repeat.\\
+$\textsc{Repeat}([v_0,...,v_r], v') :=  \textsc{Exact}(v',v_r,\bot) \land \Land_{i\in[0,r-1]} \textsc{Exact}( v_i, v_{i+1},v_{i+1})$\\
+% $\textsc{Exact}(v,v') := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i))$ \\
+% $\textsc{Exact}(\bot,v') := \bot, \textsc{Exact}(v,\bot) := \bot, \textsc{Exact}(\bot,\bot) := \bot$\\
+$\textsc{Exact}(\bot,v',\_) := \lfalse, \textsc{Exact}(v,\bot,\_) := \lfalse,$\\
+$ \textsc{Exact}(\bot,\bot,\_) := \ltrue,\textsc{Exact}(v_s,\_,v_s) := \ltrue$\\
+$\textsc{Exact}(v,v',v_s) := (M(v) = M(v')) \land \Land_{i \in [1,w]} \textsc{Exact}(C(v,i),C(v',i),v_s)$\\
+% $\textsc{Repeat}(v, v', r, d) :=  \textsc{Repeat}(v, ancestor(v'), r, d-1)$\\
+% $\textsc{Repeat}(v, v', r, 0) := (compart(v) = compart(v')) \land \textsc{Repeat}(v, C^{2d}(v'), r-1, d)$ \\
+% $\textsc{Repeat}(v, v', 0, d) :=  \textsc{ExactMatch}(v,C^{2d}(v'))$ \\
+% $\textsc{Repeat}(\bot, \bot, r, d) :=  \top  \quad \textsc{Repeat}(v, \bot, r, d) :=  \bot
+% \quad \textsc{Repeat}(\bot, v', r, d) :=  \bot $\\
+% \begin{align*}
+%   \textsc{Repeat}(m, m', r, d) :=\;&  \textsc{Repeat}(m, ancestor(m'), r, d-1)\\
+%   \textsc{Repeat}(m, m', r, 0) :=\;& (compart(m) = compart(m')) \land \textsc{Repeat}(m, C^{2d}(m'), r-1, d) \\
+%   \textsc{Repeat}(m, m', 0, d) :=\;&  \textsc{ExactMatch}(m,C^{2d}(m')) \\
+%   \textsc{Repeat}(\bot, \bot, r, d) :=\;&  \top \\
+%   \textsc{Repeat}(m, \bot, r, d) :=\;&  \bot \\
+%   \textsc{Repeat}(\bot, m', r, d) :=\;&  \bot \\
+%   \textsc{ExactMatch}(m,m') :=\;& (M(m) = M(m')) \land \Land_{i \in [1,w]} \textsc{ExactMatch}(C(m,i),C(m',i)) \\
+%   \textsc{ExactMatch}(\bot,m') :=\;& \bot \\
+%   \textsc{ExactMatch}(m,\bot) :=\;& \bot 
+% \end{align*}
+  % where $ancestor(m')$ is the parent node of $m'$ and $C^{2*d}$ is the application of $C(m',i)$, $2d$ times and $i$ comes from the reverse of the path traversed by m' to reach it's $d^{th}$ parent.
+%   The constraints for allowing runaway reactions are added after~\ref{line:consNewR} to modify \textsc{SugarSynth}.\\
+% \begin{algorithmic}[1]
+%   \vspace{1ex}
+%   \setcounterref{ALG@line}{line:consNewR}
+%   \State $mCons \landplus \Lor_{1 \leq r \leq r_0} \Lor_{1 \leq d \leq d_0} \Lor_{m \in \mu} (M(m) \neq M(\hat{m})) \implies  \textsc{Repeat}(m, \hat{m}, r, d) $
+%   \vspace{1ex}
+% \end{algorithmic}
+% where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
+We modify the third constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules not in $\mu$. We change the definition of $Neq$ as follows.\\
+$Neq(v, v') :=\; (\nu(v) \neq M(v') \land \lor_{r\in[1,r_0],d\in[1,d_0]}\textsc{Repeat}(\textsc{RepeatHeads}(v,r,d), v')) $\\
+\mbox{}\hspace{30mm}$\lor \Lor_{{i \in [1,w]}} Neq( C(v,i), C(v',i ) ),$\\
+where $d_0$ and $r_0$ are limits on the depth of the repeating subtrees and the number of repetitions respectively.
+The change will accept molecules with repeating patterns as positive samples.
+
+\paragraph{Non-monotonic rules :}
+Some production rules can not be applied if another node is present in a sibling.
+We call such rule non-monotonic because it may get disabled as the molecule grows.
 This feature of rules help in producing an exact set of desired molecules.
+We add an extra bit on each node of rule
+template called $\textsc{HardEnds}$.
+If the node is absent, its parent is present, and  $\textsc{HardEnds}$ bit is true, then
+no node must be present in the matching pattern at the time of the application of the rule.
 We modify the function $\textsc{MatchTree}$ by inserting the following constraints after~\ref{line:matchtree-cons}:
+\begin{minipage}{1.0\linewidth}
 \begin{algorithmic}[1]
@@
   \State $c \landplus \textsc{HardEnds}(v_r) \implies (mark \leq \tau(v))$
-  \vspace{1ex}
 \end{algorithmic}  
-In our method, \textsc{HardEnds} sets a constraint on the given rule to have hard ends.
+\end{minipage}
 The constraints state that if the applicable rule has \textsc{HardEnds}, then it has to be added at a time earlier than the current time of the molecule, effectively restricting the addition of further rules.

### 2020-10-15 src/sugar/sugar-encode.cpp

@@ -576,7 +576,7 @@ not_contain_mols( const sugar_mol_ptr& m1, const sugar_mol_ptr& m2,
       //std::cout<<"yes, parent exists"<<std::endl;
       not_m2_and_m1 = not_m2_and_m1 &&
         m2->get_parent()->get_unknown_sugar()->get_exists_cons();
-    }    
+    }
     strict.push_back( not_m2_and_m1 );
     VecExpr no_matches;
     no_fast_can_extend_unknown(m2,no_matches);
@@ -636,8 +636,6 @@ not_contain_mols( const sugar_mol_ptr& m1, const sugar_mol_ptr& m2 ) {
   VecExpr blocked;
   not_contain_mols( m1,  m2, dists, strict, blocked );
   //m1 contains m2 and all expansions of m2 are blocked
-  // VecExpr no_matches3;
-  // no_fast_can_extend(m2,no_matches3);
   //diagnostics_cons.push_back( mk_and(ctx ,blocked) );
   //dump( mk_or(ctx,strict) && mk_and(ctx ,blocked) );
   return mk_or( ctx, dists ) || (mk_or(ctx,strict) && mk_and(ctx ,blocked));
@@ -921,40 +919,7 @@ no_fast_match_cons( const sugar_mol_ptr& m, VecExpr& no_matches ) {
   no_fast_match_cons( m_parent, m_child_num, m_comp, no_matches );
 }
 
-//
-//     [m_parent] <-- point of operation
-//       /    
-//      /     
-//    m=[NULL]  Sibling
-//
-//  No fast rule is available to create m
-// z3::expr sugar_encoding::
-// no_fast_match_cons_at_null_leaf( const sugar_mol_ptr& m  ) {
-// void sugar_encoding::no_fast_can_extend_this(const sugar_mol_ptr& m, VecExpr& ans){  
-//   std::vector<sugar_mol*> parent_trail;
-//   std::vector< unsigned > sibling_trail;
-//   parent_trail.push_back( m.get() );
-//   // sibling_trail.push_back( 0 );
-//   for(auto r: rs){
-//     if( !r->check_is_empty() ) {
-//   for( unsigned d = 1; d < r->get_depth(); d++ ) {
-//     // find d deep match in past
-//     auto m_g = parent_trail.back();
-//     if( m_g->get_parent() ) {
-//       parent_trail.push_back( m_g->get_parent() );
[…]

### 2020-10-16 papers/sugar-synth/algo.tex
a few issues fixed

@@
 The domain of the first variables is $S \union \{\bot\}$.
-Let $SVars$ be the set of uncountable variables with the domain.
+Let $SVars$ be the set of unbounded variables with the domain.
 We will use the pool of $SVars$ to add variables to the templates.
@@
 The third situation is when a node position is in the matching part
-and has expanding decadents.
+and has expanding descendants.
 The nodes on the solid path from $v_0$ to the root $v_e$ of the expanding part are in the third situation.
@@
 indicate the set of four situations.
-Let $KVars$ be the set of uncountable variables with domain $K$. 
+Let $KVars$ be the set of unbounded variables with domain $K$. 
 % We will use the pool of $KVars$ to add variables to the templates.

### 2020-10-16 papers/sugar-synth/experiments.tex
variations first cut
a few issues fixed

@@
 %
-We have developed a specialized file format for giving input to~\ourtool.
-%
-The format allows one to declare monomers, list molecules over the monomers
-in the SMTLIB like format,
-and pass the input parameters of the method.
+% We have developed a specialized file format for giving input to~\ourtool.
+% %
+% The format allows one to declare monomers, list molecules over the monomers
+% in the SMTLIB like format,
+% and pass the input parameters of the method.
 
@@
 
-The results suggest that the tool is potentially applicable to larger data sets with
-the variants discussed earlier. However, we have not fully modeled all the information
-and biological intuition --- for example, the objective function for the rules.
+The results suggest that the tool is potentially applicable to larger data sets.
+% with
+% the variants discussed earlier.
+% However, we have not fully modeled all the information
+% and biological intuition --- for example, the objective function for the rules.
 

### 2020-10-16 papers/sugar-synth/experiments.tex
benchmark

@@
 {\em Benchmark: }
-We have applied our tool on four sets of molecules form
+We have applied our tool on three sets of real data (D1,D2,D3) and two sets of synthetic data (D4,D5). The molecules have been obtained from
 respiratory mucins of a cystic fibrosis patient (D1),
-human chorionic gonadotropin from a cancer cell line (D2), horse chorionic gonadotropin (D3), and
-hydra (D4)~\cite{Jaiman2018,hydra}.
+horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{}.
 The availability of clean data, where we are clear about the source and the collection
@@
 
-For D1, we synthesize the rules in 1.2 seconds. If we reduced any of the first
-two parameters, we failed.
-Giving an extra compartment in the first row did not impact the performance.
-We synthesize the rules for D2, which is also our motivating example, in 0.4 seconds.
-The performance of the tool is also robust against variations of the parameters.
-We synthesize the rules for D3 in 0.9 seconds if we use optimal values for the parameter.
+For D1, we synthesize the rules in 1.60 seconds. Even by reducing the first two parameters, we were able to synthesize the rules but it took longer time.
+Giving an extra compartment in the third row did not impact the performance.  We also observe the trade-off between the number of compartments
+and the depth of the rules at the second and third row of D1 and how it impacts performance.
+We synthesize the rules for D2, involving runaway reactions,  in 7.97 seconds.
+For D2, we also learned large rules, i.e., they are adding many nodes at a single time.
+We synthesize the rules for D3 in 0.57 seconds. 
 However, if we use too many resources or too little, the tool runs for a long time,
@@
 
-The synthesis for D4 takes a long time. We also observe the trade-off between the number of compartments
-and the depth of the rules at the first and second row of D4.
-For D3, we also learned large rules, i.e., they are adding many nodes at a single time.
+The synthesis for D4 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
+
 Currently, the tool does not have any objective function to optimize the synthesized rules.

### 2020-10-16 papers/sugar-synth/experiments.tex
conflict changes

--- 2020-10-16 papers/sugar-synth/experiments.tex
@@
 respiratory mucins of a cystic fibrosis patient (D1),
-human chorionic gonadotropin from a cancer cell line (D2), horse chorionic gonadotropin (D3), and
-hydra (D4)~\cite{Jaiman2018,hydra}.
+horse chorionic gonadotropin (D2),
+SARS-COV-2 (D3),
+human chorionic gonadotropin from a cancer cell line (D4),
+and
+synthetic data (D5)~\cite{Jaiman2018,SarsCov2}.
 The availability of clean data, where we are clear about the source and the collection
-methodology used, limits our choices for the data.
+methodology used, limits our choices.
 
@@
 parameter combinations to illustrate the relative performance of the tool.
-If we did not give large enough parameters, then the tool fails to synthesize the rules.
+If we did not budget large enough parameters, then the tool fails to
+synthesize the rules.
+We present the rules learned after giving minimum resources.
+Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
+in reasonable time.
 
-For D1, we synthesize the rules in 1.2 seconds. If we reduced any of the first
-two parameters, we failed.
-Giving an extra compartment in the first row did not impact the performance.
-We synthesize the rules for D2, which is also our motivating example, in 0.4 seconds.
-The performance of the tool is also robust against variations of the parameters.
-We synthesize the rules for D3 in 0.9 seconds if we use optimal values for the parameter.
-However, if we use too many resources or too little, the tool runs for a long time,
-which is typical. The search in combinatorial space is highly sensitive to the parameters.
-
-The synthesis for D4 takes a long time. We also observe the trade-off between the number of compartments
-and the depth of the rules at the first and second row of D4.
-For D3, we also learned large rules, i.e., they are adding many nodes at a single time.
 Currently, the tool does not have any objective function to optimize the synthesized rules.
@@
 for a reasonable objective function.
+The results suggest that the tool is potentially applicable to larger data sets.
+
+% For D1, we synthesize the rules in 1.2 seconds. If we reduced any of the first
+% two parameters, we failed.
+% Giving an extra compartment in the first row did not impact the performance.
+% We synthesize the rules for D2, which is also our motivating example, in 0.4 seconds.
+% The performance of the tool is also robust against variations of the parameters.
+% We synthesize the rules for D3 in 0.9 seconds if we use optimal values for the parameter.
+% However, if we use too many resources or too little, the tool runs for a long time,
+% which is typical. The search in combinatorial space is highly sensitive to the parameters.
+
+% The synthesis for D4 takes a long time. We also observe the trade-off between the number of compartments
+% and the depth of the rules at the first and second row of D4.
+% For D3, we also learned large rules, i.e., they are adding many nodes at a single time.
+% %
 
@@
 
-The results suggest that the tool is potentially applicable to larger data sets.
 % with

### 2020-10-16 src/sugar/sugar-encode.cpp

@@ -181,7 +181,10 @@ sugar_encoding( z3::context& ctx_, sugar_t_vec& sugars,
                 sugar_mol_vec& seed_ms_, sugar_mol_vec& ms_,
                 std::vector< std::vector<unsigned> > group_mols_,
                 unsigned num_rules, unsigned rule_depth_,
-                unsigned max_compartments_ )
+                unsigned max_compartments_,
+                // solving configurations
+                bool quantified_neg_cons_
+                )
   : ctx(ctx_)
   , sugars(sugars)
   , pos_ms(ms_)
@@ -190,6 +193,7 @@ sugar_encoding( z3::context& ctx_, sugar_t_vec& sugars,
   , rule_max_children(0)
   , max_compartments( max_compartments_ )
   , unknown_m(nullptr)
+  , is_quantified_neg_cons( quantified_neg_cons_ )
 {
   for( sugar_t_ptr s : sugars) {
     auto c =  s->get_children_num();

### 2020-10-16 src/sugar/sugar-encode.h

@@ -73,7 +73,10 @@ public:
                   sugar_mol_vec& seed_ms_, sugar_mol_vec& ms_,
                   std::vector< std::vector<unsigned> > group_mols_,
                   unsigned num_rules, unsigned rule_depth,
-                  unsigned max_compartments_ );
+                  unsigned max_compartments_,
+                  // solving configurations
+                  bool quantified_neg_cons
+                  );
 
   // constraints making calls
   z3::expr distinct_rules();
@@ -105,7 +108,7 @@ private:
   unsigned m_depth_max = 0;
 
   // solving confiuration
-  bool is_quantified_neg_cons = true;
+  bool is_quantified_neg_cons;
 
   // a number to immidate a large number
   // all other numbers may be asked to be smaller than this

### 2020-10-16 src/sugar/sugar.cpp

@@ -427,7 +427,10 @@ void synth_sugar( std::string input ) {
   //   std::cout << "\n";
   // }
   sugar_encoding s( ctx, sugars, seed_mols, mols, group_mols, num_rules,
-                    rule_depth, max_compartments);
+                    rule_depth, max_compartments,
+                    // solving configurations
+                    quantified_neg_cons
+                    );
 
   s.do_synth();
 

### 2020-10-16 papers/sugar-synth/experiments.tex

--- 2020-10-16 papers/sugar-synth/experiments.tex
@@
 respiratory mucins of a cystic fibrosis patient (D1),
-horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{}.
+horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{Jaiman2018,SarsCov2}.
 The availability of clean data, where we are clear about the source and the collection
-methodology used, limits our choices for the data.
+methodology used, limits our choices.
+
+% horse chorionic gonadotropin (D2),
+% SARS-COV-2 (D3),
+% human chorionic gonadotropin from a cancer cell line (D4),
+% and
+% synthetic data (D5)~\cite{}.
 
@@
 
-{\em Results:} We have applied \ourtool~on the data set. For each data set, we choose several
-parameter combinations to illustrate the relative performance of the tool.
-If we did not give large enough parameters, then the tool fails to synthesize the rules.
+{\em Results:} % We have applied \ourtool~on the data set. For each data set, we choose several
+% parameter combinations to illustrate the relative performance of the tool.
+% If we did not budget large enough parameters, then the tool fails to
+% synthesize the rules.
+% We present the rules learned after giving minimum resources.
+% Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
+% in reasonable time.
 
@@
 
-Currently, the tool does not have any objective function to optimize the synthesized rules.
-We are hoping to see more data sets such that we can develop a clearer picture
-for a reasonable objective function.
+% Currently, the tool does not have any objective function to optimize the synthesized rules.
+% We are hoping to see more data sets such that we can develop a clearer picture
+% for a reasonable objective function.
+% The results suggest that the tool is potentially applicable to larger data sets.
+
+% ==============================================================================
 

### 2020-10-16 papers/sugar-synth/experiments.tex
citation added

@@
 respiratory mucins of a cystic fibrosis patient (D1),
-horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{Jaiman2018,SarsCov2}.
+horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{Jaiman2018,10.1093/glycob/cwaa042}.
 The availability of clean data, where we are clear about the source and the collection
@@
 
-{\em Results:} % We have applied \ourtool~on the data set. For each data set, we choose several
-% parameter combinations to illustrate the relative performance of the tool.
-% If we did not budget large enough parameters, then the tool fails to
-% synthesize the rules.
-% We present the rules learned after giving minimum resources.
+{\em Results:} We have applied \ourtool~on the data set. For each data set, we choose several
+parameter combinations to illustrate the relative performance of the tool.
+If we did not budget large enough parameters, then the tool fails to
+synthesize the rules.
+We present the rules learned after giving minimum resources.
 % Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
@@
 The synthesis for D4 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
+The results suggest that the tool is potentially applicable to larger data sets.
 
@@
 % for a reasonable objective function.
-% The results suggest that the tool is potentially applicable to larger data sets.
 

### 2020-10-16 papers/sugar-synth/experiments.tex
some space saving

@@
-In this section, we  present our implementation of the proposed method for the synthesis
-and discuss the interesting optimizations in the implementation.
-%
+In this section, we  present our implementation of the proposed method.
 We also apply our tool on several data sets and illustrate the usability of our method.
@@
 {\em Implementation:} 
-We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool} for
-analyzing the performance of the technique in practice.
+We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool}.
 %

### 2020-10-16 papers/sugar-synth/experiments.tex
D6

--- 2020-10-16 papers/sugar-synth/experiments.tex
@@
 {\em Benchmark: }
-We have applied our tool on three sets of real data (D1,D2,D3) and two sets of synthetic data (D4,D5). The molecules have been obtained from
+We have applied our tool on three sets of real data (D1,D2,D3,D4) and two sets of synthetic data (D5,D6). The molecules have been obtained from
 respiratory mucins of a cystic fibrosis patient (D1),
@@
 
-{\em Results:} % We have applied \ourtool~on the data set. For each data set, we choose several
+{\em Results:}
+% We have applied \ourtool~on the data set. For each data set, we choose several
 % parameter combinations to illustrate the relative performance of the tool.
@@
 % in reasonable time.
-
 For D1, we synthesize the rules in 1.60 seconds. Even by reducing the first two parameters, we were able to synthesize the rules but it took longer time.
@@
 For D2, we also learned large rules, i.e., they are adding many nodes at a single time.
-We synthesize the rules for D3 in 0.57 seconds. 
+We synthesize the rules for D3 in 0.57 seconds. We synthesize the rules for D4, which is also our motivating example, in 0.85 seconds.
 However, if we use too many resources or too little, the tool runs for a long time,
@@
 
-The synthesis for D4 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
+The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
 

### 2020-10-16 papers/sugar-synth/experiments.tex
first cut pushed

--- 2020-10-16 papers/sugar-synth/experiments.tex
@@
 In this section, we  present our implementation of the proposed method.
+%
 We also apply our tool on several data sets and illustrate the usability of our method.
@@
 {\em Benchmark: }
-We have applied our tool on three sets of real data (D1,D2,D3) and two sets of synthetic data (D4,D5). The molecules have been obtained from
+We have applied our tool on three sets of real data (D1,D2,D3,D4) and two sets of synthetic data (D5,D6). The molecules have been obtained from
 respiratory mucins of a cystic fibrosis patient (D1),
-horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3)~\cite{Jaiman2018,10.1093/glycob/cwaa042}.
-The availability of clean data, where we are clear about the source and the collection
-methodology used, limits our choices.
+horse chorionic gonadotropin (D2), SARS-CoV-2 spike protein T323/S325 (D3), and human chorionic gonadotropin from a cancer cell line (D4)~\cite{Jaiman2018,10.1093/glycob/cwaa042}.
+The availability of clean data, where we are clear about the source
+% and the collection
+% methodology used
+, limits our choices.
 
@@
 % SARS-COV-2 (D3),
-% human chorionic gonadotropin from a cancer cell line (D4),
+% 
 % and
@@
 
-{\em Results:} We have applied \ourtool~on the data set. For each data set, we choose several
+{\em Results:}
+We have applied \ourtool~on the data set. For each data set, we choose several
 parameter combinations to illustrate the relative performance of the tool.
@@
 For D2, we also learned large rules, i.e., they are adding many nodes at a single time.
-We synthesize the rules for D3 in 0.57 seconds. 
+We synthesize the rules for D3 in 0.57 seconds. We synthesize the rules for D4, which is also our motivating example, in 0.85 seconds.
 However, if we use too many resources or too little, the tool runs for a long time,
 which is typical. The search in combinatorial space is highly sensitive to the parameters.
+The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
 
-The synthesis for D4 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
-The results suggest that the tool is potentially applicable to larger data sets.
+% The results suggest that the tool is potentially applicable to larger data sets.
 

### 2020-10-16 papers/sugar-synth/motivation.tex
variations first cut
first cut pushed

@@
 \label{fig:dataset-gly}
-\vspace{-6mm}
+% \vspace{-6mm}
 \end{figure}
@@
 \ourtool, the tool that implements the method, reported the synthesized rules from
-Figure~\ref{fig:ex-gly}(b) in 0.4 seconds.
+Figure~\ref{fig:ex-gly}(b) in 0.85 seconds.
 In our tool, we first construct a synthesis query using the templates to encode
@@
 We need to iterate further.
-After 19 iterations, our tool synthesizes a set of rules that
+After 8 iterations, our tool synthesizes a set of rules that
 satisfies the requirements.

### 2020-10-16 papers/sugar-synth/experiments.tex
second version

@@
 {\em Benchmark: }
-We have applied our tool on three sets of real data (D1,D2,D3,D4) and two sets of synthetic data (D5,D6). The molecules have been obtained from
+We have applied our tool on three sets of real data (D1,D2,D3,D4) and two sets of synthetic data (D5,D6,D7,D8). The molecules have been obtained from
 respiratory mucins of a cystic fibrosis patient (D1),
@@
 which is typical. The search in combinatorial space is highly sensitive to the parameters.
-The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
-
-% The results suggest that the tool is potentially applicable to larger data sets.
+The performance of our tool on synthetic data is also similar.
+The results suggest that the tool is potentially applicable to larger data sets.
+% The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
 

### 2021-03-29 papers/sugar-synth/main.tex
variations first cut
a few issues fixed

@@
 \input{macro.tex}
+\setlength{\textfloatsep}{0pt}%
 
-\title{%Automated inference
-  Synthesis of production rules for glycans}
-\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
+\title{Automated inference of production rules for glycans}
+% \author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
 %\institute{IITB and NCBS}
@@
 In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell.
-We have build a tool based on the method and applied it to infer rules based on published glycan data.
+We have built a tool based on the method and applied it to infer rules based on published glycan data.
 \end{abstract}
@@
 
-\section{Related work}
-\label{sec:related}
-\input{related.tex}
+% \section{Related work}
+% \label{sec:related}
+% \input{related.tex}
 

### 2021-03-29 papers/sugar-synth/variations.tex

@@
 
-{\em Compartments : } The production rules may live in different ordered compartments.
+{\em Compartments : } The production rules may live in different compartments.
 % The compartments are ordered.
 A molecule moves from one compartment to the next.
-The rules of the current compartment apply to the molecule.
+% The rules of the current compartment apply to the molecule.
 % We have presented the formal description of the compartments in
@@
 % where \textsc{Repeat} adds the constraints described above due to which runaway reactions are allowed.
-We modify the third constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules not in $\mu$. We change the definition of $Neq$ as follows.\\
+We modify constraints of $ \textsc{MolTemplateCorrectness}(\hat{m},\mu)$, which encodes that negative molecules are not in $\mu$. We replace the definition of $Neq$ as follows.\\
 $Neq(v, v') :=\; (\nu(v) \neq M(v') \land \lor_{r\in[1,r_0],d\in[1,d_0]}\textsc{Repeat}(\textsc{RepeatHeads}(v,r,d), v')) $\\
@@
 \paragraph{Non-monotonic rules :}
-Some production rules can not be applied if another node is present in a sibling.
-We call such rule non-monotonic because it may get disabled as the molecule grows.
-This feature of rules help in producing an exact set of desired molecules.
+Some production rules are not applicable if another node is present in a sibling.
+We call such a rule non-monotonic because it may get disabled as the molecule grows.
+This feature of rules helps in producing an exact set of desired molecules.
 We add an extra bit on each node of rule

### 2021-04-02 papers/sugar-synth/algo.tex
paper reviewed

@@
 line~\ref{line:ruleCorr} creates the above constraints
-and store them in $tCons$.
+and stores them in $tCons$.
 Next, we will also be searching for the molecules that may be produced by the learned rules.
@@
 In our method, the call to $\textsc{MolTemplateCorrectness}$ at
-line~\ref{line:molCorr} implements above constraints and store them in
+line~\ref{line:molCorr} implements above constraints and stores them in
 $mCons$.
@@
 We use shorthand $F \landplus G$ for $ F := F \land G$.
-As we will see that $\textsc{EnocdeProduce}$ introduces fresh variable
+As we will see that $\textsc{EncodeProduce}$ introduces fresh variable
 maps $\tau$ and $cuts$ in the encoding.
-Since we negate the returned formula by $\textsc{EnocdeProduce}$ 
+Since we negate the returned formula by $\textsc{EncodeProduce}$ 
 and then we check the satisfiability.

### 2021-04-02 papers/sugar-synth/experiments.tex
paper reviewed

--- 2020-10-16 papers/sugar-synth/experiments.tex
@@
 parameter combinations to illustrate the relative performance of the tool.
-If we did not budget large enough parameters, then the tool fails to
+If we did not budget large enough parameters such as size of unknown molecules, number of rules etc, then the tool fails to
 synthesize the rules.
@@
 We synthesize the rules for D3 in 0.57 seconds. We synthesize the rules for D4, which is also our motivating example, in 0.85 seconds.
-However, if we use too many resources or too little, the tool runs for a long time,
-which is typical. The search in combinatorial space is highly sensitive to the parameters.
+However, if we use too many resources or too little, the tool runs for a long time as the search in combinatorial space is highly sensitive to the parameters.
 The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.

### 2021-04-02 papers/sugar-synth/intro.tex
a few issues fixed
paper reviewed

@@
 method for the production rules.
-Our method takes the observed glycan molecules a cell and synthesizes
+Our method takes the observed glycan molecules in a cell as input and synthesizes
 the possible production rules that may explain the observation.
@@
 of searched molecules.
-We generate another set of constraints using the templates for the unknown molecule encoding that
-are there other molecules that can be generated using the synthesized rules.
+We generate another set of constraints using the templates for the unknown molecule.
 We again solve the constraints using an SMT solver.
@@
 
+\todo{Related work}
 We have implemented the method in our tool~\ourtool. The tool takes input data in its own format.

### 2021-04-02 papers/sugar-synth/model.tex
variations first cut
paper reviewed

@@
 \end{df}
-
-Let $m = (V,M,C,v_0)$ and $m' = (V',M',C',v_0')$ be molecules.
+\todo{Rewrite the following paragraph}
+\todo{definitions mixed with algorithms seems strange I think you should define all the formal stuff first and then present the algorithms}
+Let $\mathit{m = (V,M,C,v_0)}$ and $m' = (V',M',C',v_0')$ be molecules.
 With some abuse of notation, we write $v \in m$ to denote $v \in V$.
@@
 The expanded molecule is
-$$
+$
 m' = (V \uplus V', M \uplus M', C \uplus C' \uplus \{(v,i) \mapsto v_0'\}, v_0)
-$$ if
-$Match( r, v_{0r}, m', ancestor(m', v_0', d) )$.
+$ if
+$Match( r, v_{0r}, m', ancestor(m', v_0', d) )$ where $\uplus$ is the disjoint union.
 The match condition states that after attaching the new nodes $V'$
@@
 
+Our algorithm terminates in every case.
+At each iteration, atleast one solution i.e set of rules is discarded.
+Since the set of all possible rules is finite, we can guarantee termination.
+In the worst case, the entire search space is explored.
+For a problem having 5 node types, 3 as maximum rule depth, 4 rules to be synthesized, 3 compartments and fast-slow reactions, the search space can only be as large as $\approx 10^{21}$ rules.
+Our templating mechanism can be applied to all aspects of biology. The viability of the synthesized rules can only be verified by conducting wet experiments.
 \subsection{Synthesis problem}

### 2021-04-02 papers/sugar-synth/variations.tex
paper reviewed

@@
 \paragraph{Non-monotonic rules :}
-Some production rules are not applicable if another node is present in a sibling.
-We call such a rule non-monotonic because it may get disabled as the molecule grows.
-This feature of rules helps in producing an exact set of desired molecules.
+Some production rules can not be applied if another node is present in a sibling.
+We call such rule non-monotonic because it may get disabled as the molecule grows.
+This feature of rules help in producing an exact set of desired molecules.
 We add an extra bit on each node of rule

### 2021-04-13 papers/sugar-synth/main.tex
paper reviewed
more experiments

@@
 \title{Automated inference of production rules for glycans}
-% \author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
+\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
 %\institute{IITB and NCBS}
@@
 instead of looking for the theoretical conditions.
+\section{Experiments with variations of the synthesis problem}
+\label{sec:appendix}
+We conducted separate experiments for the variations of the synthesis problem - non-monotonic rules and unbounded molecules using synthetic data.
+%
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c|c|c|c|c|c|}\hline
+     & \#molecules& \#Rules & Rule depth & \#Compartments & success? & Time (in secs.) 
+          \\\hline
+    %%cf-mucin.sugar%%
+         &   & 6  & 2 & 1 & Yes &  2.85 \\\cline{3-7}
+    N1   & 6 & 6  & 2 & 5 & Yes & {\bf 1.07}  \\\cline{3-7}
+         &   & 5  & 3 & 1 & No & 0.81  \\\hline
 
+    %%horse-cg.sugar%%
+         &   & 6  & 2 & 1 & Yes & 1.02  \\\cline{3-7}
+    N2   & 6 & 6  & 2 & 5 & Yes & {\bf 0.76}  \\\cline{3-7}
+         &   & 6  & 4 & 1 & Yes & 0.82 \\\hline
+
+    %%sars-cov2.sugar%%     
+         &   & 9  & 2  & 1  & Yes & 3.18  \\\cline{3-7}
+    N3   & 9 & 8  & 2 & 1 & Yes & {\bf 2.39}  \\\cline{3-7}
+         &   & 8  & 3 & 2 & Yes  &  16.69 \\\hline
+
+    %%human-cg.sugar%%     
+         &   & 6  & 3  & 1  & Yes & 73.61  \\\cline{3-7}
+    R1   & 5 & 5  & 3 & 2 & Yes & {\bf 47.72}  \\\cline{3-7}
+         &   & 4  & 4 & 1 & No  &  16.99 \\\hline
+
+    % %%D4.sugar%     
+         &   & 6  & 3 & 3 & Yes &  5.25 \\\cline{3-7}
+    R2   & 8 & 4  & 3 & 2 & Yes & {\bf 3.44}  \\\cline{3-7}
+         &   & 4  & 3 & 5 & Yes  &  4.70 \\\hline
+
+    % %%D5.sugar%     
+         &   & 7  & 3 & 2 & Yes &  6.85 \\\cline{3-7}
+    R3   & 8 & 7  & 3 & 4 & Yes & {\bf 2.09}  \\\cline{3-7}
+         &   & 6  & 3 & 1 & No  & 0.77  \\\hline
+  \end{tabular}   
+  \end{table}
+ We applied our tool on the synthetic datasets N1, N2 and N3 which required some rules to have $\textsc{HardEnds}$.
+There is a clear trend of increasing time as the size of our dataset increases.
+Between N1 and N2 which have same number of molecules in the dataset, N1 has bulkier molecules and hence the time taken is higher.
+Reducing the number of compartments increases the time slightly however is successful in both N1 and N2.
+By Reducing the number of rules to be learnt, the tool fails in N1 as it required minimum 6 rules.
+Rules produced in these datasets have $\textsc{HardEnds}$ in the appropriate places to ensure that the exact set of molecules can be made.
+ %
+
+  To demonstrate the working of our tool in case of runaway reactions or datasets with unbounded molecules, we applied our tool on the synthetic datasets R1, R2 and R3.
+The molecules in these datasets have repeating patterns and we observe the synthesized rules to capture this pattern.
+The constraints added as part of unbounded molecues take care not to declare the molecules created by the repeated application of these rules to be negative.
+The time taken for synthesizing the rules is particularly high for R1 as the molecules in the dataset are relatively bulky.
+Increasing the number of compartments or Increasing the number of rules to learn increases the time in R2.
+The search in the combinatorial space is expensive, hence the time taken is more.
 % 

### 2021-04-19 papers/sugar-synth/algo.tex
asd

@@
 Here, we are considering only single compartment case.
-The generalizations are discussed in section~\ref{sec:variations}.
+The generalizations are discussed in appendix~\ref{sec:variations}.
 Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
@@
 The method assumes that the input set $\mu$ is finite.
-Even if a set of rules can produce
-an unbounded number of molecules.
-No biology will exhibit such a set in a cell.
+This is a reasonable assumption because
+even if a set of rules can produce
+an unbounded number of molecules,
+no biology will exhibit an infinite set in a cell.
 Our method
@@
 If the method fails to find production rules,
-the user can always call the method with larger parameters.
+the user may call the method with larger parameters.
 % However, we do not expect to find a large number of or sized rules
@@
 The domain of the first variables is $S \union \{\bot\}$.
-Let $SVars$ be the set of unbounded variables with the domain.
+Let $SVars$ be the unbounded set of variables with the domain.
 We will use the pool of $SVars$ to add variables to the templates.
@@
 indicate the set of four situations.
-Let $KVars$ be the set of unbounded variables with domain $K$. 
+Let $KVars$ be the unbounded set of variables with domain $K$. 
 % We will use the pool of $KVars$ to add variables to the templates.
@@
 \end{df}
+
+For a node $v$ in a template if we assign $\kappa(v) = Absent$,
+we call the node {\em absent}. Otherwise, we call the node {\em present}.
+We will also be searching for the molecules that may be produced by the learned rules.
+Therefore, we need to define the search space for the molecules.
+We use templates for defining the search space.
+We limit the template size using a parameter,
+namely the height of the template.
+\begin{df}
+For given integers $h$ and $w$,
+a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
+each internal node has $w$ children, where 
+$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
+$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
+and
+$v_{0m} \in V$ is the root.
+\end{df}
+
 In the algorithm at line~\ref{line:createRtemp}, we call \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
@@
 % to the variables of the template.
-For a node $v$ in a template if we assign $\kappa(v) = Absent$,
-we call the node {\em absent}. Otherwise, we call the node {\em present}.
 % Not all the assignments may result in a valid rule.
@@
 and stores them in $tCons$.
-Next, we will also be searching for the molecules that may be produced by the learned rules.
-Therefore, we need to define the search space for the molecules.
-We use templates for defining the search space.
-We limit the template size using a parameter,
-namely the height of the template.
-\begin{df}
-For given integers $h$ and $w$,
-a {molecule template} $\hat{m} = (V, \nu, C, v_{0m})$ is a labeled full tree with height $h$ and 
-each internal node has $w$ children, where 
-$V$ is a set of nodes, $\nu : V \maps SVars$ maps nodes to sugar choice variables,
-$C : V \times \naturals \pmaps V$ maps the indexed children of nodes,
-and
-$v_{0m} \in V$ is the root.
-\end{df}
+% Next, 
 At line~\ref{line:createMtemp}, we use the call to \textsc{MakeTemplateMol}($S$, $h$, $w$) to
-create a template of height $h$ and children width $w$.
+create a molecule template of height $h$ and children width $w$.
 %
@@
 % a parameter of the whole method.
-
 %
@@
 In our method, the call to $\textsc{MolTemplateCorrectness}$ at
-line~\ref{line:molCorr} implements above constraints and stores them in
+line~\ref{line:molCorr} generates the above constraints and stores them in
 $mCons$.
@@
 % due to the fresh variables.
-Therefore, we need to introduce universal quantifiers to the
+Therefore, we need to introduce universal quantifiers for the
 fresh variables.
@@
 % application patterns of the learned rules do not produce $m'$.
-Next, the loop goes to the next iteration.
+Afterwords, the loop goes to the next iteration.
 Since the domain of all the variables is finite, eventually the loop must
@@
 Now let us look at the encoding generated by $\textsc{EncodeProduce}$.
-The process of production of molecules adds pieces of trees of nodes
+The process of production of molecules adds pieces of trees
 one after another.
@@
 and traverses the nodes in $t$.
-The while loop at~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
+The while loop at line~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
 the $\ell$th ancestor from $v$,
@@
 
+Our algorithm terminates in every case.
+At each iteration, atleast one solution i.e set of rules is discarded.
+Since the set of all possible rules is finite, we can guarantee termination.
+In the worst case, the entire search space is explored.
+For a problem having 5 node types, 3 as maximum rule depth, 4 rules to be synthesized, 3 compartments and fast-slow reactions, the search space can only be as large as $\approx 10^{21}$ rules.
+Our templating mechanism can be applied to all aspects of biology. The viability of the synthesized rules can only be verified by conducting wet experiments.
+
 % On the other hand,

### 2021-04-19 papers/sugar-synth/intro.tex
asd

@@
 the applicability of a rule on the leaf.
-Identifying the exact mechanism of the production by extensive
+Identifying the exact set of the production rules by extensive
 biochemical experiments are costly and often needs an initial
@@
 and using prior knowledge of biochemistry.
+We estimate that there are approximately $10^{70}$
+possible rule sets if we consider all biological variations.
 This is an error-prone process since the production rules must generate
@@
 %as illustrated in Figure~\ref{fig:illus-method}.
-Our tool first constructs constraints encoding that a set of unknown rules
+We first construct constraints encoding that a set of unknown rules
 defined by templates can assemble the input set of molecules.
@@
 a condition, we have quantifiers in the constraints.
-Now onward we use a solver that handles quantifiers over finite range variable in
+We use a solver that handles quantifiers over finite range variable in
 the synthesis query.

### 2021-04-19 papers/sugar-synth/model.tex
asd

--- 2020-10-14 papers/sugar-synth/model.tex
@@
 \end{df}
-
-Let $m = (V,M,C,v_0)$ and $m' = (V',M',C',v_0')$ be molecules.
-With some abuse of notation, we write $v \in m$ to denote $v \in V$.
+\todo{Rewrite the following paragraph}
+\todo{definitions mixed with algorithms seems strange I think you should define all the formal stuff first and then present the algorithms}
+Let us define notations related to the tree structure.
+Let $\mathit{m = (V,M,C,v_0)}$ and $m' = (V',M',C',v_0')$ be molecules.
+With an abuse of notation, we write $v \in m$ to denote $v \in V$.
 For each $v \in m$, if $(v,n)$ is not in the domain of $C$, we write $C(v,n) = \bot$.
@@
 For for $d > 0$, let $ancestor(m,v,d) = ancestor(m,v', d-1)$ if $C(v',i) = v$ for some $i$.
+
+Since we will be matching the parts of the trees and applying rules to expand them,
+let us introduce notations for matching.
 Let recursively-defined predicate $Match(m,v,m',v')$ state that 
@@
   \label{fig:exrule}
-  \vspace{-8mm}
+  \vspace{-6mm}
 \end{wrapfigure}
@@
 \subsection{Synthesis problem}
-We observe a set of glycan molecules $\mu$ present in a cell.
+In nature, we observe a set of glycan molecules $\mu$ present in a cell.
 %
-However, we do not know the production rules to produce the molecules.
+However, we do not necessarily know the production rules to produce the molecules.
 %

### 2021-04-19 src/sugar/sugar-encode.cpp

@@ -182,6 +182,7 @@ sugar_encoding( z3::context& ctx_, sugar_t_vec& sugars,
                 std::vector< std::vector<unsigned> > group_mols_,
                 unsigned num_rules, unsigned rule_depth_,
                 unsigned max_compartments_,
+                bool is_fast_reactions_,
                 // solving configurations
                 bool quantified_neg_cons_
                 )
@@ -193,6 +194,7 @@ sugar_encoding( z3::context& ctx_, sugar_t_vec& sugars,
   , rule_max_children(0)
   , max_compartments( max_compartments_ )
   , unknown_m(nullptr)
+  , is_fast_reactions( is_fast_reactions_ )
   , is_quantified_neg_cons( quantified_neg_cons_ )
 {
   for( sugar_t_ptr s : sugars) {

### 2021-04-19 src/sugar/sugar-encode.h

@@ -74,6 +74,7 @@ public:
                   std::vector< std::vector<unsigned> > group_mols_,
                   unsigned num_rules, unsigned rule_depth,
                   unsigned max_compartments_,
+                  bool is_fast_reactions,
                   // solving configurations
                   bool quantified_neg_cons
                   );
@@ -106,6 +107,7 @@ private:
   sugar_mol_ptr unknown_m;
   VecExpr unknown_m_ls_exist;
   unsigned m_depth_max = 0;
+  bool is_fast_reactions;
 
   // solving confiuration
   bool is_quantified_neg_cons;

### 2021-04-19 src/sugar/sugar.cpp

@@ -339,7 +339,8 @@ void read_file( std::istream& in, sugar_t_map& sugar_map,
                 std::vector< std::vector<unsigned> > group_mols,
                 unsigned& num_rules, unsigned& rule_depth,
                 unsigned& max_compartments,
-                bool& quantified_neg_cons ) {
+                bool& quantified_neg_cons,
+                bool& fast_reactions ) {
   std::vector<unsigned> local_group;
   bool group_active = false;
   unsigned sugar_idx = 0;
@@ -380,6 +381,8 @@ void read_file( std::istream& in, sugar_t_map& sugar_map,
       max_compartments = read_unsigned( in );
     }else if( cmd == "quantified-neg-cons" ){
       quantified_neg_cons = (read_unsigned( in ) > 0);
+    }else if( cmd == "allow-fast-reactions" ){
+      fast_reactions = (read_unsigned( in ) > 0);
     }else{
       bio_synth_error( "sugar_parse", "unknown command " << cmd << " !");
     }
@@ -403,6 +406,7 @@ void synth_sugar( std::string input ) {
   unsigned rule_depth = 0;
   unsigned max_compartments = 1;
   bool quantified_neg_cons = true;
+  bool fast_reactions = true;
 
   std::ifstream in;
   in.open( input );
@@ -411,7 +415,7 @@ void synth_sugar( std::string input ) {
     //File does not exist code here
   }
   read_file( in, sugar_map, seed_mols, mols, group_mols, num_rules, rule_depth,
-             max_compartments, quantified_neg_cons );
+             max_compartments, quantified_neg_cons, fast_reactions );
   if( seed_mols.size() == 0 ) {
     bio_synth_error( "sugar_parse", "no seed mols declared!" );
   }
@@ -429,7 +433,8 @@ void synth_sugar( std::string input ) {
   sugar_encoding s( ctx, sugars, seed_mols, mols, group_mols, num_rules,
                     rule_depth, max_compartments,
                     // solving configurations
-                    quantified_neg_cons
+                    quantified_neg_cons,
+                    fast_reactions,
                     );
 
   s.do_synth();

### 2021-04-19 papers/sugar-synth/algo.tex
asdd

@@
 a set of production rules that produce a given set $\mu$ of molecules.
+Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
 Here, we are considering only single compartment case.
 The generalizations are discussed in appendix~\ref{sec:variations}.
-Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
 
@@
 \begin{wrapfigure}{r}{0.37\textwidth}
-    \vspace{-3mm}
+    \vspace{-10mm}
         \center
@@
 % are added at or after $mark$.
-The local variable $c$ collects a conjunction of constraints as they are
+The local variable $c$ collects the conjunction of constraints as they are
 generated. % and $c$ is initially $\ltrue$.
@@
 nodes inside the expanding part are added together.
-Therefore, there should be no cuts with in the set of added nodes.
+Therefore, there should be no cuts within the set of added nodes.
 Furthermore, if there are children nodes below the added nodes,
@@
 
-Our algorithm terminates in every case.
-At each iteration, atleast one solution i.e set of rules is discarded.
-Since the set of all possible rules is finite, we can guarantee termination.
-In the worst case, the entire search space is explored.
-For a problem having 5 node types, 3 as maximum rule depth, 4 rules to be synthesized, 3 compartments and fast-slow reactions, the search space can only be as large as $\approx 10^{21}$ rules.
-Our templating mechanism can be applied to all aspects of biology. The viability of the synthesized rules can only be verified by conducting wet experiments.
+Our algorithm always terminates. At each iteration, at least one solution, i.e, set of
+rules is discarded.
+In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
+Since the set of all possible rules is finite, we
+guarantee termination. In the worst case, the entire search space is explored. For a problem having 10 node types, 3 as maximum rule depth, 10 rules to be synthesized, 3 compartments
+and fast-slow reactions, the search space can only be as large as $\approx 10^{70}$ rules.
+Our templating mechanism can be applied to all aspects of biology.
+The viability of the synthesized rules can only be verified by conducting wet experiments.
 

### 2021-04-19 papers/sugar-synth/intro.tex
asdd

@@
 
-In this paper, we are presenting a {\em novel} automated synthesis
+In this paper, we are presenting the {\em first} automated synthesis
 method for the production rules.
@@
 and synthesizing missing components in programs~\cite{sygus,Solar-Lezama2005}.
-We also use templates to model the space of the production rules.
+We also use templates to model the production rules.
 % The templates usually have parameters that limit the search space.

### 2021-04-19 papers/sugar-synth/main.tex
asd
asdd

@@
 The rules by which the enzymes operate are not fully understood.
-In this paper, we present a novel SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell.
+In this paper, we present the first SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell.
 We have built a tool based on the method and applied it to infer rules based on published glycan data.
@@
 
-\section{Variations of the synthesis problem}
-\label{sec:variations}
-\input{variations.tex}
-
 \section{Experiments}
@@
 
-\section{Conclusion}
+\section{Conclusion and future work}
 % ##limitations in the model because of limited information,
@@
 instead of looking for the theoretical conditions.
+
+\section{Variations of the synthesis problem}
+\label{sec:variations}
+\input{variations.tex}
+
 \section{Experiments with variations of the synthesis problem}
-\label{sec:appendix}
+\label{sec:ex-variants}
 We conducted separate experiments for the variations of the synthesis problem - non-monotonic rules and unbounded molecules using synthetic data.

### 2021-04-19 papers/sugar-synth/model.tex
asdd
sdaasd

@@
 \end{df}
-\todo{Rewrite the following paragraph}
-\todo{definitions mixed with algorithms seems strange I think you should define all the formal stuff first and then present the algorithms}
+% \todo{Rewrite the following paragraph}
+% \todo{definitions mixed with algorithms seems strange I think you should define all the formal stuff first and then present the algorithms}
 Let us define notations related to the tree structure.
@@
 
-\subsection{Synthesis problem}
+% \subsection{Synthesis problem}
 In nature, we observe a set of glycan molecules $\mu$ present in a cell.

### 2021-04-19 papers/sugar-synth/variations.tex
changes to appendix

@@
 the biological problem.
+However, in real settings, we often encounter many more variants which require us to support additional constraints to model the set of molecules and possible rules in that particular problem.
+Hence, we developed the following variations of the method described in the main text to support more realistic problems.
 % However, the biological problem has more details.
-We have developed variations of the above method to support more realistic problems.
+% We have developed variations of the above method to support more realistic problems.
 

### 2021-04-21 papers/sugar-synth/bio.tex
variations first cut
typos

@@
 
-The process of glycosylation is stochastic, since enzymes can operate in different time orders \cite{Spahn2016}.
+The process of glycosylation is stochastic since enzymes can operate in different time orders \cite{Spahn2016}.
 It is as if factory workers could operate in many different orders while building the car, first adding doors and later windows.
@@
 In \cite{Jaiman440792}, the authors had reported a method to infer the production rules when a single glycan is produced.
-However, the biologically-interesting case is when the data set contains many glycan trees.
+However, the biologically interesting case is when the data set contains many glycan trees.
 This raises the following question: given a set of glycan trees produced by a cell,
@@
   \end{minipage}
-\vspace{-9mm}
+% \vspace{-9mm}
 \end{figure}

### 2021-04-21 papers/sugar-synth/intro.tex
sdaasd
typos

@@
 Familiar examples of this process from biology include the synthesis of linear DNA from nucleotide building blocks and the synthesis of linear proteins from amino-acid building blocks.
-In both these examples the synthesis is templated: the new DNA or protein molecule is essentially copied from an existing molecule.
+In both these examples, the synthesis is templated: the new DNA or protein molecule is essentially copied from an existing molecule.
 However, most biological assembly proceeds without a template.
@@
 
-A given cell produces a specific set of glycan molecules that are present in the cell. Since different cells produce different sets of the molecules,
+A given cell produces a specific set of glycan molecules that are present in the cell. Since different cells produce different sets of molecules,
 the assembly process must be programmable: the assembly process includes a set of production rules.
@@
 Our work is the {\em first} to consider the computational problem.
-Our method of the synthesis is similar to counterexample guided
+Our method of synthesis is similar to counterexample guided
 inductive synthesis(CEGIS)~\cite{cegis}.
@@
 constraints.
-We solve the constraints using off-the-shelf SMT solver.
+We solve the constraints using an off-the-shelf SMT solver.
 We call the query to the solver {\em synthesis query}.
@@
 stay in different compartments.
-The molecule under-assembly may flow through the compartments.
-The distribution of stay of the molecules in a compartment also affects
+The molecule under assembly may flow through the compartments.
+The distribution of the stay of the molecules in a compartment also affects
 the execution model.
 Furthermore, we may have variations in the type and quality of data available to us.
-We have added the support for the variations in our method.
+We have added support for the variations in our method.
 
-\todo{Related work}
+% \todo{Related work}
 We have implemented the method in our tool~\ourtool. The tool takes input data in its own format.
@@
 In section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
-In section~\ref{sec:model}, we present the formal model of the glycans,
+In section~\ref{sec:model}, we present the formal model of the glycans
 and their production rules.

### 2021-04-21 papers/sugar-synth/variations.tex
typos

@@
 \paragraph{Fast and slow reactions:}
-There is rate associated with chemical reactions. We abstract this by defining slow and fast rules.
+There is a rate associated with chemical reactions. We abstract this by defining slow and fast rules.
 The fast rules dominate the slow rule.
@@
 However, a set of production rules may be capable of producing an unboundedly large number of molecules.
-In such cases, rules produce molecules the have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules.
+In such cases, rules produce molecules that have repeating patterns of a subtree while rest of the tree being exactly same as one of the input molecules.
 The rules may be acceptable in some biological settings.
@@
 Some production rules can not be applied if another node is present in a sibling.
-We call such rule non-monotonic because it may get disabled as the molecule grows.
-This feature of rules help in producing an exact set of desired molecules.
+We call such rules non-monotonic because it may get disabled as the molecule grows.
+This feature of rules helps in producing an exact set of desired molecules.
 We add an extra bit on each node of rule

### 2021-04-21 papers/sugar-synth/algo.tex
sdaasd

@@
 % the nodes that are ancestors of expanding part among the matching nodes.
-Finally, the fourth situation is the rest of node positions
+Finally, the fourth situation is the rest of the node positions
 in the matching part, which is the light gray area.
@@
 % application patterns of the learned rules do not produce $m'$.
-Afterwords, the loop goes to the next iteration.
+Afterwards, the loop goes to the next iteration.
 Since the domain of all the variables is finite, eventually the loop must
@@
 % We will discuss the handling of this situation at the end of this subsection.
-$\textsc{EncodeProduce}$ uses help of three other supporting functions,
+$\textsc{EncodeProduce}$ uses the help of three other supporting functions,
 $\textsc{EncodeP}$, $\textsc{MatchTree}$, and $\textsc{MatchCut}$.
@@
 % Therefore, if $v$ is a cut node, some template must be applied. 
-Since we are matching with template rule, we do not know the
+Since we are matching with a template rule, we do not know the
 position of the root of expanding nodes.
@@
 nodes are added in the molecule in the correct order.
-The last two inputs of the function is timestamp $mark$ and a bit $isExpand$,
+The last two inputs of the function are timestamp $mark$ and a bit $isExpand$,
 which tells us that the matching nodes should be added before or after $mark$.
@@
 Our algorithm always terminates. At each iteration, at least one solution, i.e, set of
-rules is discarded.
+rules are discarded.
 In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
@@
 and fast-slow reactions, the search space can only be as large as $\approx 10^{70}$ rules.
-Our templating mechanism can be applied to all aspects of biology.
+Our templates are sufficiently expressive to cover all aspects of biology.
 The viability of the synthesized rules can only be verified by conducting wet experiments.

### 2021-04-21 papers/sugar-synth/experiments.tex
more experiments
asdd

@@
 In this section, we present our implementation of the proposed method.
-%
-We also apply our tool on several data sets and illustrate the usability of our method.
+% %
+% We also apply our tool on several data sets and illustrate the usability of our method.
 
@@
 {\em Benchmark: }
-We have applied our tool on three sets of real data (D1,D2,D3,D4) and two sets of synthetic data (D5,D6). The molecules have been obtained from
+We have applied our tool to three sets of real data (D1, D2, D3, D4) and two sets of synthetic data (D5, D6). The molecules have been obtained from
 respiratory mucins of a cystic fibrosis patient (D1),
@@
 parameter combinations to illustrate the relative performance of the tool.
-If we did not budget large enough parameters such as size of unknown molecules, number of rules etc, then the tool fails to
+If we did not budget large enough parameters such as the size of unknown molecules, number of rules etc, then the tool fails to
 synthesize the rules.
@@
 The synthesis for D5 takes 0.72 seconds. We can observe that by reducing the number of rules to learn, the tool fails to learn rules as it required minimum 7 rules.
+Our synthesized production is in line with the reported rules
+in the literature~\cite{Jaiman2018}.
+More experiments on the variations of the problem
+are included in the appendix~\ref{sec:ex-variants}. %\hyperref[sec:appendix]{}.
 

### 2021-04-21 papers/sugar-synth/main.tex
sdaasd
changes to appendix

@@
 
-\section{Implementation optimizations}
+\section{Extended related work}
 
-% \subsection{Optimizations}
-We have incorporated several optimizations to gain efficiency or reduce
-the number of iterations in the method. Let us discuss some of the remarkable
-optimizations.
-
-\paragraph{Ordered templates}
-Our method constructs a list of templates.
-Since the templates are symmetric, there is a potential of unnecessary
-iterations.
-Let us suppose the method rejects a set of learned rules.
-In the subsequent queries include the constraints that reject the set of rules,
-the solvers may choose the same set of rules again by permuting the mapping
-from the templates to the learned rule.
-To avoid these iterations, we define an arbitrary total order over the rules.
-We add ordering constraints stating that any assignments to the list of templates should produce
-ordered rules with respect to the total order.
-
-\paragraph{Quantifier instantiations}
-The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
-For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
-to ensure that the query is not unsatisfiable.
-The instantiations sufficiently slow down the solving process.
-We assist the solver by also adding an instantiation of the quantifiers that is equal
-to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
-We observe that for some inputs only adding the instantiations and not the universally quantified
-formula is more effective.
-
-\paragraph{Constraints for counterexample molecules}
-In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
-at line~\ref{line:consNewR} in each iteration.
-The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
-the construction of many terms repeatedly. In our implementation, we construct the constraints
-once outside the while loop.
-We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
-Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
-of the template variables that were returned by the solver at line~\ref{line:posModel}.
-Due to the assignments, the templates become concrete rules in the context of solving.
-
-\section{Features to support in future}
-
-\subsection{Full organism data vs single cell data}
-
-The experiments that observe the set of glycan molecules are of two kinds.
-In one kind, we isolate a single cell type of an organism and identify all the glycans present in the cells.
-In our presentation, we have assumed this kind of source of data.
-However, the experiments are difficult to conduct.
-In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
-So the information that which glycans are coming from which
-cells is unknown.
-In this situation, the synthesis has another task to map the molecules
-in $\mu$ to several cell types.
-A glycan may be present in multiple cell types.
-
-Our method is easily adapted to consider this kind of data.
-We search for a small covering set of subsets of
-$\powerset{\mu}$ such that each set in the cover allows a small
-set of production rules.
-The cover sets indicate the different types of cells in terms of
-the presence of glycans.
-This variation makes the problem particularly hard.
-There can be a potentially large number of covering subsets.
-So far, we have not encountered a large enough data set such that
-we apply the variation effectively.
-
-\subsection{Incomplete data}
-
-The experiments are imperfect. They may not detect all possible glycans
-in a cell.
-We may need to leave the possibility of allowing a few more molecules to be
-producible beyond $\mu$.
-We replace the no extra molecule constraint by a constraint that allow molecules
-that are `similar' to molecules in $\mu$.
-However, we could not imagine any measure of similarity that naturally stems from
-biological intuition.
-For now, our tool supports the count of the number of monomer differences
-as a measure of similarity.
-
-\section{Extended related work}
+\todo{Related work: Kappa calculus framework, citations on applications of formal methods to biological structures,
+some work done in theoretical chemistry might be relevant (e.g., the work of Christoph Flamm from University of Vienna), cite extended related in main text}
 
@@
 %
-The solvers attempt to find a solution of the satisfiability problem.
+The solvers attempt to find a solution to the satisfiability problem.
 %
@@
 % response timing to stimulus~\cite{response-timing-paper}.
-In~\cite{delayedCTMC}, a methods based on
+In~\cite{delayedCTMC}, a method based on
 model-checking is used to estimate the time evolution of the probability
@@
 GRNs with unknown interactions are modeled using
-a automata-like model with missing components
+an automata-like model with missing components
 and the specification is the variations of the behavior of the system under mutations.
@@
 There is a clear trend of increasing time as the size of our dataset increases.
-Between N1 and N2 which have same number of molecules in the dataset, N1 has bulkier molecules and hence the time taken is higher.
+Between N1 and N2 which have the same number of molecules in the dataset, N1 has bulkier molecules and hence the time taken is higher.
 Reducing the number of compartments increases the time slightly however is successful in both N1 and N2.
-By Reducing the number of rules to be learnt, the tool fails in N1 as it required minimum 6 rules.
+By Reducing the number of rules to be learnt, the tool fails in N1 as it required a minimum of 6 rules.
 Rules produced in these datasets have $\textsc{HardEnds}$ in the appropriate places to ensure that the exact set of molecules can be made.
@@
 The molecules in these datasets have repeating patterns and we observe the synthesized rules to capture this pattern.
-The constraints added as part of unbounded molecues take care not to declare the molecules created by the repeated application of these rules to be negative.
+The constraints added as part of unbounded molecules take care not to declare the molecules created by the repeated application of these rules to be negative.
 The time taken for synthesizing the rules is particularly high for R1 as the molecules in the dataset are relatively bulky.
@@
 The search in the combinatorial space is expensive, hence the time taken is more.
+
+\section{Implementation optimizations}
+
+% \subsection{Optimizations}
+We have incorporated several optimizations to gain efficiency or reduce
+the number of iterations in the method. Let us discuss some of the remarkable
+optimizations.
+
+\paragraph{Ordered templates}
+Our method constructs a list of templates.
+Since the templates are symmetric, there is a potential for unnecessary
+iterations.
+Let us suppose the method rejects a set of learned rules.
+In the subsequent queries include the constraints that reject the set of rules,
+the solvers may choose the same set of rules again by permuting the mapping
+from the templates to the learned rule.
+To avoid these iterations, we define an arbitrary total order over the rules.
+We add ordering constraints stating that any assignments to the list of templates should produce
+ordered rules with respect to the total order.
+
+\paragraph{Quantifier instantiations}
+The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
+For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
+to ensure that the query is not unsatisfiable.
+The instantiations sufficiently slow down the solving process.
+We assist the solver by also adding an instantiation of the quantifiers that is equal
+to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
+We observe that for some inputs only adding the instantiations and not the universally quantified
+formula is more effective.
+
+\paragraph{Constraints for counterexample molecules}
+In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
+at line~\ref{line:consNewR} in each iteration.
+The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
+the construction of many terms repeatedly. In our implementation, we construct the constraints
+once outside the while loop.
+We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
+Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
+of the template variables that were returned by the solver at line~\ref{line:posModel}.
+Due to the assignments, the templates become concrete rules in the context of solving.
+
+\section{Features to support in future}
+
+\subsection{Full organism data vs single cell data}
+
+The experiments that observe the set of glycan molecules are of two kinds.
+In one kind, we isolate a single cell type organism and identify all the glycans present in the cells.
+In our presentation, we have assumed this kind of source of data.
+However, the experiments are difficult to conduct.
+In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
+So the information that which glycans are coming from which
+cells is unknown.
+In this situation, the synthesis has another task to map the molecules
+in $\mu$ to several cell types.
+A glycan may be present in multiple cell types.
+
+Our method is easily adapted to consider this kind of data.
+We search for a small covering set of subsets of
+$\powerset{\mu}$ such that each set in the cover allows a small
+set of production rules.
+The cover sets indicate the different types of cells in terms of
+the presence of glycans.
+This variation makes the problem particularly hard.
+There can be a potentially large number of covering subsets.
+So far, we have not encountered a large enough data set such that
+we apply the variation effectively.
+
+\subsection{Incomplete data}
+
+The experiments are imperfect. They may not detect all possible glycans
+in a cell.
+We may need to leave the possibility of allowing a few more molecules to be
+producible beyond $\mu$.
+We replace the no extra molecule constraint by a constraint that allows molecules
+that are `similar' to molecules in $\mu$.
+However, we could not imagine any measure of similarity that naturally stems from
+biological intuition.
+For now, our tool supports the count of the number of monomer differences
+as a measure of similarity.
+
 % 

### 2021-04-21 papers/sugar-synth/model.tex

@@
   The solid tree with three nodes is the initial molecule.
-  The middle node $A$ and its right child $B$ forms a pattern, where the rule is applicable.
+  The middle node $A$ and its right child $B$ form a pattern, where the rule is applicable.
   %Upon applying the rule,
-  The rule adds a left child with label $A$ to the middle node.
+  The rule adds a left child with the label $A$ to the middle node.
   The rule is not applicable at the root $A$ due to pattern mismatch.

### 2021-04-21 papers/sugar-synth/motivation.tex
paper reviewed
asd

@@
 \centering
-\includegraphics[width=0.8\linewidth]{gfig2.png}
+\includegraphics[width=1\linewidth]{gfig2.png}
 \caption{A glycan data set. Figure credit: Anjali Jaiman, PhD thesis.}
@@
 glycan molecules present in a cell consists of three molecules, which are structurally similar to the three glycan molecules in Figure~\ref{fig:dataset-gly}.
-To keep illustration simple, we dropped the third glycan molecule.
+To keep illustration simple, we have dropped the
+third glycan molecule from the earlier set.
 The molecules contain four types of monomers.
 As we are considering the abstract case, we have named them $A$, $B$, $C$, $D$.
-Each monomer is associated with an arity, i.e., the maximum number of potential children.
+Each monomer is associated with an arity, i.e.,
+the maximum number of potential children.
 The arities of the monomers are $2$, $1$, $1$, and $0$, respectively.
@@
 that produce the molecules.
-All the rules are in the same compartment.
+All the rules are in the same compartment, i.e., they can be applied in arbitrary order.
 The rules have two kinds of nodes.
@@
 three as the limit on the rule heights.
-All rules are in single compartment.
+All rules are in a single compartment.
 
@@
 After the first query, we obtain the rules presented in Figure~\ref{fig:ex-gly}(e).
-The rule set can produce molecules that are not are not in input.
+The rule set can produce molecules that are not in input.
 % Next, we look for another molecule that may be produced by the rule.

### 2021-07-01 papers/sugar-synth/algo.tex
minor changes

@@
 a set of production rules that produce a given set $\mu$ of molecules.
-Our synthesis method \textsc{SugarSynth} is presented in algorithm~\ref{alg:sugar-synth}.
-Here, we are considering only single compartment case.
-The generalizations are discussed in appendix~\ref{sec:variations}.
+Our synthesis method \textsc{SugarSynth} is presented in Algorithm~\ref{alg:sugar-synth}.
+Here, we are considering only the single compartment case.
+The generalizations are discussed in Appendix~\ref{sec:variations}.
 
@@
 
-\begin{wrapfigure}{r}{0.37\textwidth}
-    \vspace{-10mm}
+\begin{wrapfigure}{r}{0.35\textwidth}
+
         \center
@@
 \end{wrapfigure}
+
 A node in a production rule can be in four situations.
@@
 The second situation is when a node position is not in the rule.
-The dashed area are the absent nodes.
-Among the matching nodes, we have two cases.
-% , namely if the expanding
-% nodes are the descendants of the matching node or not.
+The dashed area are the absent nodes. Among the matching nodes, we have two cases.
 The third situation is when a node position is in the matching part
@@
 
-In the algorithm at line~\ref{line:createRtemp}, we call \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
+In the Algorithm at line~\ref{line:createRtemp}, we call \textsc{MakeTemplatesRule}( $S$, $d$, $w$, $n$) to create $n$ templates for
 height $d$ and children width $w$.
@@
 
-\input{alg-encode-produce}
-
 \subsection{\textsc{EncodeProduce} in detail}
@@
 
+\input{alg-encode-produce.tex} 
+\vspace{-3mm}
 Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
@@
 This will cover the case when the $\textsc{EncodeProduce}$ is called at
-line~\ref{line:molenc} in algorithm~\ref{alg:sugar-synth}.
+line~\ref{line:molenc} in Algorithm~\ref{alg:sugar-synth}.
 However, at line \ref{line:consNewR} the molecule is a template
@@
 % All of them are presented in algorithm~\ref{alg:produce}.
-%
 
@@
 
-Our algorithm always terminates. At each iteration, at least one solution, i.e, set of
-rules are discarded.
+Our algorithm always terminates. At each iteration, at least one solution, i.e,
+a set of rules is discarded.
 In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
@@
 guarantee termination. In the worst case, the entire search space is explored. For a problem having 10 node types, 3 as maximum rule depth, 10 rules to be synthesized, 3 compartments
-and fast-slow reactions, the search space can only be as large as $\approx 10^{70}$ rules.
+and fast-slow reactions, the search space can only be as large as $\approx 10^{74}$ rules $(2^{10}*{(10+3-1) \choose (3-1)}*10^{(2^3 - 1)*10})$.
 Our templates are sufficiently expressive to cover all aspects of biology.

### 2021-07-01 papers/sugar-synth/experiments.tex
minor changes

@@
 The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
-to discharge the generated satisfiability queries.
+to discharge the generated satisfiability queries. All the experiments have been conducted on a laptop with 8GB (1x8GB) DDR4 at 2400MHz.
 %
@@
 synthesize the rules.
-We present the rules learned after giving minimum resources.
+We present the rules learned after giving minimum resources. However, the set of rules reported need not be either unique or minimal.
 % Our tool gave output that matched with the analysis of~\cite{Jaiman2018}
@@
 More experiments on the variations of the problem
-are included in the appendix~\ref{sec:ex-variants}. %\hyperref[sec:appendix]{}.
+are included in the Appendix~\ref{sec:ex-variants}. %\hyperref[sec:appendix]{}.
 

### 2021-07-01 papers/sugar-synth/intro.tex
minor change
minor changes

@@
 defined by templates can assemble the input set of molecules.
-The generated constraints involve Boolean variable, finite range variables, and integer ordering
+The generated constraints involve Boolean variables, finite range variables, and integer ordering
 constraints.
@@
 The method always terminates because the search space of rules is finite.
+The set of rules synthesized need not minimal.
+They also need not be unique.
+The solver reports the first set which satisfies the constraints.
 
@@
 We organize the paper as follows.
-In section~\ref{sec:bio}, we introduce the biological background.
+In Section~\ref{sec:bio}, we introduce the biological background.
 %of the synthesis problem and present the basis of the formal model.
 % If the reader wants to avoid getting into the biological details, she may skip the section.
-In section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
-In section~\ref{sec:model}, we present the formal model of the glycans
+In Section~\ref{sec:motivation}, we present a motivating example to illustrate our method.
+In Section~\ref{sec:model}, we present the formal model of the glycans
 and their production rules.
-In section~\ref{sec:algo}, we present our method for the synthesis problem.
-In section~\ref{sec:variations}, we present the variations of the problem.
+In Section~\ref{sec:algo}, we present our method for the synthesis problem.
 %and the modifications in our methods to support the variations.
-In section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
+In Section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
+In Section~\ref{sec:conclusion}, we conclude the paper.
+We present the extended related work in Appendix~\ref{sec:related}.
+
+% We have added several appendix sections.
+% In appendix~\ref{sec:variations}, we present the variations of the problem.
 % In section~\ref{sec:related} and section~\ref{sec:conclusion},
-% we present the related work, and conclude the paper.
-In section~\ref{sec:conclusion}, we conclude the paper.
 

### 2021-07-01 papers/sugar-synth/main.tex
minor changes

@@
 \title{Automated inference of production rules for glycans}
-\author{Ansuman Biswas, Ashutosh Gupta, Meghana Missula, and Mukund Thattai}
-%\institute{IITB and NCBS}
+\author{Ansuman Biswas$^2$, Ashutosh Gupta$^1$, Meghana Missula$^1$, and Mukund Thattai$^2$}
+\institute{$^1$IITB and $^2$NCBS}
 
@@
 \input{conclusion.tex}
-
+\nocite{Charles2013}
 \bibliographystyle{unsrt}
@@
 \section{Extended related work}
-
-\todo{Related work: Kappa calculus framework, citations on applications of formal methods to biological structures,
-some work done in theoretical chemistry might be relevant (e.g., the work of Christoph Flamm from University of Vienna), cite extended related in main text}
+\label{sec:related}
 
@@
 \paragraph{Quantifier instantiations}
-The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in algorithm~\ref{alg:sugar-synth}.
+The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in Algorithm~\ref{alg:sugar-synth}.
 For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers

### 2021-07-01 papers/sugar-synth/motivation.tex
minor changes

@@
 \caption{A glycan data set. Figure credit: Anjali Jaiman, PhD thesis.}
+
 \label{fig:dataset-gly}
+\vspace{7.5mm}
 % \vspace{-6mm}
@@
 gonadotropin~\cite{Harrd1992}.
-The data set has four glycan oligomers (shown in boxes and numbered). If we assume that all these oligomers
+The data set has four glycan oligomers (shown in boxes and numbered). We assume that all these oligomers
 are built by starting from a root GalNAc (yellow square) by adding one monomer at a
@@
 In the lower part of the figure, we illustrate if the enzymes are split into three compartments
-(separated by dotted lines), then certain reactions are prevented from occurring, so the set of structures
-made is reduced.
+(separated by dotted lines), then certain reactions are prevented from occurring. Thus, reducing the set of structures.
 In this case, we assume that only the terminal (bottom-most) structures will be produced as outputs.
@@
 The goal of this paper is to infer the rules.
+% \vspace{-2.5mm}
 
@@
 % the three molecules,
+
 Let us first consider six rules in Figure~\ref{fig:ex-gly}(b)

### 2021-07-04 papers/sugar-synth/algo.tex
dsadsa

@@
 Here, we are considering only the single compartment case.
-The generalizations are discussed in Appendix~\ref{sec:variations}.
+The generalizations are discussed in
+\choosefinal{extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}
 
@@
 % We will use the pool of $KVars$ to add variables to the templates.
+Our templates are sufficiently expressive to cover all aspects of biology.
 The templates are defined as follows.
@@
 
-\input{alg-encode-produce.tex} 
-\vspace{-3mm}
 Algorithm~\ref{alg:produce} presents function $\textsc{EncodeProduce}$
@@
 % (see if condition at line \ref{line:encodep-jth-child-cond}).
+
+\vspace{-3mm}
+\input{alg-encode-produce.tex}
 
@@
 
+\begin{theorem}
+  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h$, the it is in $\mu$ (soundness).
+  If it fails to synthesize rules, there are no rule set from the budget of depth $d$ and number $n$ (completeness).
+\end{theorem}
+
+The soundness is true due to the construction of constraints.
 Our algorithm always terminates. At each iteration, at least one solution, i.e,
@@
 Since the set of all possible rules is finite, we
-guarantee termination. In the worst case, the entire search space is explored. For a problem having 10 node types, 3 as maximum rule depth, 10 rules to be synthesized, 3 compartments
-and fast-slow reactions, the search space can only be as large as $\approx 10^{74}$ rules $(2^{10}*{(10+3-1) \choose (3-1)}*10^{(2^3 - 1)*10})$.
-Our templates are sufficiently expressive to cover all aspects of biology.
-The viability of the synthesized rules can only be verified by conducting wet experiments.
+guarantee termination.
+Therefore, the completeness holds.
+In the worst case, the entire search space is explored.
 

### 2021-07-04 papers/sugar-synth/appendix.tex
dsadsa

--- 2021-07-01 papers/sugar-synth/main.tex
@@
-\documentclass{llncs}
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-% \documentclass[conference,10pt]{IEEEtran}
-
-% \bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-% \citestyle{acmauthoryear}   %% For author/year citations
-
-\input{macro.tex}
-\setlength{\textfloatsep}{0pt}%
-
-\title{Automated inference of production rules for glycans}
-\author{Ansuman Biswas$^2$, Ashutosh Gupta$^1$, Meghana Missula$^1$, and Mukund Thattai$^2$}
-\institute{$^1$IITB and $^2$NCBS}
-
-\date{May 2019}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-  Glycans are tree-like polymers made up of sugar monomer building blocks.
-They are found on the surface of all living cells, and distinct glycan trees act as identity markers for distinct cell types.
-Proteins called GTase enzymes assemble glycans via the successive addition of monomer building blocks.
-The rules by which the enzymes operate are not fully understood.
-In this paper, we present the first SMT-solver-based iterative method that infers the assembly process of the glycans by analyzing the set of glycans from a cell.
-We have built a tool based on the method and applied it to infer rules based on published glycan data.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:intro}
-\input{intro}
-
-\section{Production of Glycans}
-\label{sec:bio}
-\input{bio.tex}
-
-% \clearpage
-
-\section{Motivating example}
-\label{sec:motivation}
-\input{motivation.tex}
-
-\section{Modelling of the synthesis problem}
-\label{sec:model}
-\input{model.tex}
-
-% \clearpage
-
-\section{Method for the synthesis problem}
-\label{sec:algo}
-\input{algo.tex}
-
-\section{Experiments}
-\label{sec:experiments}
-\input{experiments.tex}
-
-% \section{Related work}
-% \label{sec:related}
-% \input{related.tex}
-
-\section{Conclusion and future work}
-% ##limitations in the model because of limited information,
-% a)the boundary between two compartments is fuzzy, 
-% b) the growths of two branches are correlated/ not correlated..and some more points like these, for which wealth of data is not available, hence the model was unable to incorporate the specifics
-\label{sec:conclusion}
-\input{conclusion.tex}
-\nocite{Charles2013}
-\bibliographystyle{unsrt}
-\bibliography{biblio}
-
-\appendix
-
 \section{Extended related work}
@@
 
-% 
-% To resolve this paradox we borrow from the field of
-% al-gorithmic self-assembly, which explores how small buildingblocks
-% with stochastic local interactions assemble into various global
-% configurations [23]–[25].  A central inverse problemin self-assembly
-% is to design building blocks that assembleinto a target shape.  This
-% process has been studied using thetheoretical framework of Wang tiles:
-% square building blockswith colored sides that stick to one another
-% along sides withmatching colors [26].  For efficient assembly of Wang
-% tiles,their coloring must be cleverly chosen so the target shape isthe
-% unique terminal output of the stochastic assembly process[27].   That
-% is, their assembly must be algorithmic.  Glycans may be considered a
-% natural realization of the Wang construct,with monomers acting like
-% tiles whose stickiness is encodedby GTase enzymes.  If glycan
-% biosynthesis could be madealgorithmic, cells could suppress unwanted
-% byproducts andgenerate only desired glycan oligomers with high yield.
+%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------
 
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% End:

### 2021-07-04 papers/sugar-synth/intro.tex
dsadsa

@@
 and using prior knowledge of biochemistry.
-We estimate that there are approximately $10^{70}$
+We estimate that there may be more then $10^{70}$
 possible rule sets if we consider all biological variations.
+\footnote{For a problem having 10 monomers, 10 rules, 3 as rule size, 3 compartments
+and fast-slow reactions, the search space is $\approx 10^{74}$ rules $(2^{10}*{(10+3-1) \choose (3-1)}*10^{(2^3 - 1)*10})$.}
 This is an error-prone process since the production rules must generate
@@
 Furthermore, we may have variations in the type and quality of data available to us.
-We have added support for the variations in our method.
+\choosefinal{We support the variations, which is presented in 
+  the extended version of this paper is available at~\cite{arxiv-this}.}{We support the variations.}
 
 % \todo{Related work}
-We have implemented the method in our tool~\ourtool. The tool takes input data in its own format.
-We have applied the tool on data sets from published sources (these are available in the database UniCarbKB \cite{Campbell2013}).
+We have implemented the method in our tool~\ourtool. %The tool takes input data in its own format.
+We have applied the tool on data sets from published sources (available in the database UniCarbKB \cite{Campbell2013}).
 % We have obtained the rules for the data sets.
@@
 
-The key contributions of this paper are the following.
-\begin{itemize}
-\item We identified {\em a new area} of application for formal methods in biology.
-\item We have developed a {\em novel} synthesis method for discovering the production rules of glycan molecules from the output of the rules.
-\item We have implemented the method in a tool and applied it to data sets.
-\end{itemize}
+% The key contributions of this paper are the following.
+% \begin{itemize}
+% \item We identified {\em a new area} of application for formal methods in biology.
+% \item We have developed a {\em novel} synthesis method for discovering the production rules of glycan molecules from the output of the rules.
+% \item We have implemented the method in a tool and applied it to data sets.
+% \end{itemize}
 
@@
 In Section~\ref{sec:conclusion}, we conclude the paper.
-We present the extended related work in Appendix~\ref{sec:related}.
+\choosefinal{}{We present the extended related work in Appendix~\ref{sec:related}.}
 

### 2021-07-04 papers/sugar-synth/main.tex
dsadsa

@@
 
-\section{Extended related work}
-\label{sec:related}
-
-% Success of formal methods
-Formal methods have successful in applications to a vast range of
-problems from the analysis of systems.
-%
-In particular,
-the verification of hardware designs
-using SAT solver based model-checking has been applied
-to industrial-scale problems~\cite{biere1999symbolic2}.
-%
-The verification of software systems is a much harder problem, and
-many methods like bounded model-checking~\cite{biere2003bounded},
-abstract interpretation~\cite{lattice77}, and
-counterexample guided abstraction refinement (CEGAR)~\cite{ClarkeCEGAR} have
-been successful for the problem.
-%
-
-Synthesis of programs that implements a given behavior
-has been a focus of research for a while~\cite{PnueliSynthesis}.
-%
-In recent years, the synthesis of programs using SAT/SMT solvers
-has gained momentum.
-%
-The approach encodes the search of a program that exhibits a certain
-behavior into a satisfiability problem.
-%
-The solvers attempt to find a solution to the satisfiability problem.
-%
-The solution is the synthesized program.
-%
-In~\cite{SrivastavaSynthesis,Solar-Lezama2005},
-the method has been successfully applied
-to find programs that satisfy quantified specifications like sorting
-programs, and have missing ``holes'' and need proper implementations
-for them.
-%
-In~\cite{exampleSynth},
-a set of pairs of input and output examples is the specification of a program,
-and the synthesis method searches for programs in a space defined by a
-template~\cite{sygus}.
-
-% Boolean networks
-
-% Program synthesis
-Since the formal methods have been exhibiting effectiveness in the vast
-range of problems, the
-systems biology community has been applying the methods for various
-biological problems~\cite{fisher2007executable}.
-The formal methods approach is distinctly different from the statistical
-approach traditionally used in biology.
-In formal methods, we aim to match precisely the expected behavior rather
-than learn approximate artifacts from biological data.
-The use of formal methods belongs to two broad categories:
-the analysis of biological models and the synthesis of the models.
-
-The key focus has been to model gene regulatory networks (GRNs), since
-they are the core of the central dogma of biology.
-Boolean networks are often used to model
-GRNs to find stable points or attractors, i.e., nodes in the networks
-where a system eventually reaches no matter where it starts~\cite{wang2012BooleanOverview}.
-% In~\cite{boolean-networks-stabilty}, the networks are used to evaluate
-% stability of GRNs under mutations.
-The Boolean networks are not often sufficient to adequately model the behavior
-of GRNs.
-The continuous-time Markov chains (CTMCs) are used to bring in the aspects of
-timing and probabilistic constraints to GRNs.
-The transient behaviors of GRNs are also crucial for various aspects
-of biology.
-% For example, the decision making in some cells is based on
-% response timing to stimulus~\cite{response-timing-paper}.
-In~\cite{delayedCTMC}, a method based on
-model-checking is used to estimate the time evolution of the probability
-distributions over states of GRNs.
-
-There are many biological processes where we can observe the system behavior,
-but the exact mechanisms of the processes are not known.
-%
-Recently, the methods for formal synthesis are finding their applications in biology
-\cite{dunn2014defining,xuPluripotency,booleanModelKarp13,paoletti2014analyzing}.
-%
-In~\cite{koksal2013synthesis},
-GRNs with unknown interactions are modeled using
-an automata-like model with missing components
-and the specification is the variations of the behavior of the system under mutations.
-Their approach uses a counterexample-guided inductive synthesis (CEGIS) based algorithm~\cite{cegis}.
-CEGIS is a framework for synthesis. It first finds a program that may satisfy `samples' from
-the specification. If the synthesized program satisfies the specifications, the CEGIS terminates.
-Otherwise, the method learns a new sample where the program violates the specifications.
-It adds the sample to the set and goes to the next iteration.
-Typically, the constraint solvers find the programs during intermediate steps of CEGIS by
-solving the generated queries.
-%
-Our method follows a similar pattern, where
-a set of sampled observations is a specification, i.e., output molecules,
-and we need to find the governing programs, i.e., production rules.
-%
-In a subsequent work~\cite{fisher2015synthesising}, a Boolean network model is used,
-the functions attached to the nodes are considered unknown, and
-a different kind of data sets provides the behavior specification.
-In this approach, their earlier method is adopted to work in the new situation.
-%
-
-As far as we know, there has been no similar computation based analysis of glycan production rules.
-%
-However, there has been a theoretical analysis of the properties of the production
-rules in~\cite{Jaiman2018}. The work identifies the conditions for the production
-of a finite set of molecules and the cases of unique production methods.
-%
-In our work, we are taking the computational approach of synthesis from formal methods
-instead of looking for the theoretical conditions.
-
-\section{Variations of the synthesis problem}
-\label{sec:variations}
-\input{variations.tex}
-
-\section{Experiments with variations of the synthesis problem}
-\label{sec:ex-variants}
-We conducted separate experiments for the variations of the synthesis problem - non-monotonic rules and unbounded molecules using synthetic data.
-%
-\begin{table}
-  \centering
-  \begin{tabular}{|c|c|c|c|c|c|c|}\hline
-     & \#molecules& \#Rules & Rule depth & \#Compartments & success? & Time (in secs.) 
-          \\\hline
-    %%cf-mucin.sugar%%
-         &   & 6  & 2 & 1 & Yes &  2.85 \\\cline{3-7}
-    N1   & 6 & 6  & 2 & 5 & Yes & {\bf 1.07}  \\\cline{3-7}
-         &   & 5  & 3 & 1 & No & 0.81  \\\hline
-
-    %%horse-cg.sugar%%
-         &   & 6  & 2 & 1 & Yes & 1.02  \\\cline{3-7}
-    N2   & 6 & 6  & 2 & 5 & Yes & {\bf 0.76}  \\\cline{3-7}
-         &   & 6  & 4 & 1 & Yes & 0.82 \\\hline
-
-    %%sars-cov2.sugar%%     
-         &   & 9  & 2  & 1  & Yes & 3.18  \\\cline{3-7}
-    N3   & 9 & 8  & 2 & 1 & Yes & {\bf 2.39}  \\\cline{3-7}
-         &   & 8  & 3 & 2 & Yes  &  16.69 \\\hline
-
-    %%human-cg.sugar%%     
-         &   & 6  & 3  & 1  & Yes & 73.61  \\\cline{3-7}
-    R1   & 5 & 5  & 3 & 2 & Yes & {\bf 47.72}  \\\cline{3-7}
-         &   & 4  & 4 & 1 & No  &  16.99 \\\hline
-
-    % %%D4.sugar%     
-         &   & 6  & 3 & 3 & Yes &  5.25 \\\cline{3-7}
-    R2   & 8 & 4  & 3 & 2 & Yes & {\bf 3.44}  \\\cline{3-7}
-         &   & 4  & 3 & 5 & Yes  &  4.70 \\\hline
-
-    % %%D5.sugar%     
-         &   & 7  & 3 & 2 & Yes &  6.85 \\\cline{3-7}
-    R3   & 8 & 7  & 3 & 4 & Yes & {\bf 2.09}  \\\cline{3-7}
-         &   & 6  & 3 & 1 & No  & 0.77  \\\hline
-  \end{tabular}   
-  \end{table}
- We applied our tool on the synthetic datasets N1, N2 and N3 which required some rules to have $\textsc{HardEnds}$.
-There is a clear trend of increasing time as the size of our dataset increases.
-Between N1 and N2 which have the same number of molecules in the dataset, N1 has bulkier molecules and hence the time taken is higher.
-Reducing the number of compartments increases the time slightly however is successful in both N1 and N2.
-By Reducing the number of rules to be learnt, the tool fails in N1 as it required a minimum of 6 rules.
-Rules produced in these datasets have $\textsc{HardEnds}$ in the appropriate places to ensure that the exact set of molecules can be made.
- %
-
-  To demonstrate the working of our tool in case of runaway reactions or datasets with unbounded molecules, we applied our tool on the synthetic datasets R1, R2 and R3.
-The molecules in these datasets have repeating patterns and we observe the synthesized rules to capture this pattern.
-The constraints added as part of unbounded molecules take care not to declare the molecules created by the repeated application of these rules to be negative.
-The time taken for synthesizing the rules is particularly high for R1 as the molecules in the dataset are relatively bulky.
-Increasing the number of compartments or Increasing the number of rules to learn increases the time in R2.
-The search in the combinatorial space is expensive, hence the time taken is more.
-
-\section{Implementation optimizations}
-
-% \subsection{Optimizations}
-We have incorporated several optimizations to gain efficiency or reduce
-the number of iterations in the method. Let us discuss some of the remarkable
-optimizations.
-
-\paragraph{Ordered templates}
-Our method constructs a list of templates.
-Since the templates are symmetric, there is a potential for unnecessary
-iterations.
-Let us suppose the method rejects a set of learned rules.
-In the subsequent queries include the constraints that reject the set of rules,
-the solvers may choose the same set of rules again by permuting the mapping
-from the templates to the learned rule.
-To avoid these iterations, we define an arbitrary total order over the rules.
-We add ordering constraints stating that any assignments to the list of templates should produce
-ordered rules with respect to the total order.
-
-\paragraph{Quantifier instantiations}
-The constraints for rejecting the counterexample molecule is universally quantified at line~\ref{line:negCons} in Algorithm~\ref{alg:sugar-synth}.
-For any assignment, the solver needs to try sufficiently many instantiations of the quantifiers
-to ensure that the query is not unsatisfiable.
-The instantiations sufficiently slow down the solving process.
-We assist the solver by also adding an instantiation of the quantifiers that is equal
-to the values of the corresponding variables in the assignment $a$ at line~\ref{line:negModel}.
-We observe that for some inputs only adding the instantiations and not the universally quantified
-formula is more effective.
-
-\paragraph{Constraints for counterexample molecules}
-In the presentation of \textsc{SugarSynth}, we construct constraints of counterexample molecules
-at line~\ref{line:consNewR} in each iteration.
-The repeated work is impractical because the formula management system of Z3 will be overwhelmed by
-the construction of many terms repeatedly. In our implementation, we construct the constraints
-once outside the while loop.
-We pass the templates as the second parameter instead of concrete rules to~\textsc{EncodeProduce}.
-Later at the solving time in line~\ref{line:negModel}, constraints are added to assign values
-of the template variables that were returned by the solver at line~\ref{line:posModel}.
-Due to the assignments, the templates become concrete rules in the context of solving.
-
-\section{Features to support in future}
-
-\subsection{Full organism data vs single cell data}
-
-The experiments that observe the set of glycan molecules are of two kinds.
-In one kind, we isolate a single cell type organism and identify all the glycans present in the cells.
-In our presentation, we have assumed this kind of source of data.
-However, the experiments are difficult to conduct.
-In another and more convenient way, we smash the whole organism and identify all the glycans present in all the cells.
-So the information that which glycans are coming from which
-cells is unknown.
-In this situation, the synthesis has another task to map the molecules
-in $\mu$ to several cell types.
-A glycan may be present in multiple cell types.
-
-Our method is easily adapted to consider this kind of data.
-We search for a small covering set of subsets of
-$\powerset{\mu}$ such that each set in the cover allows a small
-set of production rules.
-The cover sets indicate the different types of cells in terms of
-the presence of glycans.
-This variation makes the problem particularly hard.
-There can be a potentially large number of covering subsets.
-So far, we have not encountered a large enough data set such that
-we apply the variation effectively.
-
-\subsection{Incomplete data}
-
-The experiments are imperfect. They may not detect all possible glycans
-in a cell.
-We may need to leave the possibility of allowing a few more molecules to be
-producible beyond $\mu$.
-We replace the no extra molecule constraint by a constraint that allows molecules
-that are `similar' to molecules in $\mu$.
-However, we could not imagine any measure of similarity that naturally stems from
-biological intuition.
-For now, our tool supports the count of the number of monomer differences
-as a measure of similarity.
-
 % 
@@
 \end{document}
+
+%--------------------- DO NOT ERASE BELOW THIS LINE --------------------------
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2021-07-05 papers/sugar-synth/algo.tex
dsadsa

@@
 The generalizations are discussed in
-\choosefinal{extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}
+\choosefinal{the extended version~\cite{arxiv-this}.}{ Appendix~\ref{sec:variations}.}
 
@@
 \begin{theorem}
-  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h$, the it is in $\mu$ (soundness).
+  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h+1$, the it is in $\mu$ (soundness).
   If it fails to synthesize rules, there are no rule set from the budget of depth $d$ and number $n$ (completeness).

### 2021-07-05 papers/sugar-synth/intro.tex
dsadsa

@@
 the possible production rules that may explain the observation.
-Our work is the {\em first} to consider the computational problem.
+To our knowledge,
+our work is the {\em first} to consider the computational problem.
 Our method of synthesis is similar to counterexample guided
@@
 The method always terminates because the search space of rules is finite.
-The set of rules synthesized need not minimal.
-They also need not be unique.
-The solver reports the first set which satisfies the constraints.
+The set of rules synthesized need not be minimal or unique. The solver reports
+the first set which satisfies the constraints.
+However, our method is adoptable. We can add various optimization criteria to
+find optimal rules for the given objectives, e.g., smallest rule sizes,
+number of rules, etc.
 
@@
 the execution model.
-Furthermore, we may have variations in the type and quality of data available to us.
+Furthermore, we may have variations in the type and quality of data available to us. For example, we may have missed a produced molecule in experiments.
 \choosefinal{We support the variations, which is presented in 
-  the extended version of this paper is available at~\cite{arxiv-this}.}{We support the variations.}
+  the extended version of this paper~\cite{arxiv-this}.}{We support the variations.}
 
@@
 %and the modifications in our methods to support the variations.
-In Section~\ref{sec:experiments}, we present our implementation  and experiments. %on some glycan data.
-In Section~\ref{sec:conclusion}, we conclude the paper.
+In Section~\ref{sec:experiments}-~\ref{sec:conclusion}, we present our experiments and conclude the paper. %on some glycan data.
+% In Section~\ref{sec:conclusion}, we conclude the paper.
 \choosefinal{}{We present the extended related work in Appendix~\ref{sec:related}.}

### 2021-07-05 papers/sugar-synth/bio.tex
dsadsa
asd

@@
 Glycosylation is carried out, not by a single enzyme, but by a large collection of so-called GTase enzymes that assemble one sugar monomer at a time into a final glycan tree.
+This process involves an ordered series of reactions, in which an enzyme first recruits the correct monomer, the enzyme-monomer complex binds to the target glycan at the appropriate motif, and finally a chemical reaction occurs which serves to bind the new monomer at the correct place on the glycan.
+The enzyme's binding motif can corresponding to a single monomer, or a large sub-structure of the entire glycan several nodes deep \cite{biswas2020promiscuity}.
 This process is reminiscent of a factory assembly line to make a car \cite{Jaiman2018}.
@@
 
-The process of glycosylation is stochastic since enzymes can operate in different time orders \cite{Spahn2016}.
+The process of glycosylation is stochastic, governed by the Poisson statistics of single-step chemical reactions.
+One result of this stochasticity is that the enzymes can operate in different time orders \cite{Spahn2016}.
 It is as if factory workers could operate in many different orders while building the car, first adding doors and later windows.
@@
 
-\begin{figure}[t]
-  \centering
-  \begin{minipage}{0.54\linewidth}
-    \includegraphics[width=0.9\linewidth]{gfig1.png}    
-  \end{minipage}
-  \begin{minipage}{0.44\linewidth}
-    \caption{Biological details of glycan production. There are many types of sugar monomer building blocks; %(each represented by standard colored shapes);
-      for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
-    \label{fig:glycan-rule}
-  \end{minipage}
-% \vspace{-9mm}
-\end{figure}
-
 In Figure~\ref{fig:glycan-rule}, we present details of glycan production.
@@
 
+% 3 body collision
+%
+% It’s not clear why the reviewer sees these reactions as implausible. The
+% monomer addition reaction proceeds in three steps. First, the monomer
+% itself binds to the relevant enzyme. Second, this monomer-enzyme complex
+% interacts with the glycan, and through a process of stochastic binding and
+% unbinding it identifies a particular region of the glycan, i.e. a motif,
+% which can be several nodes deep. The enzyme makes contacts with several
+% nodes of the glycan, and the actual depth of the recognised motif can vary
+% from one to several. Finally, the monomer is transferred from the enzyme to
+% the relevant node of the glycan, a reaction catalysed at the reaction
+% centre of the enzyme. At no step are there 3-body collisions, all these
+% reactions occur as a series of 2-body collisions. These reactions have been
+% understood in molecular detail. See Biswas & Thattai, Biochem Soc Trans
+% 48:891 (2020) for several examples of motifs of varying depths, and for a
+% graphical representation of the monomer and glycan in complex with the
+% enzyme.
+
+% Stochastic.
+%
+%Glycosylation occurs as an enzyme-catalysed reaction whose
+% deterministic kinetics are described by Michaelis-Menten or similar rate
+% equations. However, these are only valid in the large number limit. In the
+% cell, the number of molecules is small and the process is dominated by
+% stochastic chemical kinetics. Individual reactions occur as Poisson
+% processes whose parameters can be associated with the deterministic
+% equations. The steady state distributions (or time-dependent distributions
+% in case of a non-stationary process) of such stochastic chemical kinetic
+% systems can be derived using the standard application of stochastic process
+% tools and these match very well with the known data in many cellular
+% processes.
+
+% rule size and optimally
+%
+% The actual depth associated with different enzymes varies, and at least up
+% to depths of 4 or 5 there is no real physical restriction. So one may
+% consider a maximum depth criterion. See Biswas & Thattai, 2020. If there is
+% an optimisation question, it can be framed as “what is the minimal number
+% of compartments required for enzymes that only read to a certain maximum
+% depth”. Or more useful perhaps, what is the minimum depth required upto a
+% certain maximum number of compartments (typically mammalian Golgi have only
+% 3 compositionally distinct compartments). The answer to this question then
+% allows a more rigorous biochemical analysis of which enzyme is responsible
+% for which addition reaction. Overall, therefore, it’s not that we’re making
+% a claim about optimality, but we are making claims about limits on enzyme
+% depths or required compartments, both of which are experimentally
+% measurable features.
+
+% application of synthesis
+%
+% See above answer to start with. The problem is that, although the
+% donor/acceptor monomers are well known for most enzymes, the actual motifs
+% are not known except for some well-studied enzymes. So at this point in
+% time, it’s not immediate possible to associate our enzyme/depth predictions
+% with real enzymes. However, what is known is the number of distinct
+% compartments. If we can show for example that at some maximum depth, it
+% takes 5 compartments to generate some glycan set, we know something is
+% wrong either in the dataset (incomplete?) or the enzyme depth assumption,
+% since the number of compartments is 3. The reviewer is correct that there
+% are likely to be many *more* constraints than simply depth constraints on
+% real enzymes, this means that real enzymes would need even more
+% compartments than we predict. I.e. we lower-bound the required number of
+% compartments. It is possible that this approach will also help identify
+% glycans that exit the Golgi at an intermediate compartment. This is, for
+% e.g., known to happen with the SARS-CoV-2 virus, which exits after the
+% first cisterna. Such predictions will have cell-biological implications.
+
+% existing available solution
+%
+% See Biswas & Thattai for several examples of concrete rules. The reason we
+% did not focus on concrete rules is that, in fact, such rules are only known
+% for a handful of well-studied enzymes. At this stage, it is more relevant
+% to try to place broad constraints on the space of rules that are suggested
+% by the data. See Jaiman & Thattai for concrete examples of entire reaction
+% sets.
+
+\begin{figure}[t]
+  \centering
+  \begin{minipage}{0.54\linewidth}
+    \includegraphics[width=0.9\linewidth]{gfig1.png}    
+  \end{minipage}
+  \begin{minipage}{0.44\linewidth}
+    \caption{Biological details of glycan production. There are many types of sugar monomer building blocks; %(each represented by standard colored shapes);
+      for example, GalNAc (yellow square), GlcNAc (blue square), Galactose (yellow circle), Sialic Acid (purple diamond), Fucose (red triangle) and so on \cite{Varki2017}.}
+    \label{fig:glycan-rule}
+  \end{minipage}
+% \vspace{-9mm}
+\end{figure}
+
 %--------------------- DO NOT ERASE BELOW THIS LINE --------------------------

### 2021-07-06 papers/sugar-synth/algo.tex
fig spacing
15 page fit

@@
 a set of production rules that produce a given set $\mu$ of molecules.
-Our synthesis method \textsc{SugarSynth} is presented in Algorithm~\ref{alg:sugar-synth}.
+Our method \textsc{SugarSynth} is Algorithm~\ref{alg:sugar-synth}.
 Here, we are considering only the single compartment case.
@@
 We use templates to model the search space of rules.
-A template is also a tree that has a depth and
+A template is a tree that has a depth and
 the internal nodes of the tree have the same number of children.
-Two variables label each node of the template.
+Two variables label each node. %of the template.
 One variable is for choosing the sugar at the node and the other is for describing
@@
 
+A node in a production rule can be in four situations.
+In Figure~\ref{fig:rule-parts}, we illustrate the situations.
+The first situation is when a node is in the expanding part, which are shown in 
+dark gray. % nodes are the expanding nodes.
+The second situation is when a node is not in the rule.
+The dashed area are the absent nodes.
+Among the matching nodes, we have two cases.
 \begin{wrapfigure}{r}{0.35\textwidth}
-
+  % \vspace{-12mm}
         \center
@@
 \end{wrapfigure}
-
-A node in a production rule can be in four situations.
-In Figure~\ref{fig:rule-parts}, we illustrate the situations.
-The first situation is when a node position is in the expanding part.
-The dark gray nodes are the expanding nodes.
-The second situation is when a node position is not in the rule.
-The dashed area are the absent nodes. Among the matching nodes, we have two cases.
-The third situation is when a node position is in the matching part
+The third situation is when a node is in the matching part
 and has expanding descendants.
@@
 % the nodes that are ancestors of expanding part among the matching nodes.
-Finally, the fourth situation is the rest of the node positions
-in the matching part, which is the light gray area.
+Finally, the fourth situation is the rest of the nodes
+in the matching part, which is in light gray.
 % The nodes in the rest of the light gray area are in the fourth situation.
@@
 Let $K =\{Expand,Absent,MatchAns,Match\}$ be the set of symbols to
-indicate the set of four situations.
-Let $KVars$ be the unbounded set of variables with domain $K$. 
+indicate the situations.
+Let $KVars$ be an unbounded set of variables with domain $K$. 
 % We will use the pool of $KVars$ to add variables to the templates.
-Our templates are sufficiently expressive to cover all aspects of biology.
-The templates are defined as follows.
+Our templates are sufficiently expressive to cover all aspects of biology, which are defined as follows.
 
@@
 Since we negate the returned formula by $\textsc{EncodeProduce}$ 
-and then we check the satisfiability.
+and then we check the satisfiability,
 % the encoding will not be correct
 % due to the fresh variables.
-Therefore, we need to introduce universal quantifiers for the
-fresh variables.
-We introduce universal quantifiers over $\tau$ and $cuts$
-variables before adding to $nCons$.
+we need to introduce universal quantifiers over the
+fresh variables. %before adding to $nCons$.
 % Intuitively, the universal quantifiers say that all possible
@@
 Afterwards, the loop goes to the next iteration.
-Since the domain of all the variables is finite, eventually the loop must
-terminate.
+% Since the domain of all the variables is finite, eventually the loop must
+% terminate.
 
@@
 find the node to match the root of $t$.
-Naturally, it returns $\lfalse$ if there
+% Naturally,
+It returns $\lfalse$ if there
 is no $\ell$th ancestor of $v$. % i.e., there is no match.
-The variable $mark$ is the timestamp for node $v$.
+The variable $mark$ is the timestamp for $v$.
 % During matching, we will add constraints stating that the matching part of the
@@
 % are added at or after $mark$.
-The local variable $c$ collects the conjunction of constraints as they are
+The variable $c$ collects the  constraints as they are
 generated. % and $c$ is initially $\ltrue$.
-The local variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
+The variable $v_r$ is initially equal to the root $v_{0r}$ of $t$
 and traverses the nodes in $t$.
-The while loop at line~\ref{line:encodep-while} in $\textsc{EncodeP}$ starts with
+The while loop at line~\ref{line:encodep-while} starts with
 the $\ell$th ancestor from $v$,
 matches all the ancestors up to the parent of $v$.
-As the loop traverses down, it also traverses $t$ along the matching
+As the loop traverses down, it also traverses $t$ 
 using variable $v_r$.
@@
 \ref{line:encodep-next-vr} and \ref{line:encodep-update-vr}
-if the ancestors of $v$ also traverse along with the $j$th child.
+if the ancestors of $v$ also traverse along the $j$th child.
 % (see if condition at line \ref{line:encodep-jth-child-cond}).
 
-\vspace{-3mm}
+\vspace{-10mm}
 \input{alg-encode-produce.tex}
@@
 At line~\ref{line:encodep-ans-match}, the loop adds constraints stating that
-the corresponding node $v_r$ in the template rule is
+the corresponding node $v_r$ in $t$ is
 $MatchAns$, sugar matches in $v_r$ and $v'$, and
-node $v'$ is added before $mark$.
+node $v'$ was added before $mark$.
 The loop at line~\ref{line:encodep-sub-for-loop}, iterates over
@@
 $v$, it is matched at line~\ref{line:mtree} with the corresponding
-child of $v_r$ by calling $\textsc{MatchTree}$, which traverses
-rule template and molecule and generate constraints to encode that their
-nodes match.
-We will discuss $\textsc{MatchTree}$ shortly.
+child of $v_r$ by calling $\textsc{MatchTree}$, which
+will be discussed shortly.
+% traverses
+% rule template and molecule and generate constraints to encode that their
+% nodes match.
+% We will discuss $\textsc{MatchTree}$ shortly.
 % $\textsc{MatchTree}$ is also called with a third parameter
@@
 concrete rules.
-We need to swap the roles of $\nu$ and $M$ at their occurrences
+We swap the roles of $\nu$ and $M$ at their occurrences
 at line \ref{line:encodep-ans-match} in \textsc{EncodeP} and
 line \ref{line:matchtree-cons} in \textsc{MatchTree}.
-Furthermore, $\kappa(v_r)$ is a concrete value depending on the situation of $v_r$ in
-the rule.
+Furthermore, $\kappa(v_r)$ is a concrete value.
+% depending on the situation of $v_r$ in the rule.
 In the functions, the variable is to be replaced by the
@@
 \begin{theorem}
-  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns a set of rules $R$, then if $R$ produces a molecule that has depth less than $h+1$, the it is in $\mu$ (soundness).
-  If it fails to synthesize rules, there are no rule set from the budget of depth $d$ and number $n$ (completeness).
+  If \textsc{SugarSynth}($\mu$, $d$, $n$) returns  rules $R$, then if $R$ produces a molecule that has depth less than $h+1$, the it is in $\mu$ (soundness).
+  If it fails to find rules, there are no rule sets within the budget of $d$ and  $n$ (completeness).
 \end{theorem}
 
-The soundness is true due to the construction of constraints.
-Our algorithm always terminates. At each iteration, at least one solution, i.e,
+The soundness is true by construction.
+% Our algorithm always terminates.
+At each iteration, at least one solution, i.e,
 a set of rules is discarded.
-In fact, we discard many because in each iteration we reject production of a counterexample molecule, but this may discard whole range of rule sets. 
-Since the set of all possible rules is finite, we
-guarantee termination.
+In fact, we discard many because in each iteration we reject all rule sets that may produce a counterexample molecule. 
+Since the set of all possible rules is finite, we terminate.
 Therefore, the completeness holds.
-In the worst case, the entire search space is explored.
+% In the worst case, the entire search space is explored.
 

### 2021-07-06 papers/sugar-synth/experiments.tex
dsadsa
15 page fit

--- 2021-04-21 papers/sugar-synth/experiments.tex
@@
-In this section, we present our implementation of the proposed method.
+% In this section, we present our implementation of the proposed method.
 % %
@@
 {\em Implementation:} 
-We have implemented method \textsc{SugarSynth} in a prototype tool {\ourtool}.
+We have implemented \textsc{SugarSynth} in a tool {\ourtool}~\cite{Charles2013}.
 %
 The tool, written in {\tt C++}, uses {\zthree}\cite{z3} as the SMT solver
-to discharge the generated satisfiability queries.
+to solve the satisfiability queries. We ran the experiments  on a laptop with 8GB RAM and 2.4GHz CPU.
 %
@@
 in the literature~\cite{Jaiman2018}.
+\choosefinal{}{
 More experiments on the variations of the problem
-are included in the appendix~\ref{sec:ex-variants}. %\hyperref[sec:appendix]{}.
+are included in the Appendix~\ref{sec:ex-variants}.} %\hyperref[sec:appendix]{}.
 

### 2021-07-06 papers/sugar-synth/model.tex
minor changes
15 page fit

@@
 rules $R$ from a set of molecules $Q$.
-We have discussed in section~\ref{sec:bio} that all the production rules
+We have discussed in Section~\ref{sec:bio} that all the production rules
 are not applied at the same time.
@@
 %
-However, we do not necessarily know the production rules to produce the molecules.
+However, we may not know the production rules to produce the molecules.
 %
-We will be developing a method to find the production rules.
+We will be developing a method to find the rules.
 %
